### 2022-07-26 paper.tex
Initial Overleaf Import

\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cite}
\usepackage[letterpaper, total={6.5in, 9in}]{geometry}
\usepackage{mathpazo}
\usepackage{sourcecodepro}

\usepackage{hyperref}

\newcommand{\setcomp}[2]{\left\{ #1 \ \Big|\ #2 \right\}}
\newcommand{\rngto}[1]{1{:}#1}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\absdet}[1]{\abs{#1}}
\newcommand{\dv}[1]{\mathrm{d}{#1}}
\newcommand{\Exp}{\mathrm{Exp}}
\newcommand{\vect}{\mathrm{vec}}
\newcommand{\vectu}{\mathrm{vecu}}

\begin{document}

\title{Efficient Unconstraining
  Parameter Transforms for Hamiltonian Monte Carlo}
\author{Meenal Jhajaria \\ \small Flatiron Institute 
\and Seth Axen \\ \small University of T\"ubingen 
\and Adam Haber \\ \small Weizmann Institute 
\and Sean Pinkney \\ \small Omnicom Media Group
\and Bob Carpenter \\ \small Flatiron Institute}
\date{DRAFT: \today}
\maketitle

\begin{abstract}
  \noindent
  This paper evaluates the statistical and computational
  efficiency of unconstraining parameter transforms for Hamiltonian
  Monte Carlo sampling.
\end{abstract}

\section{Introduction}

In statistical computing, we often need to compute high-dimensional
integrals over densities $\pi(x)$ (e.g., Bayesian estimation or
prediction, $p$-value calculations, etc.).  The only black-box
techniques that work for general high-dimensional integrals are
Markov chain Monte Carlo (MCMC) methods.  The most effective MCMC
method in high dimensions is Hamiltonian Monte Carlo (HMC).  HMC works
by simulating the Hamiltonian dynamics of a fictitious particle
representing the value being sampled coupled with a momentum term.

Although it's possible to write HMC samplers that work for simply
constrained values such as upper- and/or lower-bounds
\cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is
much more challenging to do the same for complex constraints such as
simplexes or positive definite matrices or for densities involving
multiple constrained values.  Instead, it is far more common to map
the constrained values to unconstrained values before sampling
\cite{}.  This presents the issue of selecting which transform to use
among an infinite set of options, which is the topic of this paper.

%Sampling algorithms play an important role in Statistics, in this case
%we refer to drawing samples from a probability density or distribution
%(not survey sampling). Monte Carlo Markov chain methods are commonly
%used for high dimensional sampling. In this paper, we look at
%Hamiltonian Monte Carlo, which uses hamiltonian dynamics to propose
%samples for the Metropolis algorithm. Bayesian Inference often uses
%parameters with constraints, but HMC optimizes on the unconstrained
%space. It is difficult to use Monte Carlo estimation on a constrained
%domain ~\cite{neal2008optimal}

\section{Changes of Variables}

If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
smooth, monotonic, bijection $f$, then
\[
  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
\]
where the Jacobian of the inverse transform is defined by
\[
  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
\]
and
\[
  \absdet{J_{f^{-1}}(y)}
  = \abs{\det J_{f^{-1}}(y)}.
\]

\section{Transforms}

To draw samples from a distribution $p_X(x)$, defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y}=\mathbb{R}^m$ for $m \ge n$.

Let $f\colon \mathcal{Y} \to \mathcal{X}$ be a smooth and continuous map.
Only when $m = n$ can $f$ be bijective.
When $m > n$, multiple values of $y \in \mathcal{Y}$ may map to a given value of $x = f(y)$.
In this case, to make the function bijective, we consider an additional smooth and continuous map $g\colon \mathcal{Y} \to \mathcal{Z}$, where $\mathcal{Z}$ can be uniquely parameterized by $m - n$ real degrees of freedom.
We then define $h\colon \mathcal{Y} \to \mathcal{X} \times \mathcal{Z}: y \mapsto (f(y), g(y)) = (x, z)$.
$f$ and $g$ must be chosen so that $h$ is bijective almost-everywhere;
that is, the values of $y$ for which $h$ is non-bijective must have no probability mass.
When $h$ is non-bijective at single points, we call these singularities.
While singularities themselves are in general not a problem for sampling with MCMC, the region of a target density near the singularity tends to have high curvature.

Given a proper joint distribution with a smooth and continuous density $p_{X,Y}(x, y | \theta)$ for some parameters $\theta$, we can use the usual change of variables formula to write

\[
  p_Y(y | \theta) = p_{X,Z}(f(y), g(y) | \theta) |J_h(y)|.
\]

To sample from a target density $p_X(x | \theta)$, let $p_{X,Z}(x, z | \theta) = p_X(x | \theta) p_Z(z | x, \theta)$.
Then we must choose some proper distribution $p_Z$ that is smooth and continuous.
Given that choice, we find that
\[
  p_Y(y | \theta) = p_X(f(y) | \theta) p_Z(g(y) | f(y), \theta) |J_h(y)|.
\]

Note that when $m > n$, it is insufficient to specify a transform $h$.
One must also select a prior $p_Z$, and it is desirable to establish some general properties that a given pair $h$ and $p_Z$ should have.
In addition to the above properties, the optimal pair would
1. discard as few degrees of freedom $|m - n|$ as possible.
2. produce a $p_Z$ and $|J_y|$ and their gradients that are efficient to compute.
3. produce a $p_Y$ that is easy to sample from.

We can expand a bit on the latter point.
As a general rule, the more normal $p_Y$ is, the easier it is to sample from (REF?).
Additionally, near singularities, small perturbations to $y$ tend to result in larger changes to $h(y)$ than far from the singularity.
This warps the volume around the singularity, which causes $|J_y(y)|$ to smoothly approach infinity near the singularity.
If $m$ is large, this is generally not a problem, as the "curse of dimensionality" works in our favor, and there is exponentially more volume away from the singularity than near the singularity.
However, for low $m$, it is necessary that $p_{X,Z}$ goes to 0 near the singularity faster than $|J_y(y)|$ goes to infinity.
While $p_X$ is generally fixed by the user-specified model and may not have this necessary property, the user may then choose a $p_Z$ that has this property.

\section{Preliminaries}

[SA: TODO: eliminate or merge with following section]

To compute Jacobians of functions with matrix-valued inputs or outputs, we need to implicitly or explicitly choose for the inputs and outputs a set of coordinates with dimension equal to the number of degrees of freedom of the matrix
Unless otherwise specified, we choose the coordinates to be the vectorization of the matrix.

We define the (bijective) vectorization of a matrix $X \in \mathbb{R}^{N \times K}$ with columns $x_k$ for $k \in \{1, ..., K\}$ as the vector
\[\vect(X) = \begin{bmatrix}x_1 \\ x_2 \\ \vdots \\ x_M\end{bmatrix}\]
formed by stacking the columns of $X$.

For $X \in \mathbb{R}^{N \times K}$, let $\vectu(X)$ be the map that stacks only the parts of the columns of $X$ above and including the diagonal.
This map is sometimes called the half-vectorization and is bijective for upper-triangular or symmetric $X$.

Similarly, $\vectu_-(X)$ extracts the same elements except the diagonal.
When $X$ is strictly upper-triangular or skew-symmetric (i.e. $X = -X^\top$), then this map is bijective.

\section{Computing Jacobian determinants}

[SA: TODO: remove everything we do not use later, add strategy for making non-square Jacobian square]

Here we introduce several strategies for computing Jacobian determinants.

The vectorization has the useful property that for all matrices $A$, $X$, and $B$, $\vect(AYB) = (B^\top \otimes A) \vect(Y)$, where $\otimes$ is the Kronecker product.
This allows us to convert expressions of matrices to expressions of vectors, from which the Jacobian can be read off.
For example, let $X = AYB$. Then $\dv{X} = A \dv{Y} B$, and $\vect(\dv{X}) = (B^\top \otimes A) \vect(\dv{Y})$, and the Jacobian is $J = B^\top \otimes A$.

When the previously discussed strategies are insufficient to compute the Jacobian, we bypass the Jacobian computation entirely and use the exterior product $\wedge$.
For example, let $X \in \mathbb{R}^{N \times K}$.
The exterior product
\[
  \dv{X_{11} }\wedge \dv{X_{21}} \wedge \ldots \wedge \dv{X_{NK}} = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X}_{ij}
\]
is another more explicit way of writing the integration measure $\dv{X_{11}} \dv{X_{21}} \dots \dv{X_{NK}}$ that has useful properties for computing Jacobian determinants.
The basic properties we will use are

\begin{itemize}
  \item $a\dv{x} \wedge b\dv{y} = (ab)(\dv{x} \wedge \dv{y})$ for $a,b \in \mathbb{R}$
  \item $(\dv{x} + \dv{y}) \wedge \dv{z} = \dv{x} \wedge \dv{z} + \dv{y} \wedge \dv{z}$
  \item $\dv{x} \wedge \dv{y} = - \dv{y} \wedge \dv{x}$
  \item $\dv{x} \wedge \dv{x} = 0$
\end{itemize}

The connection between the exterior product and the Jacobian determinant is given by the following expression.
Given $f: y \mapsto x$ for $x,y \in \mathbb{R}^N$,
\[\bigwedge_{i=1}^N \dv{x_i} = \det(J_f(y)) \bigwedge_{i=1}^N \dv{y_i}.\]

We can then find the Jacobian determinant by first writing all differential forms $\dv{x_i}$ or $\dv{y_i}$ and then computing the exterior product until the expression looks like the above.
For our purposes, we will ignore the sign of the Jacobian determinant.

A related useful property is that for $X,Y,B \in \mathbb{R}^{N \times K}$, if $\dv{X} = B \dv{Y}$, then
\[\bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_i} = \det(B)^K \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{Y_i}\]

Following [REF, Edelman], we use the notation $(\cdot)^\wedge$ as a shorthand to represent the exterior product of the unique elements of $\cdot$.

For example,
\begin{itemize}
  \item Rectangular matrix $\{X \in \mathbb{R}^{N \times K}\}$: $(\dv{X})^\wedge = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_{ij}}$
  \item Lower-triangular rectangular matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i < j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{\min(i, K)} \dv{X_{ij}}$
  \item Diagonal matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i \ne j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^{\min(N, K)} \dv{X_{ii}}$
  \item Symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = X_{ji}\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^i \dv{X_{ij}}$
  \item Skew-symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = -X_{ji}, X_{ii} = 0\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{i-1} \dv{X_{ij}}$
\end{itemize}

Using this notation, for example, $(B \dv{X})^\wedge = \det(B)^K (\dv{X})^\wedge$

\section{Unit simplex}

A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
values that sums to one.  As such, there are only $N$ degrees of
freedom, because if $x$ is an $N$-simplex, then
\[
  x_N = -(x_1 + x_2 + \cdots + x_{N-1}).
\]
Simplexes are useful for representations of multinomial probabilities
(e.g., probabilities of categories in a classification problem).

The set of unit $N$-simplexes is conventionally denoted
\[
  \Delta^N = \setcomp{x \in \mathbb{R}^{N + 1}}{\textrm{sum}(x) = 1
    \textrm{ and }
    x_n \geq 0 \textrm{ for } n \in \rngto{N}}
\]
Geometrically, an $N$-simplex is the convex closure of $N+1$ points
that are 1 in one coordinate and 0 elsewhere.  For example, the
3-simplex is the complex closure of
$\begin{bmatrix}1 & 0 & 0 \end{bmatrix},
\begin{bmatrix} 0 & 1 & 0 \end{bmatrix}$,
and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$.

\subsection{StickBreaking Transform}

The StickBreaking transform ican be be understood from the stick-breaking construction for Dirichlet \cite{sethurman}. Intuitively, this comprises of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$. Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  R^{N-1}$, for $1 \leq i \leq N$ as:	
\[
y_i
= \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
   \right) \text{for break proportion} \, 
   z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
\]

The inverse transform $ f^{-1} \colon R^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  \Delta^{N-1}$ is defined as:
\[
x_i =
\left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right) \text{for break proportion} \, z_i = \mathrm{logit}^{-1} \left( y_i
                             + \log \left( \frac{1}{N - i}
                                            \right)\right).
                                            \]

An $N$ dimensional unit simplex $\Delta^{N-1}$ has $N-1$ degrees of freedom (Notice that this inverse transform only maps the first $N-1$ elements, the last element of the simplex $x_{N} = 1 - \sum_1^{N-1}{x_i}$). The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix. Much like the alr transform, for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
\begin{align*}
\mathbf{J}_{i, i} &= \frac{\partial x_i}{\partial y_i}
=
\frac{\partial x_i}{\partial z_i} \,
\frac{\partial z_i}{\partial y_i}\\
\mathbf{J}_{i, i} &= \left(
  1 - \sum_{k' = 1}^{k-1} x_{k'}
   \right) z_k (1 - z_k),
\end{align*}

Absolute determinant of the diagonal matrix $\mathbf{J}$ is the product of its diagonal entries:
\begin{align*}
	\abs{\, det \, \textbf{J} \,} = \prod_{i=1}^{N-1} \textbf{J}_{i,i}
\end{align*}

The correction term $p_Y(y) = p_X(f^{-1}(y))\,
\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
\subsection{Additive log ratio transform}

The unconstraining transform for the identified softmax is known as
the additive log ratio (ALR) transform
\cite{aitchison1982statistical}, which is a bijection
$\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
$x \in \Delta^{N-1}$ by
\[
  \textrm{alr}(x)
  = \begin{bmatrix}\displaystyle
    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
  \end{bmatrix}.
\]

The inverse additive log ratio transform maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{alr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,
\[
  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
\]

To calculate the change of variables adjustment, we consider only the
first $N-1$ coordinates of the result, because the last is defined in
terms of the first $N-1$.  For convenience, we define a function 
$s:\mathbb{R}^{N-1} \rightarrow \mathbb{R}^{N-1}$ that operates on
$N-1$ unconstrained variables and returns the first $N-1$ components
of the $\textrm{alr}^{-1}(y)$, which is
defined for $y \in \mathbb{R}^{N-1}$ by
\[
  s(y) = \frac{\exp(y)}{\textrm{sum}(\exp(y)) + 1}
  = \begin{bmatrix}
    \textrm{alr}^{-1}(y)_1
    & \cdots &
    \textrm{alr}^{-1}(y)_{N-1}
    \end{bmatrix}.
\]
Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
we can transform to a density over unconstrained parameters $y \in
\mathbb{R}^{N-1}$ by applying the inverse ALR transform and adjusting
for the change of variables, which yields
\[
  p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{s}(y)},
\]
where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
and $\absdet{J_s(y)}$ is the absolute value of its determinant.

\subsubsection{Softmax Transform}

To calculate the determinant of the Jacobian of the inverse transform,
we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
$\textrm{exp}$ is the elementwise exponential function and
\textrm{norm} is defined by
\[
  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
\]
As such, the resulting Jacobian determinant is the product of the
Jacobian determinants of the component functions,
\[
  \absdet{J_s(y)}
  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
\]
where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
function is diagonal, so the determinant is the product of the
diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
\[
  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
\]
As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
$\textrm{norm}$ to derive the Jacobian,
\[
  J_{\textrm{norm}}
  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
  \textrm{vector}_{N-1}(1)^{\top},
\]
where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
the matrix determinant lemma,\footnote{The matrix determinant lemma
  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
    \textrm{det}(A).\]}
we have
\begin{eqnarray*}
  \textrm{absdet}(J_{\textrm{norm}}(z))
  & = &
  \left(
    1
    + \textrm{vector}_{N-1}(1)^{\top}
    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
    \frac{-z}{(1 + \textrm{sum}(z))^2}
    \right)
    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
        \right)
  \\[6pt]
  & = &
  \left(
    1 
    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
        \textrm{sum}(z))^2} \right)
  \right)
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = &
        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
\end{eqnarray*}
Thus the entire absolute determinant of the Jacobian is defined by the
product, 
\[
  \absdet{J_s(y)}
  \ = \
  \textrm{prod}(\exp(y))
  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
\]
and our final expression for densities for unconstrained $y \in
\mathbb{R}^{N-1}$ is
\[
  p_Y(y)
  = p_X(\textrm{alr}^{-1}(y))
  \, \textrm{prod}(\exp(y))
  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
\]

%\subsection{Softmax Transform}
%
%The softmax function can be understood from Multinomial Logistic
%Regression employed for predicting probabilities of a Categorically
%distributed variable. Geometrically it maps $R^K$ to the boundary of a
%unit K-simplex(it is simply the convex hull of $k+1$ affinely
%independent points in $R^K$. Essentially it transforms a vector of
%size K to another vector of size K where the outputs sum to 1. It is
%worth noting that the mapping is actually from $R^K$ to $R^{K-1}$, so
%when a vector of size $K$ is transformed the $K_th$ vector is simply
%$1$- sumof k-1 vectors

\subsection{Simplex softmax parameterization}

Let $\Delta^n$ indicate the unit $n$-simplex with $n$ positive
elements that sum to 1 and $\Delta^n_-$ indicate the first $n-1$
elements, from which the final element can be uniquely determined.

We define the transformation
$\phi: \mathbb{R}^{n-1} \to \Delta^n_-: y \mapsto x_-$, where
$x=\begin{pmatrix}x_- \\ \frac{1}{r}\end{pmatrix} \in \Delta^n$,
$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$, and
$r = 1 + \sum_{i=1}^{n-1} e^{y_i}$.

First we compute the scalar derivatives:
\[
\begin{aligned}
  \frac{\mathrm{d} r}{\mathrm{d} y_j}
  &= e^{y_j} = r x_j\\
  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i}
  \frac{\mathrm{d} r}{\mathrm{d} y_j}
  = \delta_{ij} x_i - x_i x_j, \quad 1 \le i \le n-1
\end{aligned}
\]
where
$\delta_{ij} = \begin{cases} 1 &\text{if } i = j \\ 0
  &\text{otherwise}\end{cases}$ is the Kronecker delta.

If $\mathrm{diag}(x)$ is the diagonal matrix whose diagonal are the
elements of $x$, then the Jacobian is
\[
  J = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top) \operatorname{diag}(x_-),
\]
where $\boldsymbol{1}_n$ is the $n$-vector of ones.

Using Sylvester's determinant theorem,
$|I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = 1 -
\boldsymbol{1}_{n-1}^\top x_- = 1 - \sum_{i=1}^{n-1} x_i = x_n$, so
$$\mathrm{correction} = |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$

\subsection{Simplex softmax parameterization through parameter expansion}

We define the transformation
$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
(x_-, r)$, where $r = \sum_{i=1}^n e^{y_i}$ and
$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$..

First we compute the scalar derivatives:
\[
\begin{aligned}
  \frac{\mathrm{d} r}{\mathrm{d} y_j}
  &= e^{y_j} = r x_j
  \\
  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i} \frac{\mathrm{d} r}{\mathrm{d} y_j} = \delta_{ij} x_i - x_i x_j,
\end{aligned}
\]
which corresponds to the Jacobian
\[
  J = \begin{pmatrix}I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top & -x_- \\
    r \boldsymbol{1}_{n-1}^\top & r \end{pmatrix} \mathrm{diag}(x).
\]

For invertible $A$, the determinant of the block matrix
$\begin{pmatrix}A & B \\ C & D\end{pmatrix}$ is $|A| |D-CA^{-1}B|$.  A
square matrix is invertible iff its determinant is non-yero.  From the
previous section,
\[
  |I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = x_n > 0,
\]
so the determinant of the Jacobian is
\[
  |J| = x_n \left|r + r \boldsymbol{1}_{n-1}^\top (I_{n-1} - x_-
    \boldsymbol{1}_{n-1}^\top)^{-1} x_-\right|
  \prod_{i=1}^n x_i.
\]

Let $w = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top)^{-1} x_-$. Then,
\[
\begin{aligned}
    w - x_- \sum_{i=1}^{n-1} w_i &= x_-\\
    w &= x_- \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
    \sum_{i=1}^{n-1} w_i &= \sum_{i=1}^{n-1} \left( x_- (1 - \sum_{i=1}^{n-1} w_i) \right) = \left(\sum_{i=1}^{n-1} x_i \right) \left(1 - \sum_{i=1}^{n-1} w_i\right) = (1 - x_n)  \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
    \sum_{i=1}^{n-1} w_i &= \frac{1 - x_n}{x_n} = \frac{1}{x_n} - 1\\
    w &= x_- \left(1 - \left(\frac{1}{x_n} - 1\right)\right) = \frac{1}{x_n} x_-
  \end{aligned}
\]

Then
\[
  |J| = x_n r \left|1 + \frac{1}{x_n}\sum_{i=1}^{n-1} x_i\right|
  \prod_{i=1}^n x_i = r \prod_{i=1}^n x_i
\]

To keep the target distribution proper, we must select a prior
distribution $\pi(r)$ for $r$.  If we choose $r \sim \chi_n$, then the
product of the correction and the density of the prior for $r$ is
proportional to

\[
  \mathrm{correction}
  = \pi(r) |J| = r^n e^{-r^2/2} \prod_{i=1}^n x_i
  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{2}\left(\sum_{i=1}^n
      e^{y_i}\right)^2\right).
\]

Alternatively, if we choose $r \sim \mathrm{Gamma}(n, 1)$, then
\[
  \mathrm{correction} = \pi(r) |J| = r^n e^{-r} \prod_{i=1}^n x_i =
  \exp\left(\sum_{i=1}^n y_i - \sum_{i=1}^n e^{y_i}\right).
\]
This latter correction is equivalent to the sampling procedure from
the Dirichlet distribution with $\alpha_i=1$, where
$z_i \sim \mathrm{Exponential}(1)$ and
$y = \frac{z}{\sum_{i=1}^n z_i}$.

Both of these corrections can be captured with the generalization
\[
  \mathrm{correction}
  = \pi(r) |J|
  = r^n e^{-r^p/p} \prod_{i=1}^n x_i
  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{p} \left(\sum_{i=1}^n e^{y_i}\right)^p\right),
\]
for $p > 0$, which corresponds to $r \sim \text{Generalized-Gamma}(1, n, p)$.

\section{Semi-orthogonal matrices (Stiefel manifold)}

A matrix $X \in \mathbb{R}^{N \times K}$ for $K \le N$ is semi-orthogonal if and only if $X^\top X = I_K$.
The semi-orthogonal matrices form a space called the Stiefel manifold.
When $K=1$, $X$ is a point on the $(N-1)$-sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$.
We will consider such unit vectors in the next section.
When $K=N$, $X$ is an element of the orthogonal group $O(N)$ and additionally satisfies $\det(X) = \pm 1$.
$O(N)$ is comprised of two disconnected submanifolds, one being the Special Orthogonal group $SO(N)$ with $\det(X) = +1$, and the other having $\det(X) = -1$.
$SO(2)$ and $SO(3)$ are the groups of rotation matrices of 2D and 3D vectors, respectively.

By choosing a set of rules such as Gram-Schmidt orthonormalization of the canonical basis in $\mathbb{R}^{N \times N}$, it is possible to analytically compute from $X$ an orthogonal complement $X^\perp \in \mathbb{R}^{N \times (N - K)}$, such that $\tilde{X}^\top \tilde{X} = I_N$, where $\tilde{X} = \begin{bmatrix}X & X^\perp \end{bmatrix} \in O(N)$.

Note that if we differentiate both sides of the constraint $X^\top X = I_K$, we get $\dv{X}^\top X + X^\top \dv{X} = 0$, which leads to $(X^\top \dv{X})^\top = - X^\top \dv{X}$.
Therefore, $X^\top \dv{X}$ is a $K \times K$ skew-symmetric matrix with $\frac{K(K-1)}{2}$ unique entries in the strict upper triangle.

Densities of distributions on the Stiefel manifold are usually written with respect to the invariant metric on the Stiefel manifold, which we will define separately for the two cases $K=N$ and $K < N$.

For $X \in O(N)$ with the $i$th column written $x_i$, we define the invariant measure as the exterior product of the unique elements of this skew-symmetric matrix:
\[
  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N (X^\top \dv{X})_{ij} = \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j.
\]
This measure is called invariant because if $f(X) = A X B$ for $A,B \in O(N)$ and $\mathcal{S} \subset O(N)$ then $\int_{\mathcal{S}} (X^\top \dv{X}) = \int_{f(\mathcal{S})} (X^\top \dv{X})$.
It is also called the Haar measure on $O(N)$ \cite{muirhead2009aspects}..

To choose an invariant measure for $K < N$, we first choose an $X^\perp$ analytically computed from $X$.
Because $X^\perp$ is analytically computed, $\dv{X^\perp_{ij}} = 0$.
We can then construct the invariant measure for $K<N$ from the invariant measure for $O(N)$:

\[
  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j = \bigwedge_{j=1}^K \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j,
\]

Though we needed to choose a $X^\perp$, the invariant measure does not depend on the choice of $X^\perp$.
Moreover, this measure is invariant to left-actions by $O(N)$ and right-actions by $O(K)$.
For more details, see \cite{muirhead2009aspects}.

When $K=1$, the invariant measure is equivalent to the usual Hausdorff measure on the sphere.

\subsection{Givens parameterization of Stiefel manifold}

\subsection{Cayley parameterization of Stiefel manifold}

\subsection{Polar parameterization of Stiefel manifold}

\subsection{Householder parameterization of Stiefel manifold}

\subsection{Matrix exponential parameterization of Special Orthogonal group}

\section{Unit sphere}

The unit sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$ consists of all unit vectors $x$ of length 1, i.e. $x^\top x = 1$.
It is a special case of the Stiefel manifold.

Densities on the sphere are generally defined with respect to the spherically uniform measure.
This measure can be defined in many ways that are equivalent to within a constant factor that is not relevant for MCMC.
We will take the same approach as with the Stiefel manifold and define the spherically uniform measure as the measure that is invariant to transformations by $N \times N$ orthogonal matrices (rotations and reflections).
As discussed in the previous section, if $\tilde{X} \in O(N)$ is an orthogonal matrix whose first column is $x$, then $\tilde{X}^\top x = e_1$, where $e_1$ is the unit vector whose first element is 1, and $(\tilde{X}^\top \dv{x})_1 = 0$.
For convenience, we will assume that the rules used to produce $\tilde{X}$ enforce that $\det(\tilde{X}) = +1$.
The spherically uniform measure is then written
\[
    (\dv{x})^\wedge \equiv (\tilde{X}^\top \dv{x})^\wedge = \bigwedge_{i=2}^N (\tilde{X}^\top \dv{x})_i.
\]

\subsection{Embedding in real space using parameter expansion}

Let $f\colon y \mapsto (r, x)$ for $y \in \mathbb{R}^N$, $r > 0$, and $x \in \mathbb{S}^{N-1}$.
This map is bijective everywhere except $y=0$.

To compute the Jacobian determinant of the map, we first compute the differential form of $y=rx$:

\[\dv{y} = \dv{r} x + r \dv{x}.\]

Multiplying both sides by $\tilde{X}^\top$, we find

\[
    \tilde{X}^\top \dv{y} = \dv{r} \tilde{X}^\top x + r \tilde{X}^\top \dv{x} = \dv{r} e_1 + r (\tilde{X}^\top \dv{x}).
    \]

Then

\[
    (\tilde{X}^\top \dv{y})_i = \begin{cases}
        \dv{r} & \text{ if } i = 1 \\
        r (\tilde{X}^\top \dv{x})_i & \text{ if } i \ne 1
    \end{cases}.
\]

Note that $(\tilde{X}^\top \dv{y})^\wedge = \det({\tilde{X}}) (\dv{y})^\wedge = (\dv{y})^\wedge$.

Taking the exterior product of both sides, we then find

\[
\begin{aligned}
    (\dv{y})^\wedge &= \dv{r} \wedge \left(\bigwedge_{i=2}^N r (\tilde{X}^\top \dv{x})_i \right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge\right)\\
    \dv{r} \wedge (\dv{x})^\wedge &= r^{1-N} (\dv{y})^\wedge = \lVert y \rVert^{1-N} (\dv{y})^\wedge
\end{aligned}
\]

So the absolute Jacobian determinant is $\absdet{J_f(y)} = \lVert y \rVert^{1-N}$.

In the constrained space there is a free radial parameter $r$ that is unconstrained on the positive reals.
As a result, nothing prevents $r$ from growing arbitrarily large, which would cause the posterior to be improper, i.e. to have infinite mass.
Thus, we must also specify a prior $\pi(r)$.
While we may want to customize $\pi(r)$, a reasonable default choice is $r \sim \chi_N$, which has the density $\pi(r) \propto r^{N-1} \exp\left(-\frac{1}{2}r^2\right) $.

The combined log density correction is then
\[
  \absdet{J_f(y)} \pi(r) = \absdet{J_f(y)} \pi(\lVert y \rVert) \propto \lVert y \rVert^{1-N} \lVert y \rVert^{N-1} \exp\left(-\frac{1}{2}\lVert y \rVert^2\right) = \exp\left(-\frac{1}{2}\lVert y \rVert^2\right).
\]

That is, if $x$ is uniformly distributed on the sphere and $r$ follows this prior, then $y$ has a zero-centered standard multivariate normal distribution.

Since HMC is especially efficient at sampling normal distributions, this parameterization is very efficient when the marginal posterior distribution of $x$ is very diffuse on the sphere.
However, as the target distribution becomes concentrated on $x$, the geometry near the singularity $y=0$ may become problematic.

\subsection{Parameterization using hyperspherical coordinates}

Given $z \in \mathbb{R}^N$, $r > 0$ and angles $\phi=\{\phi_1, \ldots, \phi_{N-1}\}$, where $\phi_{N-1} \in (0, 2\pi]$ and $\phi_i \in (0, \pi]$ for $i < N-1$, we define the relationship between $z$ and $(r, \phi)$ as

\[
  z_i = \left(r \prod_{j=1}^{i-1} \sin(\phi_j)\right) \begin{cases}
    1 & \text{ if } i = N\\
    \cos(\phi_i) & \text{otherwise}
  \end{cases}.
\]

For the plane $N=2$ this works out the usual polar coordinate system:
\[
    \begin{aligned}
        z_1 &= r \cos\phi_1\\
        z_2 &= r \sin\phi_1
    \end{aligned}
\]

For $N=3$ this is \footnote{
    Note that this definition of hyperspherical coordinates differs for $N=3$ from the usual definition of spherical coordinates.
    It can be made the same through the coordinate permutation $(z_1, z_2, z_3) \to (z_2, z_3, z_1)$.
}
\[
    \begin{aligned}
        z_1 &= r \cos\phi_1\\
        z_2 &= r \sin\phi_1\cos\phi_2\\
        z_3 &= r \sin\phi_1\sin\phi_2
    \end{aligned}
\]

Our strategy is to compute the Jacobian determinant of the bijective map $g: (r, \phi) \mapsto z$ and relate it to the Jacobian determinant from the previous section to find the Jacobian determinant of the bijective map $f_2: \phi \mapsto x$ for $x \in \mathbb{S}^{N-1}$.

To do so, note first that
\[
  z_i^2 = \left(r^2 \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \begin{cases}
    1 & \text{ if } i = N\\
    \cos^2(\phi_i) & \text{otherwise}
  \end{cases}.
\]

Note also that
\[
    \begin{aligned}
        z_N^2 &= r^2 \prod_{j=1}^{N-1} \sin^2(\phi_j)\\
        z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \cos^2(\phi_{N-1}) + \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j) \right) \sin^2(\phi_{N-1}) = r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\\
        z_{N-2}^2 + z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \cos^2(\phi_{N-2}) + \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \sin^2(\phi_{N-2}) = r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\\
        &\vdots \\
        \sum_{k=1}^i z_{N - i + k}^2 &= r^2 \prod_{j=1}^{N - i} \sin^2(\phi_j)\\
        &\vdots \\
        \sum_{k=1}^N z_{N - i + k}^2 &= r^2
    \end{aligned}
\]

We then differentiate both sides of this expression to get the differential form:

\[
    \begin{aligned}
        2 z_N \dv{z_N} &= 2r\dv{r} + 2 r^2 \left(\prod_{j=1}^{N-1} \sin^2(\phi_j)\right)\left( \sum_{j=1}^{N-1} \cot(\phi_j) \dv{\phi_j} \right)\\
        2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-2} \cot(\phi_j)\right)\\
        2z_{N-2}\dv{z_{N-2}} + 2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-3} \cot(\phi_j)\right)\\
        &\vdots \\
        2\sum_{k=1}^i z_{N - i + k} \dv{z_{N - i + k}} &= 2 r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N - i} \sin^2(\phi_j)\right) \left(\sum_{k=1}^{N-i} \cot(\phi_k) \dv{\phi_k}\right)\\
        &\vdots \\
        2\sum_{k=1}^N z_{N - i + k} \dv{z_{N - i + k}} &= 2r\dv{r}
    \end{aligned}
\]

For each side separately, we now take the exterior product for all $i \in 1\ldots N$.
On the left side, we start at the top with $(2z_N dz_n) \wedge (2 z_{N-1} \dv{z_{N-1}} + 2 z_N \dv{z_N}) = 2^2 z_{N} z_{N-1} (\dv{z_N} \wedge \dv{z_{N-1}})$.
Note that the $\dv{z_N}$ from the second equation is eliminated since the exterior product already contains that term.
This trend continues as we proceed to the bottom, and the result is
\[
    \bigwedge_{i=1}^N 2z_{N-i+1} \dv{z_{N-i+1}} = \bigwedge_{i=1}^N 2z_i \dv{z_i} = 2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge
\]

Note that

\[
  \prod_{i=1}^N z_i = r \prod_{i=1}^{N-1} r \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j) = r^N \prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j)
\]

On the right side, we start at the bottom instead with
\[
(2r\dv{r} + 2 r^2 \sin(\phi_1) \cos(\phi_1) \dv{\phi_1}) \wedge (2r\dv{r}) = 4 r^3 \sin(\phi_1) \cos(\phi_1) (\dv{\phi_1} \wedge \dv{r}).
\]
Again, the $\dv{r}$ term in the second-from-bottom equation cancels since $\dv{r}$ is already included in the bottom term.
This trend also continues as we proceed to the top, and the result is then

\[
\begin{aligned}
    \left(\bigwedge_{i=1}^{N - 1} 2 r^2 \left(\prod_{j=1}^{N - i - 1} \sin^2(\phi_j)\right) \sin(\phi_{N-i}) \cos(\phi_{N-i}) \dv{\phi_{N-i}} \right) \wedge (2r\dv{r}) \\
    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_{N-i}) \cos(\phi_{N-i}) \prod_{j=1}^{N-i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
    = 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
\end{aligned}
\]

Equating the exterior products of each side, we have

\[
\begin{aligned}
    2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge &= 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
\end{aligned}
\]

As a result, the Jacobian determinant is

\[
    \absdet{J_g(r, \phi)} = r^{N-1} \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}
\]

As a quick check, for the circle ($N=1$), $\absdet{J_g(r, \phi)} = r$, and for the sphere ($N=2$), $\absdet{J_g(r, \phi)} = r^2 \sin(\phi_1)$, which are the expected results.
A similar result was given in \cite{muirhead2009aspects}.

Comparing with the previous section, we find

\[
\begin{aligned}
    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge \right) \\
    &= (-1)^{N-1} r^{N-1} \left((\dv{x})^\wedge \wedge \dv{r}\right)\\
    (\dv{x})^\wedge &= (-1)^{N-1}\left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) (\dv{\phi})^\wedge
\end{aligned},
\]

so the absolute Jacobian determinant for the bijective map from hyperspherical coordinates to the sphere is
\[
    \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}.
\]

However, we're not done yet.
Recall that $\phi_j \in (0, \pi]$ for $i < N-1$ and $\phi_{N-1} \in (0, 2\pi]$.
So far we've essentially changed the problem of sampling on the hypersphere in $N$ dimensions to the problem of sampling on $N-1$ intervals, or, alternatively, $N-2$ hemicircles and 1 circle.

There are a few considerations when choosing the remaining parameterizations.
First, and most importantly, we want a smooth geometry.
Second, we want to avoid unnecessary work.
For example, if we choose parameter expansion for these parameters, we will end up with $N-1$ extra degrees of freedom.
Third, we want to avoid costly functions if possible.
For example, computing $\sin$ and $\cos$ is about 3 times more costly than computing $\exp$ and about 20 times more costly than computing a square.

\section{Cholesky factors of correlation matrices}

A $K \times K$ correlation matrix $\Omega$ is positive definite, has a unit diagonal, and all off-diagonal entries are between -1 and 1 (exclusive). Because it is positive definite, it can be Cholesky factored (see \ref{cholesky}) to a $K \times K$ lower-triangular matrix $L$
with positive diagonal elements such that $\Omega = L \cdot L^T$. Because the correlation matrix has a unit diagonal,

\[
  \Omega_{k,k} = L_k \cdot L_k^T = 1
\]

each row vector $L_k$ of the Cholesky factor is of unit length. The length and positivity
constraint allow the diagonal elements of L to be calculated from the off-diagonal
elements, so that a Cholesky factor for a $K \times K$ correlation matrix requires only $\binom{K}{2}$ unconstrained parameters.

\subsection{Hyperbolic tangent transform}

Suppose $y$ is a vector of $\binom{K}{2}$ unconstrained values. Let $z$ be a lower-triangular matrix with zero diagonal and below diagonal entries filled by row. For example, in the $3 \times 3$ case,
\[
z = \begin{bmatrix}
0 & 0 & 0\\
\tanh{y_1} & 0 & 0\\
\tanh{y_2} & \tanh{y_3} & 0\\
\end{bmatrix}
\]

The matrix $z$, with entries in the range $( -1, 1)$, is then transformed to the Cholesky factor $x$, by taking
\[
x_{i, j}=\left\{\begin{array}{lll}
0 & \text { if } i<j & \text { [above diagonal] } \\
\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i=j & \text { [on diagonal] } \\
z_{i, j} \sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i>j & \text { [below diagonal] }
\end{array}\right.
\]

In the $3 \times 3$ case, this yields

\[
x=\left[\begin{array}{ccc}
1 & 0 & 0 \\
z_{2,1} & \sqrt{1-x_{2,1}^{2}} & 0 \\
z_{3,1} & z_{3,2} \sqrt{1-x_{3,1}^{2}} & \sqrt{1-\left(x_{3,1}^{2}+x_{3,2}^{2}\right)}
\end{array}\right]
\]

where the $z_{i,j}\in (-1, 1)$ are the tanh-transformed $y$.

\subsection{Absolute Jacobian determinant of inverse transform}

The Jacobian of the full transform is the product of the Jacobians of its component
transforms. First, for the inverse transform $z = \tanh{y}$, the derivative is
\[
\frac{d}{dy} \tanh y = \frac{1}{(\cosh y)^{2}}
\]
Second, for the inverse transform of $z$ to $x$, the resulting Jacobian matrix $J$ is of
dimension $\binom{K}{2} \times \binom{K}{2}$, with indexes $(i, j)$ for $(i > j)$. The Jacobian matrix is lower
triangular, so that its determinant is the product of its diagonal entries, of which
there is one for each $(i, j)$ pair.
\[
|\operatorname{det} J|=\prod_{i>j}\left|\frac{d}{d z_{i, j}} x_{i, j}\right|
\]

where

\[
\frac{d}{d z_{i, j}} x_{i, j}=\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}}
\]

So the combined log Jacobian determinant
of the complete inverse transform is given by

\[
\log |\operatorname{det} J|=-2 \sum_{n \leq(\substack{K \\ 2}} \log \cosh y+\frac{1}{2} \sum_{i>j} \log \left(1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}\right)
\]

\section{Positive definite matrices}

[SA: all transformations below can be generalized to SPD matrices of fixed rank, but we'd need to know what the common reference measure is for densities of SPD matrices, and I can't recall seeing one.]

An $N \times N$ positive definite (PD) matrix $X$ is a symmetric matrix that satisfies $v^\top X v > 0$ for all non-zero vectors $v$.
Covariance matrices are examples of PD matrices.

Density functions for distributions of PD matrices, such as the Wishart distribution, are generally defined with respect to the Lebesgue measure of the unique triangular elements $(\dv{X})^\wedge$.

\subsection{Cholesky parameterization}\label{cholesky}

Given a positive (semi-)definite matrix $X$ of rank $K$, we can uniquely decompose it as $X = L L^\top$ where $L$ is an $N \times N$ lower triangular matrix with $L_{jj} > 0$ for all $j \le K$ and $L_{ij} = 0$ for all $j > K$.
$L$ has $\frac{K(K+1)}{2} + (N - K)K$ degrees of freedom.
When $K=N$, $X$ is positive-definite.

We can construct $X$ by composing 2 bijective maps.
The first, $f_1$, enforces the constraint of the positive diagonal:

\[L_{ij} = f_1(Y)_{ij} = \begin{cases} \exp(Y_{ij}) & \text{ if } i = j \le K \\ Y_{ij} & \text{otherwise} \end{cases}\]

The Jacobian determinant for this map is $\absdet{J_{f_1}(Y)} = \exp(\sum_{i=1}^K Y_{ii})$.

The second map is $f_2: L \mapsto L L^\top = X$.
To compute the Jacobian of the map, we use the half-vectorization $\vectu(X)$.
Note that $X_{ij} = \sum_{k=1}^{\min(i, j)} L_{ik} L_{jk}$.
$X_{ij}$ is thus only dependent on elements of $L$ that are below and to the left of the $(i, j)$ position.
As a result, the Jacobian of $\vectu \circ f_2 \circ \vectu^{-1}$ is triangular, and its determinant is only the product of its diagonal elements: $\absdet{J_{f_2}} = \prod_{i=1}^N \prod_{j=1}^i \frac{\partial X_{ij}}{\partial L_{ij}}$.

In this index range, $\frac{\partial X_{ij}}{\partial L_{ij}} = (1 + \delta_{ij}) L_{ii}$.
As a result

\[
  \absdet{J_{f_2}(Z)} = 2^N \prod_{i = 1}^N \prod_{j = 1}^i L_{ii} = 2^N \prod_{i=1}^N |L_{ii}|^{N - i + 1} \\
\]

Composing the two bijective maps as $f = f_2 \circ f_1$, the resulting Jacobian determinant is

\[
  \absdet{J_{f}(Y)} = 2^N \exp\left( \sum_{i=1}^N (N - i + 2) Y_{ii}\right).
\]

\subsection{Matrix exponential parameterization}

For every PD matrix $X \in \mathbb{R}^{N \times N}$, there is a unique symmetric matrix $Y$ such that $X = \Exp(Y)$, where $\operatorname{Exp}(\cdot)$ is the matrix exponential, defined as
\[\Exp(Y) = \sum_{k=0}^\infty \frac{1}{k!} Y^k.\]

Let $Y = U \Lambda U^\top$ be the symmetric eigendecomposition of $Y$, where the matrix of eigenvectors $U$ is orthogonal, and the $\Lambda = \operatorname{diag}(\lambda)$ for $\lambda \in \mathbb{R}^N$.
Then $Y^2 = U \Lambda U^\top U \Lambda U^\top = U \Lambda^2 U^\top$.
In fact, $Y^c = U \Lambda^c U^\top$ for all real $c$.
As a result,
\[\Exp(Y) = U \Exp(\Lambda) U^\top,\]
where $\Exp(\Lambda)_{ij} = \exp(\lambda_i) \delta_{ij}$.

$\Exp(Y)$ is one example of a matrix function $f$, which we can roughly define for symmetric matrices as any function that obeys
\[f(Y) = U f(\Lambda) U^\top,\]
where $f(\Lambda)_{ij} = f(\lambda_i)\delta_{ij}$.
\footnote{
  In a similar way, we can generalize other scalar functions like the inverse, trigonometric, and hyperbolic functions to functions of matrices.
  All results in this section apply to these functions as well.
}

For such functions $f$, we can compute a common form of the Jacobian determinant.
Differentiating the expression, we find
\[
\begin{aligned}
  \dv{(f(Y))} &= \dv{U} f(\Lambda) U^\top + U f'(\Lambda) \dv{\Lambda} U^\top + U f(\Lambda) \dv{U}^\top\\
  U^\top \dv{(f(Y))} U &= U^\top \dv{U} f(\Lambda) + f(\Lambda) \dv{U}^\top U + f'(\Lambda)\dv{\Lambda},
\end{aligned}
\]
where $f'(\Lambda)_{ij} = \frac{\dv{(f(\lambda_i))}}{\dv{\lambda_i}}$.

Recall that $U^\top \dv{U}$ is skew-symmetric.
Note also that for skew-symmetrix $Z = -Z^\top$ and diagonal matrix $\operatorname{diag}(v)$,
\[(Z\operatorname{diag}(v) + \operatorname{diag}(v)Z^\top)_{ij} = Z_{ij} v_j + v_i Z_{ji} = Z_{ij} v_j - v_i Z_{ij} = Z_{ij} (v_j - v_i) = Z \circ V,\]
where $V_{ij} = v_j - v_i$ and $\circ$ is the Hadamard (elementwise) product.

From the right-hand side, note then that
\[(U^\top \dv{(f(Y))} U)_{ij} = \begin{cases} (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) & \text{ if } i \ne j \\ f'(\lambda_i) \dv{f(\lambda_i)} &\text{ if } i = j \end{cases}\]

The exterior product of the diagonal part is
\[
  \bigwedge_{i=1}^N f'(\lambda_i) \dv{\lambda_i} = \left(\prod_{i=1}^N f'(\lambda_i)\right) (\dv{\Lambda})^\wedge,
\]
while the exterior product of the sub-diagonal part is
\[
  \bigwedge_{i=1}^N \bigwedge_{j=i+1}^N (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) = \left( \prod_{i=1}^N \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i)) \right) \left(U^\top \dv{U} \right)^\wedge
\]

Lastly, note for any symmetric $Z$ and orthogonal matrix $U$,
\[(U^\top \dv{Z} U)^\wedge = \det(U^\top)^N \det(U)^N (\dv{Z})^\wedge = (\dv{Z})^\wedge.\]

As a result, we can compute $(\dv(f(Y)))^\wedge$ by taking the exterior product of the sub-diagonal and diagonal parts: 
\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i))\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]

If $f$ is the identity map $f(Y) = Y$, then
\[ (\dv{Y})^\wedge = \left(\prod_{i=1}^N \prod_{j=1}^{i-1} (\lambda_j - \lambda_i)\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]

Combining the latter two expressions:
\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} \frac{f(\lambda_j) - f(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge\]

[REF, Edelman's notes] previously found this result.

For $X = f(Y) = \Exp(Y)$ then,
\[ (\dv{X})^\wedge = \left(\prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge,\]

so the Jacobian determinant of the exponential map is $\absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right|$.

Note that this expression is undefined when any eigenvalues are repeated.
This is not a problem for MCMC, since the eigenvalues are non-unique with probability 0, but non-unique eigenvalues could occur when maximizing a log-density.

Because $\lim_{h \to 0} \frac{f(x) - f(x + h)}{h} = f'(x)$, we can then modify the expression to handle such cases:
\[
  \absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \prod_{j=1}^i \begin{cases}
    \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right| &\text{ if } \lambda_i \ne \lambda_j \\ 
    \exp(\lambda_i) &\text{ if } \lambda_i = \lambda_j
  \end{cases}.
\]

If the eigenvalues are sorted in ascending order, then the log Jacobian determinant can be more stably computed as

\[
  \log\absdet{J_{\Exp}(Y)} = \sum_{j=1}^N j\lambda_j + \sum_{i=1}^j \begin{cases}
    \log(1 - \exp(\lambda_i - \lambda_j)) - \log(\lambda_j - \lambda_i) &\text{ if } \lambda_i \ne \lambda_j \\ 
    0 &\text{ if } \lambda_i = \lambda_j
  \end{cases}.
\]

While the cost of computing the Jacobian determinant is $O(N^2)$, the cost of computing the eigendecomposition naively is $O(N^3)$, so this transformation is expensive.
An attractive alternative would be to define the transform directly from the eigendecomposition of $X$.

\subsection{Eigendecomposition parameterization}

To define a transform directly from the symmetric eigendecomposition of $X$, we first need to note that the eigendecomposition is not unique; if we swap any two columns $i$ and $j$ of $U$ as well as swapping the eigenvalues $\lambda_i$ and $\lambda_j$, the result is the same matrix.
Moreover, if we reverse the sign of any given eigenvector, it has no effect on the eigenvalues or $X$.

When the eigenvalues are all unique, we can make the eigendecomposition also unique by enforcing an ascending order of eigenvalues $0 < \lambda_i < \lambda_j$ for $j > i$ and by forcing the first entry of each eigenvector to be positive.

So we can construct a bijective map with the maps $f_1$ that produces an ordered vector of positive eigenvalues and $f_2$ that produces an orthogonal matrix whose first row contains only positive entries.
Consider $\lambda = f_1(y_1)$ and $U = f_2(y_2)$.

Let $f_1$ be the map that computes each $\lambda_i$ as
\[
  \lambda_i = \begin{cases}
   \exp\bigl((y_1)_1\bigr) &\text{ if } i = 1\\
    \lambda_{i-1} + \exp\bigl((y_1)_i\bigr) &\text{ if } 1 < i \le N
  \end{cases}
\]
The Jacobian of this map is triangular, so its determinant is $\absdet{J_{f_1}(y_1)} = \exp\left(\sum_{i=1}^N (y_1)_i\right)$.

We can then choose for $f_2$ any of the transforms to the orthogonal matrices previously discussed, though care must be taken to avoid multimodality.
For example, if the transform covers the entire orthogonal group, then $2^N$ matrices map to the same orthogonal matrix with positive first row, which will manifest as $2^N$ modes in the unconstrained distribution.
If, however, we can modify one of the transforms to natively produce only matrices with this constraint, then we can eliminate these undesirable modes.

The combined Jacobian determinant of $f: (y_1, y_2) \mapsto f_2(y_2) \operatorname{diag}(f_1(y_1)) f_2(y_2)^\top = X$ is then 
\[ \absdet{J_{f}(y_1, y_2)} = \absdet{J_{f_2}(y_2)} \exp\left(\sum_{i=1}^N (y_1)_i\right).\]

\subsubsection*{Acknowledgements}

We would like to thank \url{matrixcalculus.org} for providing an
easy-to-use symbolic matrix derivative calculator.

\bibliography{all}{}
\bibliographystyle{plain}

\end{document}

### 2022-07-27 dummypaper.tex
figures

--- 2022-07-26 paper.tex
@@
 \usepackage{sourcecodepro}
+\usepackage{graphicx}
 
@@
 
-\section{Changes of Variables}
-
-If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
-smooth, monotonic, bijection $f$, then
-\[
-  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
-\]
-where the Jacobian of the inverse transform is defined by
-\[
-  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
-\]
-and
-\[
-  \absdet{J_{f^{-1}}(y)}
-  = \abs{\det J_{f^{-1}}(y)}.
-\]
-
 \section{Transforms}
@@
 
-\section{Preliminaries}
+\subsection{Changes of Variables}
 
-[SA: TODO: eliminate or merge with following section]
+If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
+smooth, monotonic, bijection $f$, then
+\[
+  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
+\]
+where the Jacobian of the inverse transform is defined by
+\[
+  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
+\]
+and
+\[
+  \absdet{J_{f^{-1}}(y)}
+  = \abs{\det J_{f^{-1}}(y)}.
+\]
+
+\subsection{Computing Jacobian}
 
@@
 
-\section{Computing Jacobian determinants}
-
-[SA: TODO: remove everything we do not use later, add strategy for making non-square Jacobian square]
-
 Here we introduce several strategies for computing Jacobian determinants.
@@
 For example, let $X = AYB$. Then $\dv{X} = A \dv{Y} B$, and $\vect(\dv{X}) = (B^\top \otimes A) \vect(\dv{Y})$, and the Jacobian is $J = B^\top \otimes A$.
-
-When the previously discussed strategies are insufficient to compute the Jacobian, we bypass the Jacobian computation entirely and use the exterior product $\wedge$.
-For example, let $X \in \mathbb{R}^{N \times K}$.
-The exterior product
-\[
-  \dv{X_{11} }\wedge \dv{X_{21}} \wedge \ldots \wedge \dv{X_{NK}} = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X}_{ij}
-\]
-is another more explicit way of writing the integration measure $\dv{X_{11}} \dv{X_{21}} \dots \dv{X_{NK}}$ that has useful properties for computing Jacobian determinants.
-The basic properties we will use are
-
-\begin{itemize}
-  \item $a\dv{x} \wedge b\dv{y} = (ab)(\dv{x} \wedge \dv{y})$ for $a,b \in \mathbb{R}$
-  \item $(\dv{x} + \dv{y}) \wedge \dv{z} = \dv{x} \wedge \dv{z} + \dv{y} \wedge \dv{z}$
-  \item $\dv{x} \wedge \dv{y} = - \dv{y} \wedge \dv{x}$
-  \item $\dv{x} \wedge \dv{x} = 0$
-\end{itemize}
-
-The connection between the exterior product and the Jacobian determinant is given by the following expression.
-Given $f: y \mapsto x$ for $x,y \in \mathbb{R}^N$,
-\[\bigwedge_{i=1}^N \dv{x_i} = \det(J_f(y)) \bigwedge_{i=1}^N \dv{y_i}.\]
-
-We can then find the Jacobian determinant by first writing all differential forms $\dv{x_i}$ or $\dv{y_i}$ and then computing the exterior product until the expression looks like the above.
-For our purposes, we will ignore the sign of the Jacobian determinant.
-
-A related useful property is that for $X,Y,B \in \mathbb{R}^{N \times K}$, if $\dv{X} = B \dv{Y}$, then
-\[\bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_i} = \det(B)^K \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{Y_i}\]
-
-Following [REF, Edelman], we use the notation $(\cdot)^\wedge$ as a shorthand to represent the exterior product of the unique elements of $\cdot$.
-
-For example,
-\begin{itemize}
-  \item Rectangular matrix $\{X \in \mathbb{R}^{N \times K}\}$: $(\dv{X})^\wedge = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_{ij}}$
-  \item Lower-triangular rectangular matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i < j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{\min(i, K)} \dv{X_{ij}}$
-  \item Diagonal matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i \ne j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^{\min(N, K)} \dv{X_{ii}}$
-  \item Symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = X_{ji}\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^i \dv{X_{ij}}$
-  \item Skew-symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = -X_{ji}, X_{ii} = 0\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{i-1} \dv{X_{ij}}$
-\end{itemize}
-
-Using this notation, for example, $(B \dv{X})^\wedge = \det(B)^K (\dv{X})^\wedge$
 
@@
 
-\section{Semi-orthogonal matrices (Stiefel manifold)}
+\section{Results}
 
-A matrix $X \in \mathbb{R}^{N \times K}$ for $K \le N$ is semi-orthogonal if and only if $X^\top X = I_K$.
-The semi-orthogonal matrices form a space called the Stiefel manifold.
-When $K=1$, $X$ is a point on the $(N-1)$-sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$.
-We will consider such unit vectors in the next section.
-When $K=N$, $X$ is an element of the orthogonal group $O(N)$ and additionally satisfies $\det(X) = \pm 1$.
-$O(N)$ is comprised of two disconnected submanifolds, one being the Special Orthogonal group $SO(N)$ with $\det(X) = +1$, and the other having $\det(X) = -1$.
-$SO(2)$ and $SO(3)$ are the groups of rotation matrices of 2D and 3D vectors, respectively.
+\begin{figure}
+    \centering
+    \includegraphics[width=1.2\textwidth]{figures/ess.png}
+    \caption{Effective Sample Size vs Total LeapFrog Steps}
+    \label{fig:hyg}
+\end{figure}
 
-By choosing a set of rules such as Gram-Schmidt orthonormalization of the canonical basis in $\mathbb{R}^{N \times N}$, it is possible to analytically compute from $X$ an orthogonal complement $X^\perp \in \mathbb{R}^{N \times (N - K)}$, such that $\tilde{X}^\top \tilde{X} = I_N$, where $\tilde{X} = \begin{bmatrix}X & X^\perp \end{bmatrix} \in O(N)$.
-
-Note that if we differentiate both sides of the constraint $X^\top X = I_K$, we get $\dv{X}^\top X + X^\top \dv{X} = 0$, which leads to $(X^\top \dv{X})^\top = - X^\top \dv{X}$.
-Therefore, $X^\top \dv{X}$ is a $K \times K$ skew-symmetric matrix with $\frac{K(K-1)}{2}$ unique entries in the strict upper triangle.
-
-Densities of distributions on the Stiefel manifold are usually written with respect to the invariant metric on the Stiefel manifold, which we will define separately for the two cases $K=N$ and $K < N$.
-
-For $X \in O(N)$ with the $i$th column written $x_i$, we define the invariant measure as the exterior product of the unique elements of this skew-symmetric matrix:
-\[
-  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N (X^\top \dv{X})_{ij} = \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j.
-\]
-This measure is called invariant because if $f(X) = A X B$ for $A,B \in O(N)$ and $\mathcal{S} \subset O(N)$ then $\int_{\mathcal{S}} (X^\top \dv{X}) = \int_{f(\mathcal{S})} (X^\top \dv{X})$.
-It is also called the Haar measure on $O(N)$ \cite{muirhead2009aspects}..
-
-To choose an invariant measure for $K < N$, we first choose an $X^\perp$ analytically computed from $X$.
-Because $X^\perp$ is analytically computed, $\dv{X^\perp_{ij}} = 0$.
-We can then construct the invariant measure for $K<N$ from the invariant measure for $O(N)$:
-
-\[
-  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j = \bigwedge_{j=1}^K \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j,
-\]
-
-Though we needed to choose a $X^\perp$, the invariant measure does not depend on the choice of $X^\perp$.
-Moreover, this measure is invariant to left-actions by $O(N)$ and right-actions by $O(K)$.
-For more details, see \cite{muirhead2009aspects}.
-
-When $K=1$, the invariant measure is equivalent to the usual Hausdorff measure on the sphere.
-
-\subsection{Givens parameterization of Stiefel manifold}
-
-\subsection{Cayley parameterization of Stiefel manifold}
-
-\subsection{Polar parameterization of Stiefel manifold}
-
-\subsection{Householder parameterization of Stiefel manifold}
-
-\subsection{Matrix exponential parameterization of Special Orthogonal group}
-
-\section{Unit sphere}
-
-The unit sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$ consists of all unit vectors $x$ of length 1, i.e. $x^\top x = 1$.
-It is a special case of the Stiefel manifold.
-
-Densities on the sphere are generally defined with respect to the spherically uniform measure.
-This measure can be defined in many ways that are equivalent to within a constant factor that is not relevant for MCMC.
-We will take the same approach as with the Stiefel manifold and define the spherically uniform measure as the measure that is invariant to transformations by $N \times N$ orthogonal matrices (rotations and reflections).
-As discussed in the previous section, if $\tilde{X} \in O(N)$ is an orthogonal matrix whose first column is $x$, then $\tilde{X}^\top x = e_1$, where $e_1$ is the unit vector whose first element is 1, and $(\tilde{X}^\top \dv{x})_1 = 0$.
-For convenience, we will assume that the rules used to produce $\tilde{X}$ enforce that $\det(\tilde{X}) = +1$.
-The spherically uniform measure is then written
-\[
-    (\dv{x})^\wedge \equiv (\tilde{X}^\top \dv{x})^\wedge = \bigwedge_{i=2}^N (\tilde{X}^\top \dv{x})_i.
-\]
-
-\subsection{Embedding in real space using parameter expansion}
-
-Let $f\colon y \mapsto (r, x)$ for $y \in \mathbb{R}^N$, $r > 0$, and $x \in \mathbb{S}^{N-1}$.
-This map is bijective everywhere except $y=0$.
-
-To compute the Jacobian determinant of the map, we first compute the differential form of $y=rx$:
-
-\[\dv{y} = \dv{r} x + r \dv{x}.\]
-
-Multiplying both sides by $\tilde{X}^\top$, we find
-
-\[
-    \tilde{X}^\top \dv{y} = \dv{r} \tilde{X}^\top x + r \tilde{X}^\top \dv{x} = \dv{r} e_1 + r (\tilde{X}^\top \dv{x}).
-    \]
-
-Then
-
-\[
-    (\tilde{X}^\top \dv{y})_i = \begin{cases}
-        \dv{r} & \text{ if } i = 1 \\
-        r (\tilde{X}^\top \dv{x})_i & \text{ if } i \ne 1
-    \end{cases}.
-\]
-
-Note that $(\tilde{X}^\top \dv{y})^\wedge = \det({\tilde{X}}) (\dv{y})^\wedge = (\dv{y})^\wedge$.
-
-Taking the exterior product of both sides, we then find
-
-\[
-\begin{aligned}
-    (\dv{y})^\wedge &= \dv{r} \wedge \left(\bigwedge_{i=2}^N r (\tilde{X}^\top \dv{x})_i \right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge\right)\\
-    \dv{r} \wedge (\dv{x})^\wedge &= r^{1-N} (\dv{y})^\wedge = \lVert y \rVert^{1-N} (\dv{y})^\wedge
-\end{aligned}
-\]
-
-So the absolute Jacobian determinant is $\absdet{J_f(y)} = \lVert y \rVert^{1-N}$.
-
-In the constrained space there is a free radial parameter $r$ that is unconstrained on the positive reals.
-As a result, nothing prevents $r$ from growing arbitrarily large, which would cause the posterior to be improper, i.e. to have infinite mass.
-Thus, we must also specify a prior $\pi(r)$.
-While we may want to customize $\pi(r)$, a reasonable default choice is $r \sim \chi_N$, which has the density $\pi(r) \propto r^{N-1} \exp\left(-\frac{1}{2}r^2\right) $.
-
-The combined log density correction is then
-\[
-  \absdet{J_f(y)} \pi(r) = \absdet{J_f(y)} \pi(\lVert y \rVert) \propto \lVert y \rVert^{1-N} \lVert y \rVert^{N-1} \exp\left(-\frac{1}{2}\lVert y \rVert^2\right) = \exp\left(-\frac{1}{2}\lVert y \rVert^2\right).
-\]
-
-That is, if $x$ is uniformly distributed on the sphere and $r$ follows this prior, then $y$ has a zero-centered standard multivariate normal distribution.
-
-Since HMC is especially efficient at sampling normal distributions, this parameterization is very efficient when the marginal posterior distribution of $x$ is very diffuse on the sphere.
-However, as the target distribution becomes concentrated on $x$, the geometry near the singularity $y=0$ may become problematic.
-
-\subsection{Parameterization using hyperspherical coordinates}
-
-Given $z \in \mathbb{R}^N$, $r > 0$ and angles $\phi=\{\phi_1, \ldots, \phi_{N-1}\}$, where $\phi_{N-1} \in (0, 2\pi]$ and $\phi_i \in (0, \pi]$ for $i < N-1$, we define the relationship between $z$ and $(r, \phi)$ as
-
-\[
-  z_i = \left(r \prod_{j=1}^{i-1} \sin(\phi_j)\right) \begin{cases}
-    1 & \text{ if } i = N\\
-    \cos(\phi_i) & \text{otherwise}
-  \end{cases}.
-\]
-
-For the plane $N=2$ this works out the usual polar coordinate system:
-\[
-    \begin{aligned}
-        z_1 &= r \cos\phi_1\\
-        z_2 &= r \sin\phi_1
-    \end{aligned}
-\]
-
-For $N=3$ this is \footnote{
-    Note that this definition of hyperspherical coordinates differs for $N=3$ from the usual definition of spherical coordinates.
-    It can be made the same through the coordinate permutation $(z_1, z_2, z_3) \to (z_2, z_3, z_1)$.
-}
-\[
-    \begin{aligned}
-        z_1 &= r \cos\phi_1\\
-        z_2 &= r \sin\phi_1\cos\phi_2\\
-        z_3 &= r \sin\phi_1\sin\phi_2
-    \end{aligned}
-\]
-
-Our strategy is to compute the Jacobian determinant of the bijective map $g: (r, \phi) \mapsto z$ and relate it to the Jacobian determinant from the previous section to find the Jacobian determinant of the bijective map $f_2: \phi \mapsto x$ for $x \in \mathbb{S}^{N-1}$.
-
-To do so, note first that
-\[
-  z_i^2 = \left(r^2 \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \begin{cases}
-    1 & \text{ if } i = N\\
-    \cos^2(\phi_i) & \text{otherwise}
-  \end{cases}.
-\]
-
-Note also that
-\[
-    \begin{aligned}
-        z_N^2 &= r^2 \prod_{j=1}^{N-1} \sin^2(\phi_j)\\
-        z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \cos^2(\phi_{N-1}) + \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j) \right) \sin^2(\phi_{N-1}) = r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\\
-        z_{N-2}^2 + z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \cos^2(\phi_{N-2}) + \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \sin^2(\phi_{N-2}) = r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\\
-        &\vdots \\
-        \sum_{k=1}^i z_{N - i + k}^2 &= r^2 \prod_{j=1}^{N - i} \sin^2(\phi_j)\\
-        &\vdots \\
-        \sum_{k=1}^N z_{N - i + k}^2 &= r^2
-    \end{aligned}
-\]
-
-We then differentiate both sides of this expression to get the differential form:
-
-\[
-    \begin{aligned}
-        2 z_N \dv{z_N} &= 2r\dv{r} + 2 r^2 \left(\prod_{j=1}^{N-1} \sin^2(\phi_j)\right)\left( \sum_{j=1}^{N-1} \cot(\phi_j) \dv{\phi_j} \right)\\
-        2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-2} \cot(\phi_j)\right)\\
-        2z_{N-2}\dv{z_{N-2}} + 2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-3} \cot(\phi_j)\right)\\
-        &\vdots \\
-        2\sum_{k=1}^i z_{N - i + k} \dv{z_{N - i + k}} &= 2 r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N - i} \sin^2(\phi_j)\right) \left(\sum_{k=1}^{N-i} \cot(\phi_k) \dv{\phi_k}\right)\\
-        &\vdots \\
-        2\sum_{k=1}^N z_{N - i + k} \dv{z_{N - i + k}} &= 2r\dv{r}
-    \end{aligned}
-\]
-
-For each side separately, we now take the exterior product for all $i \in 1\ldots N$.
-On the left side, we start at the top with $(2z_N dz_n) \wedge (2 z_{N-1} \dv{z_{N-1}} + 2 z_N \dv{z_N}) = 2^2 z_{N} z_{N-1} (\dv{z_N} \wedge \dv{z_{N-1}})$.
-Note that the $\dv{z_N}$ from the second equation is eliminated since the exterior product already contains that term.
-This trend continues as we proceed to the bottom, and the result is
-\[
-    \bigwedge_{i=1}^N 2z_{N-i+1} \dv{z_{N-i+1}} = \bigwedge_{i=1}^N 2z_i \dv{z_i} = 2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge
-\]
-
-Note that
-
-\[
-  \prod_{i=1}^N z_i = r \prod_{i=1}^{N-1} r \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j) = r^N \prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j)
-\]
-
-On the right side, we start at the bottom instead with
-\[
-(2r\dv{r} + 2 r^2 \sin(\phi_1) \cos(\phi_1) \dv{\phi_1}) \wedge (2r\dv{r}) = 4 r^3 \sin(\phi_1) \cos(\phi_1) (\dv{\phi_1} \wedge \dv{r}).
-\]
-Again, the $\dv{r}$ term in the second-from-bottom equation cancels since $\dv{r}$ is already included in the bottom term.
-This trend also continues as we proceed to the top, and the result is then
-
-\[
-\begin{aligned}
-    \left(\bigwedge_{i=1}^{N - 1} 2 r^2 \left(\prod_{j=1}^{N - i - 1} \sin^2(\phi_j)\right) \sin(\phi_{N-i}) \cos(\phi_{N-i}) \dv{\phi_{N-i}} \right) \wedge (2r\dv{r}) \\
-    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_{N-i}) \cos(\phi_{N-i}) \prod_{j=1}^{N-i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
-    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
-    = 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
-\end{aligned}
-\]
-
-Equating the exterior products of each side, we have
-
-\[
-\begin{aligned}
-    2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge &= 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
-    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
-\end{aligned}
-\]
-
-As a result, the Jacobian determinant is
-
-\[
-    \absdet{J_g(r, \phi)} = r^{N-1} \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}
-\]
-
-As a quick check, for the circle ($N=1$), $\absdet{J_g(r, \phi)} = r$, and for the sphere ($N=2$), $\absdet{J_g(r, \phi)} = r^2 \sin(\phi_1)$, which are the expected results.
-A similar result was given in \cite{muirhead2009aspects}.
-
-Comparing with the previous section, we find
-
-\[
-\begin{aligned}
-    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge \right) \\
-    &= (-1)^{N-1} r^{N-1} \left((\dv{x})^\wedge \wedge \dv{r}\right)\\
-    (\dv{x})^\wedge &= (-1)^{N-1}\left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) (\dv{\phi})^\wedge
-\end{aligned},
-\]
-
-so the absolute Jacobian determinant for the bijective map from hyperspherical coordinates to the sphere is
-\[
-    \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}.
-\]
-
-However, we're not done yet.
-Recall that $\phi_j \in (0, \pi]$ for $i < N-1$ and $\phi_{N-1} \in (0, 2\pi]$.
-So far we've essentially changed the problem of sampling on the hypersphere in $N$ dimensions to the problem of sampling on $N-1$ intervals, or, alternatively, $N-2$ hemicircles and 1 circle.
-
-There are a few considerations when choosing the remaining parameterizations.
-First, and most importantly, we want a smooth geometry.
-Second, we want to avoid unnecessary work.
-For example, if we choose parameter expansion for these parameters, we will end up with $N-1$ extra degrees of freedom.
-Third, we want to avoid costly functions if possible.
-For example, computing $\sin$ and $\cos$ is about 3 times more costly than computing $\exp$ and about 20 times more costly than computing a square.
-
-\section{Cholesky factors of correlation matrices}
-
-A $K \times K$ correlation matrix $\Omega$ is positive definite, has a unit diagonal, and all off-diagonal entries are between -1 and 1 (exclusive).
-Because it is positive definite, it can be Cholesky factored (see \ref{cholesky}) to a $K \times K$ lower-triangular matrix $L$
-with positive diagonal elements such that $\Omega = L \cdot L^T$. Because the correlation matrix has a unit diagonal,
-
-\[
-  \Omega_{k,k} = L_k \cdot L_k^T = 1
-\]
-
-each row vector $L_k$ of the Cholesky factor is of unit length. The length and positivity
-constraint allow the diagonal elements of L to be calculated from the off-diagonal
-elements, so that a Cholesky factor for a $K \times K$ correlation matrix requires only $\binom{K}{2}$ unconstrained parameters.
-
-\subsection{Hyperbolic tangent transform}
-
-Suppose $y$ is a vector of $\binom{K}{2}$ unconstrained values. Let $z$ be a lower-triangular matrix with zero diagonal and below diagonal entries filled by row. For example, in the $3 \times 3$ case,
-\[
-z = \begin{bmatrix}
-0 & 0 & 0\\
-\tanh{y_1} & 0 & 0\\
-\tanh{y_2} & \tanh{y_3} & 0\\
-\end{bmatrix}
-\]
-
-The matrix $z$, with entries in the range $( -1, 1)$, is then transformed to the Cholesky factor $x$, by taking
-\[
-x_{i, j}=\left\{\begin{array}{lll}
-0 & \text { if } i<j & \text { [above diagonal] } \\
-\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i=j & \text { [on diagonal] } \\
-z_{i, j} \sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i>j & \text { [below diagonal] }
-\end{array}\right.
-\]
-
-In the $3 \times 3$ case, this yields
-
-\[
-x=\left[\begin{array}{ccc}
-1 & 0 & 0 \\
-z_{2,1} & \sqrt{1-x_{2,1}^{2}} & 0 \\
-z_{3,1} & z_{3,2} \sqrt{1-x_{3,1}^{2}} & \sqrt{1-\left(x_{3,1}^{2}+x_{3,2}^{2}\right)}
-\end{array}\right]
-\]
-
-where the $z_{i,j}\in (-1, 1)$ are the tanh-transformed $y$.
-
-\subsection{Absolute Jacobian determinant of inverse transform}
-
-The Jacobian of the full transform is the product of the Jacobians of its component
-transforms. First, for the inverse transform $z = \tanh{y}$, the derivative is
-\[
-\frac{d}{dy} \tanh y = \frac{1}{(\cosh y)^{2}}
-\]
-Second, for the inverse transform of $z$ to $x$, the resulting Jacobian matrix $J$ is of
-dimension $\binom{K}{2} \times \binom{K}{2}$, with indexes $(i, j)$ for $(i > j)$. The Jacobian matrix is lower
-triangular, so that its determinant is the product of its diagonal entries, of which
-there is one for each $(i, j)$ pair.
-\[
-|\operatorname{det} J|=\prod_{i>j}\left|\frac{d}{d z_{i, j}} x_{i, j}\right|
-\]
-
-where
-
-\[
-\frac{d}{d z_{i, j}} x_{i, j}=\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}}
-\]
-
-So the combined log Jacobian determinant
-of the complete inverse transform is given by
-
-\[
-\log |\operatorname{det} J|=-2 \sum_{n \leq(\substack{K \\ 2}} \log \cosh y+\frac{1}{2} \sum_{i>j} \log \left(1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}\right)
-\]
-
-\section{Positive definite matrices}
-
-[SA: all transformations below can be generalized to SPD matrices of fixed rank, but we'd need to know what the common reference measure is for densities of SPD matrices, and I can't recall seeing one.]
-
-An $N \times N$ positive definite (PD) matrix $X$ is a symmetric matrix that satisfies $v^\top X v > 0$ for all non-zero vectors $v$.
-Covariance matrices are examples of PD matrices.
-
-Density functions for distributions of PD matrices, such as the Wishart distribution, are generally defined with respect to the Lebesgue measure of the unique triangular elements $(\dv{X})^\wedge$.
-
-\subsection{Cholesky parameterization}\label{cholesky}
-
-Given a positive (semi-)definite matrix $X$ of rank $K$, we can uniquely decompose it as $X = L L^\top$ where $L$ is an $N \times N$ lower triangular matrix with $L_{jj} > 0$ for all $j \le K$ and $L_{ij} = 0$ for all $j > K$.
-$L$ has $\frac{K(K+1)}{2} + (N - K)K$ degrees of freedom.
-When $K=N$, $X$ is positive-definite.
-
-We can construct $X$ by composing 2 bijective maps.
-The first, $f_1$, enforces the constraint of the positive diagonal:
-
-\[L_{ij} = f_1(Y)_{ij} = \begin{cases} \exp(Y_{ij}) & \text{ if } i = j \le K \\ Y_{ij} & \text{otherwise} \end{cases}\]
-
-The Jacobian determinant for this map is $\absdet{J_{f_1}(Y)} = \exp(\sum_{i=1}^K Y_{ii})$.
-
-The second map is $f_2: L \mapsto L L^\top = X$.
-To compute the Jacobian of the map, we use the half-vectorization $\vectu(X)$.
-Note that $X_{ij} = \sum_{k=1}^{\min(i, j)} L_{ik} L_{jk}$.
-$X_{ij}$ is thus only dependent on elements of $L$ that are below and to the left of the $(i, j)$ position.
-As a result, the Jacobian of $\vectu \circ f_2 \circ \vectu^{-1}$ is triangular, and its determinant is only the product of its diagonal elements: $\absdet{J_{f_2}} = \prod_{i=1}^N \prod_{j=1}^i \frac{\partial X_{ij}}{\partial L_{ij}}$.
-
-In this index range, $\frac{\partial X_{ij}}{\partial L_{ij}} = (1 + \delta_{ij}) L_{ii}$.
-As a result
-
-\[
-  \absdet{J_{f_2}(Z)} = 2^N \prod_{i = 1}^N \prod_{j = 1}^i L_{ii} = 2^N \prod_{i=1}^N |L_{ii}|^{N - i + 1} \\
-\]
-
-Composing the two bijective maps as $f = f_2 \circ f_1$, the resulting Jacobian determinant is
-
-\[
-  \absdet{J_{f}(Y)} = 2^N \exp\left( \sum_{i=1}^N (N - i + 2) Y_{ii}\right).
-\]
-
-\subsection{Matrix exponential parameterization}
-
-For every PD matrix $X \in \mathbb{R}^{N \times N}$, there is a unique symmetric matrix $Y$ such that $X = \Exp(Y)$, where $\operatorname{Exp}(\cdot)$ is the matrix exponential, defined as
-\[\Exp(Y) = \sum_{k=0}^\infty \frac{1}{k!} Y^k.\]
-
-Let $Y = U \Lambda U^\top$ be the symmetric eigendecomposition of $Y$, where the matrix of eigenvectors $U$ is orthogonal, and the $\Lambda = \operatorname{diag}(\lambda)$ for $\lambda \in \mathbb{R}^N$.
-Then $Y^2 = U \Lambda U^\top U \Lambda U^\top = U \Lambda^2 U^\top$.
-In fact, $Y^c = U \Lambda^c U^\top$ for all real $c$.
-As a result,
-\[\Exp(Y) = U \Exp(\Lambda) U^\top,\]
-where $\Exp(\Lambda)_{ij} = \exp(\lambda_i) \delta_{ij}$.
-
-$\Exp(Y)$ is one example of a matrix function $f$, which we can roughly define for symmetric matrices as any function that obeys
-\[f(Y) = U f(\Lambda) U^\top,\]
-where $f(\Lambda)_{ij} = f(\lambda_i)\delta_{ij}$.
-\footnote{
-  In a similar way, we can generalize other scalar functions like the inverse, trigonometric, and hyperbolic functions to functions of matrices.
-  All results in this section apply to these functions as well.
-}
-
-For such functions $f$, we can compute a common form of the Jacobian determinant.
-Differentiating the expression, we find
-\[
-\begin{aligned}
-  \dv{(f(Y))} &= \dv{U} f(\Lambda) U^\top + U f'(\Lambda) \dv{\Lambda} U^\top + U f(\Lambda) \dv{U}^\top\\
-  U^\top \dv{(f(Y))} U &= U^\top \dv{U} f(\Lambda) + f(\Lambda) \dv{U}^\top U + f'(\Lambda)\dv{\Lambda},
-\end{aligned}
-\]
-where $f'(\Lambda)_{ij} = \frac{\dv{(f(\lambda_i))}}{\dv{\lambda_i}}$.
-
-Recall that $U^\top \dv{U}$ is skew-symmetric.
-Note also that for skew-symmetrix $Z = -Z^\top$ and diagonal matrix $\operatorname{diag}(v)$,
-\[(Z\operatorname{diag}(v) + \operatorname{diag}(v)Z^\top)_{ij} = Z_{ij} v_j + v_i Z_{ji} = Z_{ij} v_j - v_i Z_{ij} = Z_{ij} (v_j - v_i) = Z \circ V,\]
-where $V_{ij} = v_j - v_i$ and $\circ$ is the Hadamard (elementwise) product.
-
-From the right-hand side, note then that
-\[(U^\top \dv{(f(Y))} U)_{ij} = \begin{cases} (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) & \text{ if } i \ne j \\ f'(\lambda_i) \dv{f(\lambda_i)} &\text{ if } i = j \end{cases}\]
-
-The exterior product of the diagonal part is
-\[
-  \bigwedge_{i=1}^N f'(\lambda_i) \dv{\lambda_i} = \left(\prod_{i=1}^N f'(\lambda_i)\right) (\dv{\Lambda})^\wedge,
-\]
-while the exterior product of the sub-diagonal part is
-\[
-  \bigwedge_{i=1}^N \bigwedge_{j=i+1}^N (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) = \left( \prod_{i=1}^N \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i)) \right) \left(U^\top \dv{U} \right)^\wedge
-\]
-
-Lastly, note for any symmetric $Z$ and orthogonal matrix $U$,
-\[(U^\top \dv{Z} U)^\wedge = \det(U^\top)^N \det(U)^N (\dv{Z})^\wedge = (\dv{Z})^\wedge.\]
-
-As a result, we can compute $(\dv(f(Y)))^\wedge$ by taking the exterior product of the sub-diagonal and diagonal parts: 
-\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i))\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]
-
-If $f$ is the identity map $f(Y) = Y$, then
-\[ (\dv{Y})^\wedge = \left(\prod_{i=1}^N \prod_{j=1}^{i-1} (\lambda_j - \lambda_i)\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]
-
-Combining the latter two expressions:
-\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} \frac{f(\lambda_j) - f(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge\]
-
-[REF, Edelman's notes] previously found this result.
-
-For $X = f(Y) = \Exp(Y)$ then,
-\[ (\dv{X})^\wedge = \left(\prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge,\]
-
-so the Jacobian determinant of the exponential map is $\absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right|$.
-
-Note that this expression is undefined when any eigenvalues are repeated.
-This is not a problem for MCMC, since the eigenvalues are non-unique with probability 0, but non-unique eigenvalues could occur when maximizing a log-density.
-
-Because $\lim_{h \to 0} \frac{f(x) - f(x + h)}{h} = f'(x)$, we can then modify the expression to handle such cases:
-\[
-  \absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \prod_{j=1}^i \begin{cases}
-    \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right| &\text{ if } \lambda_i \ne \lambda_j \\ 
-    \exp(\lambda_i) &\text{ if } \lambda_i = \lambda_j
-  \end{cases}.
-\]
-
-If the eigenvalues are sorted in ascending order, then the log Jacobian determinant can be more stably computed as
-
-\[
-  \log\absdet{J_{\Exp}(Y)} = \sum_{j=1}^N j\lambda_j + \sum_{i=1}^j \begin{cases}
-    \log(1 - \exp(\lambda_i - \lambda_j)) - \log(\lambda_j - \lambda_i) &\text{ if } \lambda_i \ne \lambda_j \\ 
-    0 &\text{ if } \lambda_i = \lambda_j
-  \end{cases}.
-\]
-
-While the cost of computing the Jacobian determinant is $O(N^2)$, the cost of computing the eigendecomposition naively is $O(N^3)$, so this transformation is expensive.
-An attractive alternative would be to define the transform directly from the eigendecomposition of $X$.
-
-\subsection{Eigendecomposition parameterization}
-
-To define a transform directly from the symmetric eigendecomposition of $X$, we first need to note that the eigendecomposition is not unique; if we swap any two columns $i$ and $j$ of $U$ as well as swapping the eigenvalues $\lambda_i$ and $\lambda_j$, the result is the same matrix.
-Moreover, if we reverse the sign of any given eigenvector, it has no effect on the eigenvalues or $X$.
-
-When the eigenvalues are all unique, we can make the eigendecomposition also unique by enforcing an ascending order of eigenvalues $0 < \lambda_i < \lambda_j$ for $j > i$ and by forcing the first entry of each eigenvector to be positive.
-
-So we can construct a bijective map with the maps $f_1$ that produces an ordered vector of positive eigenvalues and $f_2$ that produces an orthogonal matrix whose first row contains only positive entries.
-Consider $\lambda = f_1(y_1)$ and $U = f_2(y_2)$.
-
-Let $f_1$ be the map that computes each $\lambda_i$ as
-\[
-  \lambda_i = \begin{cases}
-   \exp\bigl((y_1)_1\bigr) &\text{ if } i = 1\\
-    \lambda_{i-1} + \exp\bigl((y_1)_i\bigr) &\text{ if } 1 < i \le N
-  \end{cases}
-\]
-The Jacobian of this map is triangular, so its determinant is $\absdet{J_{f_1}(y_1)} = \exp\left(\sum_{i=1}^N (y_1)_i\right)$.
-
-We can then choose for $f_2$ any of the transforms to the orthogonal matrices previously discussed, though care must be taken to avoid multimodality.
-For example, if the transform covers the entire orthogonal group, then $2^N$ matrices map to the same orthogonal matrix with positive first row, which will manifest as $2^N$ modes in the unconstrained distribution.
-If, however, we can modify one of the transforms to natively produce only matrices with this constraint, then we can eliminate these undesirable modes.
-
-The combined Jacobian determinant of $f: (y_1, y_2) \mapsto f_2(y_2) \operatorname{diag}(f_1(y_1)) f_2(y_2)^\top = X$ is then 
-\[ \absdet{J_{f}(y_1, y_2)} = \absdet{J_{f_2}(y_2)} \exp\left(\sum_{i=1}^N (y_1)_i\right).\]
+\begin{figure}
+    \centering
+    \includegraphics[width=1.2\textwidth]{figures/rmse.png}
+    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
+    \label{fig:my_label}
+\end{figure}
 

### 2022-07-27 dummypaper.tex
intro and transforms updates
sync

@@
 \usepackage{graphicx}
-
 \usepackage{hyperref}
@@
   Parameter Transforms for Hamiltonian Monte Carlo}
-\author{Meenal Jhajaria \\ \small Flatiron Institute 
+\author{Meenal Jhajharia \\ \small Flatiron Institute 
 \and Seth Axen \\ \small University of T\"ubingen 
@@
 
-Although it's possible to write HMC samplers that work for simply
+Although it is possible to write HMC samplers that work for simply
 constrained values such as upper- and/or lower-bounds
@@
 the constrained values to unconstrained values before sampling
-\cite{}.  This presents the issue of selecting which transform to use
+\cite{JSSv076i01}.
+For example, the Probabilistic  Programming language Stan, defines distributions with an unconstrained support, in favour of making sampling an easier process.
+The variables with constraints (or with constrained support in this case) are transformed to an unconstrained space.
+Inverse transforms of these variables are used for log density adjustments.
+These are usually in the form of a Jacobian matrix comprised of the gradient $\frac{\partial x}{\partial y}$, where $x$ and $y$ are the constrained and unconstrained parameters respectively.
+
+These gradients are computationally expensive, certain transforms are entirely inefficient, and some work well on certain distributions or parametrizations and vice versa.
+This presents the issue of selecting which transform to use
 among an infinite set of options, which is the topic of this paper.
+We begin by defining the general theory of transforms, followed by transforms from a unit simplex $\Delta^N$ to the $R^M$.
+Finally, we discuss notions of "better" transforms defined on the simplex, using statistical measures like Effective Sample Size, Leapfrog steps taken by the sampler etc.
 
-%Sampling algorithms play an important role in Statistics, in this case
-%we refer to drawing samples from a probability density or distribution
-%(not survey sampling). Monte Carlo Markov chain methods are commonly
-%used for high dimensional sampling. In this paper, we look at
-%Hamiltonian Monte Carlo, which uses hamiltonian dynamics to propose
-%samples for the Metropolis algorithm. Bayesian Inference often uses
-%parameters with constraints, but HMC optimizes on the unconstrained
-%space. It is difficult to use Monte Carlo estimation on a constrained
-%domain ~\cite{neal2008optimal}
+\section{Unconstraining Transforms}
 
-\section{Transforms}
-
-To draw samples from a distribution $p_X(x)$, defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y}=\mathbb{R}^m$ for $m \ge n$.
-
-Let $f\colon \mathcal{Y} \to \mathcal{X}$ be a smooth and continuous map.
-Only when $m = n$ can $f$ be bijective.
-When $m > n$, multiple values of $y \in \mathcal{Y}$ may map to a given value of $x = f(y)$.
-In this case, to make the function bijective, we consider an additional smooth and continuous map $g\colon \mathcal{Y} \to \mathcal{Z}$, where $\mathcal{Z}$ can be uniquely parameterized by $m - n$ real degrees of freedom.
-We then define $h\colon \mathcal{Y} \to \mathcal{X} \times \mathcal{Z}: y \mapsto (f(y), g(y)) = (x, z)$.
-$f$ and $g$ must be chosen so that $h$ is bijective almost-everywhere;
-that is, the values of $y$ for which $h$ is non-bijective must have no probability mass.
-When $h$ is non-bijective at single points, we call these singularities.
-While singularities themselves are in general not a problem for sampling with MCMC, the region of a target density near the singularity tends to have high curvature.
-
-Given a proper joint distribution with a smooth and continuous density $p_{X,Y}(x, y | \theta)$ for some parameters $\theta$, we can use the usual change of variables formula to write
-
-\[
-  p_Y(y | \theta) = p_{X,Z}(f(y), g(y) | \theta) |J_h(y)|.
-\]
-
-To sample from a target density $p_X(x | \theta)$, let $p_{X,Z}(x, z | \theta) = p_X(x | \theta) p_Z(z | x, \theta)$.
-Then we must choose some proper distribution $p_Z$ that is smooth and continuous.
-Given that choice, we find that
-\[
-  p_Y(y | \theta) = p_X(f(y) | \theta) p_Z(g(y) | f(y), \theta) |J_h(y)|.
-\]
-
-Note that when $m > n$, it is insufficient to specify a transform $h$.
-One must also select a prior $p_Z$, and it is desirable to establish some general properties that a given pair $h$ and $p_Z$ should have.
-In addition to the above properties, the optimal pair would
-1. discard as few degrees of freedom $|m - n|$ as possible.
-2. produce a $p_Z$ and $|J_y|$ and their gradients that are efficient to compute.
-3. produce a $p_Y$ that is easy to sample from.
-
-We can expand a bit on the latter point.
-As a general rule, the more normal $p_Y$ is, the easier it is to sample from (REF?).
-Additionally, near singularities, small perturbations to $y$ tend to result in larger changes to $h(y)$ than far from the singularity.
-This warps the volume around the singularity, which causes $|J_y(y)|$ to smoothly approach infinity near the singularity.
-If $m$ is large, this is generally not a problem, as the "curse of dimensionality" works in our favor, and there is exponentially more volume away from the singularity than near the singularity.
-However, for low $m$, it is necessary that $p_{X,Z}$ goes to 0 near the singularity faster than $|J_y(y)|$ goes to infinity.
-While $p_X$ is generally fixed by the user-specified model and may not have this necessary property, the user may then choose a $p_Z$ that has this property.
-
-\subsection{Changes of Variables}
-
-If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
-smooth, monotonic, bijection $f$, then
+To draw samples from a parameter defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y} \in \mathbb{R}^m$ for $m \ge n$.
+Let $p_Y(y)$ be an improper density function defined over $\mathcal{Y} \rightarrow(0, \infty)$ with support over all elements of $\mathcal{Y}$ and a surjective, smooth and continuous map $g: \mathcal{Y} \rightarrow \mathcal{X}$.
+Then, for any proper density $\pi_{X}$ over $\mathcal{X}$, the constraining transform can be defined as $\left(p_{Y}, g\right)$: $\mathcal{Y} \rightarrow \mathcal{X}$.
+In this case, the density $p_{Y}(y)=\pi_{Y}(y) \cdot \pi_{X}(g(y))$ is proper, and $Y \sim \pi_Y \implies g(Y) \sim \pi_X$.
+The required change of variables for this unconstraining transform is:
 \[
@@
 \]
-and
-\[
-  \absdet{J_{f^{-1}}(y)}
-  = \abs{\det J_{f^{-1}}(y)}.
-\]
 
-\subsection{Computing Jacobian}
-
-To compute Jacobians of functions with matrix-valued inputs or outputs, we need to implicitly or explicitly choose for the inputs and outputs a set of coordinates with dimension equal to the number of degrees of freedom of the matrix
-Unless otherwise specified, we choose the coordinates to be the vectorization of the matrix.
-
-We define the (bijective) vectorization of a matrix $X \in \mathbb{R}^{N \times K}$ with columns $x_k$ for $k \in \{1, ..., K\}$ as the vector
-\[\vect(X) = \begin{bmatrix}x_1 \\ x_2 \\ \vdots \\ x_M\end{bmatrix}\]
-formed by stacking the columns of $X$.
-
-For $X \in \mathbb{R}^{N \times K}$, let $\vectu(X)$ be the map that stacks only the parts of the columns of $X$ above and including the diagonal.
-This map is sometimes called the half-vectorization and is bijective for upper-triangular or symmetric $X$.
-
-Similarly, $\vectu_-(X)$ extracts the same elements except the diagonal.
-When $X$ is strictly upper-triangular or skew-symmetric (i.e. $X = -X^\top$), then this map is bijective.
-
-Here we introduce several strategies for computing Jacobian determinants.
-
-The vectorization has the useful property that for all matrices $A$, $X$, and $B$, $\vect(AYB) = (B^\top \otimes A) \vect(Y)$, where $\otimes$ is the Kronecker product.
-This allows us to convert expressions of matrices to expressions of vectors, from which the Jacobian can be read off.
-For example, let $X = AYB$. Then $\dv{X} = A \dv{Y} B$, and $\vect(\dv{X}) = (B^\top \otimes A) \vect(\dv{Y})$, and the Jacobian is $J = B^\top \otimes A$.
+In this paper we consider transforms where $m = n$, for which the constraining transform is defined as $(\pi_Y, f)$: $\mathcal{Y} \rightarrow \mathcal{X}$, given that $\pi_{Y}(y)=\absdet{J_{f^{-1}}(y)}$.
 
@@
 
+%A simplex is a set of non negative vectors that sum up to a constant $a$, this can be written as:
+%\begin{equation*}
+%\Delta=\left\{\mathbf{x} \in \mathbb{R}^{n} \mid\left\langle\mathbf{x}, \mathbf{1}_{n}\right\rangle=a, \mathbf{x} \geq 0\right\}
+%\end{equation*}
+%
+%In computational works, we generally care about the unit simplex, this is the special case where $ a = 1$.
+Geometrically, an \textit{n-simplex} is the convex closure of $n+1$ affinely independent set of points $x_0, x_1, \dots x_n$.
+It is the simplest polytope for $n$-dimensions (in this case we will be looking at a simplex as a convex polytope, not abstract).
+An \textit{n-simplex} is be represented in the barycentric coordinate system with weighted vertices, as follows:
+%\begin{equation*}
+%\Delta = \left\{x_{0} u_{0}+\cdots+x_{n} u_{n} \mid \sum_{i=0}^{n} x_{i}=1\\
+% \text{ and } x_{i} \geq 0 \text { for } i=0, \ldots, k\right\}
+%\end{equation*}
+%
+%With this representation, taking $u_{i}=1$, the $n$ unit simplex can be written as: 
+%\begin{equation*}
+%\left\{x \in \mathbb{R}^{n}: x_{0}+\cdots+x_{n-1}=1, x_{i} \geq 0 \text { for } i=0, \ldots, k-1\right\}
+%\end{equation*}
+%
+%A \textit{unit simplex} on $R^n$ is an $n - 1$ dimensional simplex with unit vectors as vertices. Notice the unit weights, this set only has $n-1$ affinely independent points.
+%
+%Sum to one constraints are useful in probabilistic inference, often used for modeling categorical responses.
+The rest of the section describes various transforms used to map a simplex to an unconstrained space ($R^n$).\\
+%\\
+
 A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
-values that sums to one.  As such, there are only $N$ degrees of
-freedom, because if $x$ is an $N$-simplex, then
+values that sums to one.  Simplexes are useful for representations of multinomial probabilities
+(e.g., probabilities of categories in a classification problem). The set of unit $N$-simplexes is conventionally denoted
 \[
-  x_N = -(x_1 + x_2 + \cdots + x_{N-1}).
-\]
-Simplexes are useful for representations of multinomial probabilities
-(e.g., probabilities of categories in a classification problem).
-
-The set of unit $N$-simplexes is conventionally denoted
-\[
-  \Delta^N = \setcomp{x \in \mathbb{R}^{N + 1}}{\textrm{sum}(x) = 1
-    \textrm{ and }
-    x_n \geq 0 \textrm{ for } n \in \rngto{N}}
+  \Delta^N = \setcomp{x \in \mathbb{R}_{\ge 0}^{N + 1}}{ \sum_{i=1}^{N} x_i = 1}
 \]
@@
 \begin{bmatrix} 0 & 1 & 0 \end{bmatrix}$,
-and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$.
-
+and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$. As such, there are only $N$ degrees of
+freedom, because if $x$ is an $N$-simplex, then
+\[
+  x_N = 1 - (x_1 + x_2 + \cdots + x_{N-1}).
+\]
+We will use $\Delta^N_-$ to denote $N-1$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
 \subsection{StickBreaking Transform}
 
-The StickBreaking transform ican be be understood from the stick-breaking construction for Dirichlet \cite{sethurman}.
-Intuitively, this comprises of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
+The StickBreaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
+It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
 Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  R^{N-1}$, for $1 \leq i \leq N$ as:
@@
 = \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
-   \right) \text{for break proportion} \, 
-   z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
+   \right) 
    \]
 
+for break proportion 
+\[ 
+ z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
+\]\\
+
 The inverse transform $ f^{-1} \colon R^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  \Delta^{N-1}$ is defined as:
+
 \[
-x_i =
-\left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right) \text{for break proportion} \, z_i = \mathrm{logit}^{-1} \left( y_i
-                             + \log \left( \frac{1}{N - i}
-                                            \right)\right).
+x_i = \left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right)\\
 \]
 
+for break proportion \[z_i = \mathrm{logit}^{-1} \left( y_i  + \log \left( \frac{1}{N - i}
+                                            \right)\right)
+                                            \]\\
+
 An $N$ dimensional unit simplex $\Delta^{N-1}$ has $N-1$ degrees of freedom (Notice that this inverse transform only maps the first $N-1$ elements, the last element of the simplex $x_{N} = 1 - \sum_1^{N-1}{x_i}$).
-The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix.
-Much like the alr transform, for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
+The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix, so for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
 \begin{align*}
@@
   1 - \sum_{k' = 1}^{k-1} x_{k'}
-   \right) z_k (1 - z_k),
-\end{align*}
-
-Absolute determinant of the diagonal matrix $\mathbf{J}$ is the product of its diagonal entries:
-\begin{align*}
-	\abs{\, det \, \textbf{J} \,} = \prod_{i=1}^{N-1} \textbf{J}_{i,i}
+   \right) z_k (1 - z_k),\\
+   \abs{\, det \, \textbf{J} \,} &= \prod_{i=1}^{N-1} \textbf{J}_{i,i}
 \end{align*}
@@
     \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
-  \end{bmatrix}.
+  \end{bmatrix}
 \]
-
 The inverse additive log ratio transform maps values in
@@
 \[
-  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
+  \textrm{softmax}(u) = \frac{e^u}{\sum{e^u}}
 \]
-
-To calculate the change of variables adjustment, we consider only the
-first $N-1$ coordinates of the result, because the last is defined in
-terms of the first $N-1$.  For convenience, we define a function 
-$s:\mathbb{R}^{N-1} \rightarrow \mathbb{R}^{N-1}$ that operates on
-$N-1$ unconstrained variables and returns the first $N-1$ components
-of the $\textrm{alr}^{-1}(y)$, which is
-defined for $y \in \mathbb{R}^{N-1}$ by
+Again, the change of variables adjustment is calculated for $N-1$ dimensions.  
 \[
-  s(y) = \frac{\exp(y)}{\textrm{sum}(\exp(y)) + 1}
-  = \begin{bmatrix}
+\begin{bmatrix}
     \textrm{alr}^{-1}(y)_1
@@
     \textrm{alr}^{-1}(y)_{N-1}
-    \end{bmatrix}.
+    \end{bmatrix} = \frac{e^y}{\sum e^y + 1}
 \]
-Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
-we can transform to a density over unconstrained parameters $y \in
+From a density $p_X(x)$ defined over $x \in \Delta^{N-1}$,
+we can obtain a density over unconstrained parameters $y \in
 \mathbb{R}^{N-1}$ by applying the inverse ALR transform and adjusting
-for the change of variables, which yields
+for the change of variables
 \[
-  p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{s}(y)},
+  p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{f^{-1}}(y)}
 \]
-where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
-and $\absdet{J_s(y)}$ is the absolute value of its determinant.
-
-\subsubsection{Softmax Transform}
-
-To calculate the determinant of the Jacobian of the inverse transform,
-we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
-$\textrm{exp}$ is the elementwise exponential function and
-\textrm{norm} is defined by
+where $J_{f^{-1}}$ is the $N-1$ dimensional Jacobian matrix. To calculate the determinant of the Jacobian of the inverse transform,
+we start by noting that $alr^{-1}(y) = f \circ g$, where
+$f(y) = e^{ \circ y}$ (elementwise exponential), and $g(z)=\frac{z}{\sum z + 1}$. Here $z = e^{\circ y} \in R_{+}^{N-1}$, at this point we evaluate the Jacobian as:
 \[
-  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
+  \absdet{J_{alr^{-1}}(y)}
+  = \absdet{J_{f(y)}} \absdet{J_{g(z)}}
 \]
-As such, the resulting Jacobian determinant is the product of the
-Jacobian determinants of the component functions,
+Note that $J_{f(z)}$ only consists of diagonal entries $e^{y_{i}}$
 \[
-  \absdet{J_s(y)}
-  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
+  \absdet{J_{f(y)}} = \prod_i^n e^{y_i}
 \]
-where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
-function is diagonal, so the determinant is the product of the
-diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
+For $g$ we begin with evaluating $ J_{g} = \frac{\partial g(z)}{\partial z}$. Note that $\mathbf{1}$ is a vector of ones.
 \[
-  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
+ J_g = \frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I}_{N-1}
+  - \left(\frac{1}{1 + \sum e_{i=1}^{N} z_i^2} x \right)
+  \mathbf{1}_{N-1}
 \]
-As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
-$\textrm{norm}$ to derive the Jacobian,
-\[
-  J_{\textrm{norm}}
-  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
-  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
-  \textrm{vector}_{N-1}(1)^{\top},
-\]
-where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
-$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
-the matrix determinant lemma,\footnote{The matrix determinant lemma
+Using the matrix determinant lemma,\footnote{The matrix determinant lemma
   is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
     \textrm{det}(A).\]}
-we have
 \begin{eqnarray*}
-  \textrm{absdet}(J_{\textrm{norm}}(z))
+  |J_{g}|
   & = &
@@
     1
-    + \textrm{vector}_{N-1}(1)^{\top}
-    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
-    \frac{-z}{(1 + \textrm{sum}(z))^2}
+    + \mathbf{1}
+    \left(\frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I} \right)^{-1}
+    \frac{-z}{(1 + \sum e_{i=1}^{N} z_i)^2}
     \right)
-    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
-        \right)
+    \abs{\frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I}}
   \\[6pt]
-  & = &
-  \left(
-    1 
-    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
-        \textrm{sum}(z))^2} \right)
-  \right)
-        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
-  \\[6pt]
-  & = &
-        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
-        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
-  \\[6pt]
-  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
-        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
-  \\[6pt]
-  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
+
+  & = & \left( \frac{1}{1 + \sum e_{i=1}^{N} z_i} \right)^N.
 \end{eqnarray*} 
-Thus the entire absolute determinant of the Jacobian is defined by the
-product, 
+Finally,
 \[
-  \absdet{J_s(y)}
+  \absdet{J_{alr^{-1}}(y)}
   \ = \
-  \textrm{prod}(\exp(y))
-  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
+ \prod_i^n e^{y_i}
+  \, \left( \frac{1}{1 + \sum e_{i=1}^{N}y_i} \right)^N.
 \]
-and our final expression for densities for unconstrained $y \in
-\mathbb{R}^{N-1}$ is
+Putting $J_{alr^{-1}}$ in the change of variables adjustment, we get
 \[
@@
   = p_X(\textrm{alr}^{-1}(y))
-  \, \textrm{prod}(\exp(y))
-  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
+  \, \prod_{i=1}^{N} e^y_i)
+  \left( \frac{1}{1 + \sum_{i=1}^{N} e^y_i}\right)^N
 \]
@@
 \subsection{Simplex softmax parameterization}
-
-Let $\Delta^n$ indicate the unit $n$-simplex with $n$ positive
-elements that sum to 1 and $\Delta^n_-$ indicate the first $n-1$
-elements, from which the final element can be uniquely determined.
 
@@
     \caption{Effective Sample Size vs Total LeapFrog Steps}
-    \label{fig:hyg}
+    \label{fig:ess}
 \end{figure}
@@
     \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
-    \label{fig:my_label}
+    \label{fig:rmse}
 \end{figure}

### 2022-07-27 dummypaper.tex
updates to simplex part of the paper

@@
 
-%A simplex is a set of non negative vectors that sum up to a constant $a$, this can be written as:
-%\begin{equation*}
-%\Delta=\left\{\mathbf{x} \in \mathbb{R}^{n} \mid\left\langle\mathbf{x}, \mathbf{1}_{n}\right\rangle=a, \mathbf{x} \geq 0\right\}
-%\end{equation*}
-%
-%In computational works, we generally care about the unit simplex, this is the special case where $ a = 1$.
-Geometrically, an \textit{n-simplex} is the convex closure of $n+1$ affinely independent set of points $x_0, x_1, \dots x_n$.
-It is the simplest polytope for $n$-dimensions (in this case we will be looking at a simplex as a convex polytope, not abstract).
-An \textit{n-simplex} is be represented in the barycentric coordinate system with weighted vertices, as follows:
-%\begin{equation*}
-%\Delta = \left\{x_{0} u_{0}+\cdots+x_{n} u_{n} \mid \sum_{i=0}^{n} x_{i}=1\\
-% \text{ and } x_{i} \geq 0 \text { for } i=0, \ldots, k\right\}
-%\end{equation*}
-%
-%With this representation, taking $u_{i}=1$, the $n$ unit simplex can be written as: 
-%\begin{equation*}
-%\left\{x \in \mathbb{R}^{n}: x_{0}+\cdots+x_{n-1}=1, x_{i} \geq 0 \text { for } i=0, \ldots, k-1\right\}
-%\end{equation*}
-%
-%A \textit{unit simplex} on $R^n$ is an $n - 1$ dimensional simplex with unit vectors as vertices. Notice the unit weights, this set only has $n-1$ affinely independent points.
-%
-%Sum to one constraints are useful in probabilistic inference, often used for modeling categorical responses.
-The rest of the section describes various transforms used to map a simplex to an unconstrained space ($R^n$).\\
-%\\
-
 A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
@@
 \]
+%n not n-1
 We will use $\Delta^N_-$ to denote $N-1$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
-\subsection{StickBreaking Transform}
+\subsection{Stick-Breaking Transform}
 
-The StickBreaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
+The Stick-Breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
 It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
-Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  R^{N-1}$, for $1 \leq i \leq N$ as:
+Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \to  R^{N-1}$, for $1 \leq i \leq N$ as:
 \[
@@
 
-The inverse transform $ f^{-1} \colon R^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  \Delta^{N-1}$ is defined as:
+The inverse transform $ f^{-1} \colon R^{N-1} \to \Delta^{N-1}$ is defined as:
 
@@
 
-An $N$ dimensional unit simplex $\Delta^{N-1}$ has $N-1$ degrees of freedom (Notice that this inverse transform only maps the first $N-1$ elements, the last element of the simplex $x_{N} = 1 - \sum_1^{N-1}{x_i}$).
-The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix, so for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
-\begin{align*}
-\mathbf{J}_{i, i} &= \frac{\partial x_i}{\partial y_i}
-=
-\frac{\partial x_i}{\partial z_i} \,
-\frac{\partial z_i}{\partial y_i}\\
-\mathbf{J}_{i, i} &= \left(
-  1 - \sum_{k' = 1}^{k-1} x_{k'}
-   \right) z_k (1 - z_k),\\
-   \abs{\, det \, \textbf{J} \,} &= \prod_{i=1}^{N-1} \textbf{J}_{i,i}
-\end{align*}
-
-The correction term $p_Y(y) = p_X(f^{-1}(y))\,
-\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
+The determinant of the Jacobian
+\[
+   \abs{\mathbf{J}} = \prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right)
+\]
 \subsection{Additive log ratio transform}
-
 The unconstraining transform for the identified softmax is known as
@@
 where for $u \in \mathbb{R}^N$,
+
 \[
-  \textrm{softmax}(u) = \frac{e^u}{\sum{e^u}}
-\]
-Again, the change of variables adjustment is calculated for $N-1$ dimensions.  
-\[
-\begin{bmatrix}
-    \textrm{alr}^{-1}(y)_1
-    & \cdots &
-    \textrm{alr}^{-1}(y)_{N-1}
-    \end{bmatrix} = \frac{e^y}{\sum e^y + 1}
-\]
-From a density $p_X(x)$ defined over $x \in \Delta^{N-1}$,
-we can obtain a density over unconstrained parameters $y \in
-\mathbb{R}^{N-1}$ by applying the inverse ALR transform and adjusting
-for the change of variables
-\[
-  p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{f^{-1}}(y)}
-\]
-where $J_{f^{-1}}$ is the $N-1$ dimensional Jacobian matrix. To calculate the determinant of the Jacobian of the inverse transform,
-we start by noting that $alr^{-1}(y) = f \circ g$, where
-$f(y) = e^{ \circ y}$ (elementwise exponential), and $g(z)=\frac{z}{\sum z + 1}$. Here $z = e^{\circ y} \in R_{+}^{N-1}$, at this point we evaluate the Jacobian as:
-\[
-  \absdet{J_{alr^{-1}}(y)}
-  = \absdet{J_{f(y)}} \absdet{J_{g(z)}}
-\]
-Note that $J_{f(z)}$ only consists of diagonal entries $e^{y_{i}}$
-\[
-  \absdet{J_{f(y)}} = \prod_i^n e^{y_i}
-\]
-For $g$ we begin with evaluating $ J_{g} = \frac{\partial g(z)}{\partial z}$. Note that $\mathbf{1}$ is a vector of ones.
-\[
- J_g = \frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I}_{N-1}
-  - \left(\frac{1}{1 + \sum e_{i=1}^{N} z_i^2} x \right)
-  \mathbf{1}_{N-1}
-\]
-Using the matrix determinant lemma,\footnote{The matrix determinant lemma
-  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
-    \textrm{det}(A).\]}
-\begin{eqnarray*}
-  |J_{g}|
-  & = &
-  \left(
-    1
-    + \mathbf{1}
-    \left(\frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I} \right)^{-1}
-    \frac{-z}{(1 + \sum e_{i=1}^{N} z_i)^2}
-    \right)
-    \abs{\frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I}}
-  \\[6pt]
-
-  & = & \left( \frac{1}{1 + \sum e_{i=1}^{N} z_i} \right)^N.
-\end{eqnarray*} 
-Finally,
-\[
-  \absdet{J_{alr^{-1}}(y)}
-  \ = \
- \prod_i^n e^{y_i}
-  \, \left( \frac{1}{1 + \sum e_{i=1}^{N}y_i} \right)^N.
-\]
-Putting $J_{alr^{-1}}$ in the change of variables adjustment, we get
-\[
-  p_Y(y)
-  = p_X(\textrm{alr}^{-1}(y))
-  \, \prod_{i=1}^{N} e^y_i)
-  \left( \frac{1}{1 + \sum_{i=1}^{N} e^y_i}\right)^N
+  \textrm{softmax}(u) = \frac{exp(u)}{\sum exp(u)}
 \]
 
-%\subsection{Softmax Transform}
-%
-%The softmax function can be understood from Multinomial Logistic
-%Regression employed for predicting probabilities of a Categorically
-%distributed variable. Geometrically it maps $R^K$ to the boundary of a
-%unit K-simplex(it is simply the convex hull of $k+1$ affinely
-%independent points in $R^K$. Essentially it transforms a vector of
-%size K to another vector of size K where the outputs sum to 1. It is
-%worth noting that the mapping is actually from $R^K$ to $R^{K-1}$, so
-%when a vector of size $K$ is transformed the $K_th$ vector is simply
-%$1$- sumof k-1 vectors
-
-\subsection{Simplex softmax parameterization}
+\[
+\textrm{Here, } \abs{J} = \prod \exp(y)
+  \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
+\]
+\subsection{Augmented-Softmax Transform}
 
 We define the transformation
-$\phi: \mathbb{R}^{n-1} \to \Delta^n_-: y \mapsto x_-$, where
-$x=\begin{pmatrix}x_- \\ \frac{1}{r}\end{pmatrix} \in \Delta^n$,
-$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$, and
-$r = 1 + \sum_{i=1}^{n-1} e^{y_i}$.
-
-First we compute the scalar derivatives:
-\[
-\begin{aligned}
-  \frac{\mathrm{d} r}{\mathrm{d} y_j}
-  &= e^{y_j} = r x_j\\
-  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
-  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i}
-  \frac{\mathrm{d} r}{\mathrm{d} y_j}
-  = \delta_{ij} x_i - x_i x_j, \quad 1 \le i \le n-1
-\end{aligned}
-\]
-where
-$\delta_{ij} = \begin{cases} 1 &\text{if } i = j \\ 0
-  &\text{otherwise}\end{cases}$ is the Kronecker delta.
-
-If $\mathrm{diag}(x)$ is the diagonal matrix whose diagonal are the
-elements of $x$, then the Jacobian is
-\[
+$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
+(x_-, r)$, where  $\Delta_{-}^N$ and $x_{-}$  denote $N-1$ elements of the simplex and $x$, respectively. Here $r = \sum_{i=1}^n exp(y_i)$,
+$x_i = \frac{1}{r} exp(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function. If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $n$-vector of ones.\[
   J = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top) \operatorname{diag}(x_-),
 \]
-where $\boldsymbol{1}_n$ is the $n$-vector of ones.
 
@@
 \boldsymbol{1}_{n-1}^\top x_- = 1 - \sum_{i=1}^{n-1} x_i = x_n$, so
-$$\mathrm{correction} = |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$
-
-\subsection{Simplex softmax parameterization through parameter expansion}
-
-We define the transformation
-$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
-(x_-, r)$, where $r = \sum_{i=1}^n e^{y_i}$ and
-$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$..
-
-First we compute the scalar derivatives:
-\[
-\begin{aligned}
-  \frac{\mathrm{d} r}{\mathrm{d} y_j}
-  &= e^{y_j} = r x_j
-  \\
-  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
-  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i} \frac{\mathrm{d} r}{\mathrm{d} y_j} = \delta_{ij} x_i - x_i x_j,
-\end{aligned}
-\]
-which corresponds to the Jacobian
-\[
-  J = \begin{pmatrix}I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top & -x_- \\
-    r \boldsymbol{1}_{n-1}^\top & r \end{pmatrix} \mathrm{diag}(x).
-\]
-
-For invertible $A$, the determinant of the block matrix
-$\begin{pmatrix}A & B \\ C & D\end{pmatrix}$ is $|A| |D-CA^{-1}B|$.  A
-square matrix is invertible iff its determinant is non-yero.  From the
-previous section,
-\[
-  |I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = x_n > 0,
-\]
-so the determinant of the Jacobian is
-\[
-  |J| = x_n \left|r + r \boldsymbol{1}_{n-1}^\top (I_{n-1} - x_-
-    \boldsymbol{1}_{n-1}^\top)^{-1} x_-\right|
-  \prod_{i=1}^n x_i.
-\]
-
-Let $w = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top)^{-1} x_-$. Then,
-\[
-\begin{aligned}
-    w - x_- \sum_{i=1}^{n-1} w_i &= x_-\\
-    w &= x_- \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
-    \sum_{i=1}^{n-1} w_i &= \sum_{i=1}^{n-1} \left( x_- (1 - \sum_{i=1}^{n-1} w_i) \right) = \left(\sum_{i=1}^{n-1} x_i \right) \left(1 - \sum_{i=1}^{n-1} w_i\right) = (1 - x_n)  \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
-    \sum_{i=1}^{n-1} w_i &= \frac{1 - x_n}{x_n} = \frac{1}{x_n} - 1\\
-    w &= x_- \left(1 - \left(\frac{1}{x_n} - 1\right)\right) = \frac{1}{x_n} x_-
-  \end{aligned}
-\]
-
-Then
-\[
-  |J| = x_n r \left|1 + \frac{1}{x_n}\sum_{i=1}^{n-1} x_i\right|
-  \prod_{i=1}^n x_i = r \prod_{i=1}^n x_i
-\]
-
-To keep the target distribution proper, we must select a prior
-distribution $\pi(r)$ for $r$.  If we choose $r \sim \chi_n$, then the
-product of the correction and the density of the prior for $r$ is
-proportional to
-
-\[
-  \mathrm{correction}
-  = \pi(r) |J| = r^n e^{-r^2/2} \prod_{i=1}^n x_i
-  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{2}\left(\sum_{i=1}^n
-      e^{y_i}\right)^2\right).
-\]
-
-Alternatively, if we choose $r \sim \mathrm{Gamma}(n, 1)$, then
-\[
-  \mathrm{correction} = \pi(r) |J| = r^n e^{-r} \prod_{i=1}^n x_i =
-  \exp\left(\sum_{i=1}^n y_i - \sum_{i=1}^n e^{y_i}\right).
-\]
-This latter correction is equivalent to the sampling procedure from
-the Dirichlet distribution with $\alpha_i=1$, where
-$z_i \sim \mathrm{Exponential}(1)$ and
-$y = \frac{z}{\sum_{i=1}^n z_i}$.
-
-Both of these corrections can be captured with the generalization
-\[
-  \mathrm{correction}
-  = \pi(r) |J|
-  = r^n e^{-r^p/p} \prod_{i=1}^n x_i
-  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{p} \left(\sum_{i=1}^n e^{y_i}\right)^p\right),
-\]
-for $p > 0$, which corresponds to $r \sim \text{Generalized-Gamma}(1, n, p)$.
+$$ |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$
 
@@
     \centering
-    \includegraphics[width=1.2\textwidth]{figures/ess.png}
-    \caption{Effective Sample Size vs Total LeapFrog Steps}
-    \label{fig:ess}
+    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_density.png}
+    \caption{Effective Sample Size/Total LeapFrog Steps}
+    \label{fig:ess_density}
 \end{figure}
@@
     \centering
-    \includegraphics[width=1.2\textwidth]{figures/rmse.png}
+    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_cdf.png}
+    \caption{Effective Sample Size/Total LeapFrog Steps}
+    \label{fig:ess_cdf}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1.2\textwidth]{figures/simplex/rmse.png}
     \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
@@
 
+\appendix
+\section{Jacobian Determinants}
+\subsection{Stickbreaking Transform}
+The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix, so for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
+\begin{align*}
+\mathbf{J}_{i, i} &= \frac{\partial x_i}{\partial y_i}
+=
+\frac{\partial x_i}{\partial z_i} \,
+\frac{\partial z_i}{\partial y_i}\\
+\mathbf{J}_{i, i} &= \left(
+  1 - \sum_{k' = 1}^{k-1} x_{k'}
+   \right) z_k (1 - z_k),\\
+   \abs{\textbf{J}} &= \prod_{i=1}^{N-1} \textbf{J}_{i,i}
+\end{align*}
+
+The change of variables adjustment $p_Y(y) = p_X(f^{-1}(y))\,
+\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
+
+\subsection{Additive Log ratio transform}
+To calculate the determinant of the Jacobian of the inverse transform,
+we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
+$\textrm{exp}$ is the elementwise exponential function and
+\textrm{norm} is defined by
+\[
+  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
+\]
+As such, the resulting Jacobian determinant is the product of the
+Jacobian determinants of the component functions,
+\[
+  \absdet{J_s(y)}
+  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
+\]
+where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
+function is diagonal, so the determinant is the product of the
+diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
+\[
+  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
+\]
+As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
+$\textrm{norm}$ to derive the Jacobian,
+\[
+  J_{\textrm{norm}}
+  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
+  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
+  \textrm{vector}_{N-1}(1)^{\top},
+\]
+where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
+$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
+the matrix determinant lemma,\footnote{The matrix determinant lemma
+  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
+    \textrm{det}(A).\]}
+we have
+\begin{eqnarray*}
+  \textrm{absdet}(J_{\textrm{norm}}(z))
+  & = &
+  \left(
+    1
+    + \textrm{vector}_{N-1}(1)^{\top}
+    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
+    \frac{-z}{(1 + \textrm{sum}(z))^2}
+    \right)
+    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
+        \right)
+  \\[6pt]
+  & = &
+  \left(
+    1 
+    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
+        \textrm{sum}(z))^2} \right)
+  \right)
+        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
+  \\[6pt]
+  & = &
+        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
+        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
+  \\[6pt]
+  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
+        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
+  \\[6pt]
+  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
+\end{eqnarray*}
+Thus the entire absolute determinant of the Jacobian is defined by the
+product, 
+\[
+  \absdet{J_s(y)}
+  \ = \
+  \textrm{prod}(\exp(y))
+  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
+\]
+and our final expression for densities for unconstrained $y \in
+\mathbb{R}^{N-1}$ is
+\[
+  p_Y(y)
+  = p_X(\textrm{alr}^{-1}(y))
+  \, \textrm{prod}(\exp(y))
+  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
+\]
+
 \end{document}

### 2022-07-27 appendix.tex
remove dummy and add appendix

--- 2022-07-27 dummypaper.tex
@@
-\documentclass[11pt]{article}
+\begin{appendices}
 
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{cite}
-\usepackage[letterpaper, total={6.5in, 9in}]{geometry}
-\usepackage{mathpazo}
-\usepackage{sourcecodepro}
-\usepackage{graphicx}
-\usepackage{hyperref}
-
-\newcommand{\setcomp}[2]{\left\{ #1 \ \Big|\ #2 \right\}}
-\newcommand{\rngto}[1]{1{:}#1}
-\newcommand{\abs}[1]{\left| #1 \right|}
-\newcommand{\absdet}[1]{\abs{#1}}
-\newcommand{\dv}[1]{\mathrm{d}{#1}}
-\newcommand{\Exp}{\mathrm{Exp}}
-\newcommand{\vect}{\mathrm{vec}}
-\newcommand{\vectu}{\mathrm{vecu}}
-
-\begin{document}
-
-\title{Efficient Unconstraining
-  Parameter Transforms for Hamiltonian Monte Carlo}
-\author{Meenal Jhajharia \\ \small Flatiron Institute 
-\and Seth Axen \\ \small University of T\"ubingen 
-\and Adam Haber \\ \small Weizmann Institute 
-\and Sean Pinkney \\ \small Omnicom Media Group
-\and Bob Carpenter \\ \small Flatiron Institute}
-\date{DRAFT: \today}
-\maketitle
-
-\begin{abstract}
-  \noindent
-  This paper evaluates the statistical and computational
-  efficiency of unconstraining parameter transforms for Hamiltonian
-  Monte Carlo sampling.
-\end{abstract}
-
-\section{Introduction}
-
-In statistical computing, we often need to compute high-dimensional
-integrals over densities $\pi(x)$ (e.g., Bayesian estimation or
-prediction, $p$-value calculations, etc.).  The only black-box
-techniques that work for general high-dimensional integrals are
-Markov chain Monte Carlo (MCMC) methods.  The most effective MCMC
-method in high dimensions is Hamiltonian Monte Carlo (HMC).  HMC works
-by simulating the Hamiltonian dynamics of a fictitious particle
-representing the value being sampled coupled with a momentum term.
-
-Although it is possible to write HMC samplers that work for simply
-constrained values such as upper- and/or lower-bounds
-\cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is
-much more challenging to do the same for complex constraints such as
-simplexes or positive definite matrices or for densities involving
-multiple constrained values.  Instead, it is far more common to map
-the constrained values to unconstrained values before sampling
-\cite{JSSv076i01}.
-For example, the Probabilistic  Programming language Stan, defines distributions with an unconstrained support, in favour of making sampling an easier process.
-The variables with constraints (or with constrained support in this case) are transformed to an unconstrained space.
-Inverse transforms of these variables are used for log density adjustments.
-These are usually in the form of a Jacobian matrix comprised of the gradient $\frac{\partial x}{\partial y}$, where $x$ and $y$ are the constrained and unconstrained parameters respectively.
-
-These gradients are computationally expensive, certain transforms are entirely inefficient, and some work well on certain distributions or parametrizations and vice versa.
-This presents the issue of selecting which transform to use
-among an infinite set of options, which is the topic of this paper.
-We begin by defining the general theory of transforms, followed by transforms from a unit simplex $\Delta^N$ to the $R^M$.
-Finally, we discuss notions of "better" transforms defined on the simplex, using statistical measures like Effective Sample Size, Leapfrog steps taken by the sampler etc.
-
-\section{Unconstraining Transforms}
-
-To draw samples from a parameter defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y} \in \mathbb{R}^m$ for $m \ge n$.
-Let $p_Y(y)$ be an improper density function defined over $\mathcal{Y} \rightarrow(0, \infty)$ with support over all elements of $\mathcal{Y}$ and a surjective, smooth and continuous map $g: \mathcal{Y} \rightarrow \mathcal{X}$.
-Then, for any proper density $\pi_{X}$ over $\mathcal{X}$, the constraining transform can be defined as $\left(p_{Y}, g\right)$: $\mathcal{Y} \rightarrow \mathcal{X}$.
-In this case, the density $p_{Y}(y)=\pi_{Y}(y) \cdot \pi_{X}(g(y))$ is proper, and $Y \sim \pi_Y \implies g(Y) \sim \pi_X$.
-The required change of variables for this unconstraining transform is:
-\[
-  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
-\]
-where the Jacobian of the inverse transform is defined by
-\[
-  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
-\]
-
-In this paper we consider transforms where $m = n$, for which the constraining transform is defined as $(\pi_Y, f)$: $\mathcal{Y} \rightarrow \mathcal{X}$, given that $\pi_{Y}(y)=\absdet{J_{f^{-1}}(y)}$.
-
-\section{Unit simplex}
-
-A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
-values that sums to one.  Simplexes are useful for representations of multinomial probabilities
-(e.g., probabilities of categories in a classification problem). The set of unit $N$-simplexes is conventionally denoted
-\[
-  \Delta^N = \setcomp{x \in \mathbb{R}_{\ge 0}^{N + 1}}{ \sum_{i=1}^{N} x_i = 1}
-\]
-Geometrically, an $N$-simplex is the convex closure of $N+1$ points
-that are 1 in one coordinate and 0 elsewhere.  For example, the
-3-simplex is the complex closure of
-$\begin{bmatrix}1 & 0 & 0 \end{bmatrix},
-\begin{bmatrix} 0 & 1 & 0 \end{bmatrix}$,
-and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$. As such, there are only $N$ degrees of
-freedom, because if $x$ is an $N$-simplex, then
-\[
-  x_N = 1 - (x_1 + x_2 + \cdots + x_{N-1}).
-\]
-%n not n-1
-We will use $\Delta^N_-$ to denote $N-1$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
-\subsection{Stick-Breaking Transform}
-
-The Stick-Breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
-It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
-Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \to  R^{N-1}$, for $1 \leq i \leq N$ as:
-\[
-y_i
-= \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
-   \right) 
-   \]
-
-for break proportion 
-\[ 
- z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
-\]\\
-
-The inverse transform $ f^{-1} \colon R^{N-1} \to \Delta^{N-1}$ is defined as:
-
-\[
-x_i = \left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right)\\
-\]
-
-for break proportion \[z_i = \mathrm{logit}^{-1} \left( y_i  + \log \left( \frac{1}{N - i}
-                                            \right)\right)
-                                            \]\\
-
-The determinant of the Jacobian
-\[
-   \abs{\mathbf{J}} = \prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right)
-\]
-\subsection{Additive log ratio transform}
-The unconstraining transform for the identified softmax is known as
-the additive log ratio (ALR) transform
-\cite{aitchison1982statistical}, which is a bijection
-$\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
-$x \in \Delta^{N-1}$ by
-\[
-  \textrm{alr}(x)
-  = \begin{bmatrix}\displaystyle
-    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
-  \end{bmatrix}
-\]
-The inverse additive log ratio transform maps values in
-$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
-\mathbb{R}^{N-1}$ by
-\[
-  \textrm{alr}^{-1}(y)
-  = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
-\]
-where for $u \in \mathbb{R}^N$,
-
-\[
-  \textrm{softmax}(u) = \frac{exp(u)}{\sum exp(u)}
-\]
-
-\[
-\textrm{Here, } \abs{J} = \prod \exp(y)
-  \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
-\]
-\subsection{Augmented-Softmax Transform}
-
-We define the transformation
-$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
-(x_-, r)$, where  $\Delta_{-}^N$ and $x_{-}$  denote $N-1$ elements of the simplex and $x$, respectively. Here $r = \sum_{i=1}^n exp(y_i)$,
-$x_i = \frac{1}{r} exp(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function. If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $n$-vector of ones.\[
-  J = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top) \operatorname{diag}(x_-),
-\]
-
-Using Sylvester's determinant theorem,
-$|I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = 1 -
-\boldsymbol{1}_{n-1}^\top x_- = 1 - \sum_{i=1}^{n-1} x_i = x_n$, so
-$$ |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$
-
-\section{Results}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_density.png}
-    \caption{Effective Sample Size/Total LeapFrog Steps}
-    \label{fig:ess_density}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_cdf.png}
-    \caption{Effective Sample Size/Total LeapFrog Steps}
-    \label{fig:ess_cdf}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1.2\textwidth]{figures/simplex/rmse.png}
-    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
-    \label{fig:rmse}
-\end{figure}
-
-\subsubsection*{Acknowledgements}
-
-We would like to thank \url{matrixcalculus.org} for providing an
-easy-to-use symbolic matrix derivative calculator.
-
-\bibliography{all}{}
-\bibliographystyle{plain}
-
-\appendix
 \section{Jacobian Determinants}
@@
 \]  
-
-\end{document}
+\end{appendices}

### 2022-07-27 notes.tex
remove dummy and add appendix

--- 2022-07-26 paper.tex
@@
 \usepackage[letterpaper, total={6.5in, 9in}]{geometry}
+\usepackage[dvipsnames]{xcolor}
+\usepackage{hyperref}
+    \hypersetup{
+      colorlinks=true,
+      linkcolor=MidnightBlue,
+      citecolor=MidnightBlue,
+      filecolor=MidnightBlue,
+      urlcolor=MidnightBlue,
+      pdftitle={Efficient Unconstraining Parameter Transforms for HMC},
+      pdfpagemode=FullScreen,
+      pdfstartpage=1,
+      pdfauthor={Meenal Jhajaria et al.}
+    }
+
 \usepackage{mathpazo}
 \usepackage{sourcecodepro}
-
-\usepackage{hyperref}
 
@@
 
-\title{Efficient Unconstraining
-  Parameter Transforms for Hamiltonian Monte Carlo}
-\author{Meenal Jhajaria \\ \small Flatiron Institute 
-\and Seth Axen \\ \small University of T\"ubingen 
-\and Adam Haber \\ \small Weizmann Institute 
-\and Sean Pinkney \\ \small Omnicom Media Group
-\and Bob Carpenter \\ \small Flatiron Institute}
-\date{DRAFT: \today}
-\maketitle
-
-\begin{abstract}
-  \noindent
-  This paper evaluates the statistical and computational
-  efficiency of unconstraining parameter transforms for Hamiltonian
-  Monte Carlo sampling.
-\end{abstract}
-
-\section{Introduction}
-
-In statistical computing, we often need to compute high-dimensional
-integrals over densities $\pi(x)$ (e.g., Bayesian estimation or
-prediction, $p$-value calculations, etc.).  The only black-box
-techniques that work for general high-dimensional integrals are
-Markov chain Monte Carlo (MCMC) methods.  The most effective MCMC
-method in high dimensions is Hamiltonian Monte Carlo (HMC).  HMC works
-by simulating the Hamiltonian dynamics of a fictitious particle
-representing the value being sampled coupled with a momentum term.
-
-Although it's possible to write HMC samplers that work for simply
-constrained values such as upper- and/or lower-bounds
-\cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is
-much more challenging to do the same for complex constraints such as
-simplexes or positive definite matrices or for densities involving
-multiple constrained values.  Instead, it is far more common to map
-the constrained values to unconstrained values before sampling
-\cite{}.  This presents the issue of selecting which transform to use
-among an infinite set of options, which is the topic of this paper.
-
-%Sampling algorithms play an important role in Statistics, in this case
-%we refer to drawing samples from a probability density or distribution
-%(not survey sampling). Monte Carlo Markov chain methods are commonly
-%used for high dimensional sampling. In this paper, we look at
-%Hamiltonian Monte Carlo, which uses hamiltonian dynamics to propose
-%samples for the Metropolis algorithm. Bayesian Inference often uses
-%parameters with constraints, but HMC optimizes on the unconstrained
-%space. It is difficult to use Monte Carlo estimation on a constrained
-%domain ~\cite{neal2008optimal}
-
-\section{Changes of Variables}
-
-If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
-smooth, monotonic, bijection $f$, then
-\[
-  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
-\]
-where the Jacobian of the inverse transform is defined by
-\[
-  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
-\]
-and
-\[
-  \absdet{J_{f^{-1}}(y)}
-  = \abs{\det J_{f^{-1}}(y)}.
-\]
+\title{Notes}
 
@@
 
-The StickBreaking transform ican be be understood from the stick-breaking construction for Dirichlet \cite{sethurman}.
+The StickBreaking transform ican be be understood from the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
 Intuitively, this comprises of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.

### 2022-07-27 paper.tex
remove dummy and add appendix

@@
 \usepackage{sourcecodepro}
-
+\usepackage{graphicx}
 \usepackage{hyperref}
@@
 the constrained values to unconstrained values before sampling
-\cite{}.  This presents the issue of selecting which transform to use
+\cite{JSSv076i01}.
+For example, the Probabilistic  Programming language Stan, defines distributions with an unconstrained support, in favour of making sampling an easier process.
+The variables with constraints (or with constrained support in this case) are transformed to an unconstrained space.
+Inverse transforms of these variables are used for log density adjustments.
+These are usually in the form of a Jacobian matrix comprised of the gradient $\frac{\partial x}{\partial y}$, where $x$ and $y$ are the constrained and unconstrained parameters respectively.
+
+These gradients are computationally expensive, certain transforms are entirely inefficient, and some work well on certain distributions or parametrizations and vice versa.
+This presents the issue of selecting which transform to use
 among an infinite set of options, which is the topic of this paper.
+We begin by defining the general theory of transforms, followed by transforms from a unit simplex $\Delta^N$ to the $R^M$.
+Finally, we discuss notions of "better" transforms defined on the simplex, using statistical measures like Effective Sample Size, Leapfrog steps taken by the sampler etc.
 
-%Sampling algorithms play an important role in Statistics, in this case
-%we refer to drawing samples from a probability density or distribution
-%(not survey sampling). Monte Carlo Markov chain methods are commonly
-%used for high dimensional sampling. In this paper, we look at
-%Hamiltonian Monte Carlo, which uses hamiltonian dynamics to propose
-%samples for the Metropolis algorithm. Bayesian Inference often uses
-%parameters with constraints, but HMC optimizes on the unconstrained
-%space. It is difficult to use Monte Carlo estimation on a constrained
-%domain ~\cite{neal2008optimal}
+\section{Unconstraining Transforms}
 
-\section{Changes of Variables}
-
-If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
-smooth, monotonic, bijection $f$, then
+To draw samples from a parameter defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y} \in \mathbb{R}^m$ for $m \ge n$.
+Let $p_Y(y)$ be an improper density function defined over $\mathcal{Y} \rightarrow(0, \infty)$ with support over all elements of $\mathcal{Y}$ and a surjective, smooth and continuous map $g: \mathcal{Y} \rightarrow \mathcal{X}$.
+Then, for any proper density $\pi_{X}$ over $\mathcal{X}$, the constraining transform can be defined as $\left(p_{Y}, g\right)$: $\mathcal{Y} \rightarrow \mathcal{X}$.
+In this case, the density $p_{Y}(y)=\pi_{Y}(y) \cdot \pi_{X}(g(y))$ is proper, and $Y \sim \pi_Y \implies g(Y) \sim \pi_X$.
+The required change of variables for this unconstraining transform is:
 \[
@@
 \]
-and
-\[
-  \absdet{J_{f^{-1}}(y)}
-  = \abs{\det J_{f^{-1}}(y)}.
-\]
 
-\section{Transforms}
-
-To draw samples from a distribution $p_X(x)$, defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y}=\mathbb{R}^m$ for $m \ge n$.
-
-Let $f\colon \mathcal{Y} \to \mathcal{X}$ be a smooth and continuous map.
-Only when $m = n$ can $f$ be bijective.
-When $m > n$, multiple values of $y \in \mathcal{Y}$ may map to a given value of $x = f(y)$.
-In this case, to make the function bijective, we consider an additional smooth and continuous map $g\colon \mathcal{Y} \to \mathcal{Z}$, where $\mathcal{Z}$ can be uniquely parameterized by $m - n$ real degrees of freedom.
-We then define $h\colon \mathcal{Y} \to \mathcal{X} \times \mathcal{Z}: y \mapsto (f(y), g(y)) = (x, z)$.
-$f$ and $g$ must be chosen so that $h$ is bijective almost-everywhere;
-that is, the values of $y$ for which $h$ is non-bijective must have no probability mass.
-When $h$ is non-bijective at single points, we call these singularities.
-While singularities themselves are in general not a problem for sampling with MCMC, the region of a target density near the singularity tends to have high curvature.
-
-Given a proper joint distribution with a smooth and continuous density $p_{X,Y}(x, y | \theta)$ for some parameters $\theta$, we can use the usual change of variables formula to write
-
-\[
-  p_Y(y | \theta) = p_{X,Z}(f(y), g(y) | \theta) |J_h(y)|.
-\]
-
-To sample from a target density $p_X(x | \theta)$, let $p_{X,Z}(x, z | \theta) = p_X(x | \theta) p_Z(z | x, \theta)$.
-Then we must choose some proper distribution $p_Z$ that is smooth and continuous.
-Given that choice, we find that
-\[
-  p_Y(y | \theta) = p_X(f(y) | \theta) p_Z(g(y) | f(y), \theta) |J_h(y)|.
-\]
-
-Note that when $m > n$, it is insufficient to specify a transform $h$.
-One must also select a prior $p_Z$, and it is desirable to establish some general properties that a given pair $h$ and $p_Z$ should have.
-In addition to the above properties, the optimal pair would
-1. discard as few degrees of freedom $|m - n|$ as possible.
-2. produce a $p_Z$ and $|J_y|$ and their gradients that are efficient to compute.
-3. produce a $p_Y$ that is easy to sample from.
-
-We can expand a bit on the latter point.
-As a general rule, the more normal $p_Y$ is, the easier it is to sample from (REF?).
-Additionally, near singularities, small perturbations to $y$ tend to result in larger changes to $h(y)$ than far from the singularity.
-This warps the volume around the singularity, which causes $|J_y(y)|$ to smoothly approach infinity near the singularity.
-If $m$ is large, this is generally not a problem, as the "curse of dimensionality" works in our favor, and there is exponentially more volume away from the singularity than near the singularity.
-However, for low $m$, it is necessary that $p_{X,Z}$ goes to 0 near the singularity faster than $|J_y(y)|$ goes to infinity.
-While $p_X$ is generally fixed by the user-specified model and may not have this necessary property, the user may then choose a $p_Z$ that has this property.
-
-\section{Preliminaries}
-
-[SA: TODO: eliminate or merge with following section]
-
-To compute Jacobians of functions with matrix-valued inputs or outputs, we need to implicitly or explicitly choose for the inputs and outputs a set of coordinates with dimension equal to the number of degrees of freedom of the matrix
-Unless otherwise specified, we choose the coordinates to be the vectorization of the matrix.
-
-We define the (bijective) vectorization of a matrix $X \in \mathbb{R}^{N \times K}$ with columns $x_k$ for $k \in \{1, ..., K\}$ as the vector
-\[\vect(X) = \begin{bmatrix}x_1 \\ x_2 \\ \vdots \\ x_M\end{bmatrix}\]
-formed by stacking the columns of $X$.
-
-For $X \in \mathbb{R}^{N \times K}$, let $\vectu(X)$ be the map that stacks only the parts of the columns of $X$ above and including the diagonal.
-This map is sometimes called the half-vectorization and is bijective for upper-triangular or symmetric $X$.
-
-Similarly, $\vectu_-(X)$ extracts the same elements except the diagonal.
-When $X$ is strictly upper-triangular or skew-symmetric (i.e. $X = -X^\top$), then this map is bijective.
-
-\section{Computing Jacobian determinants}
-
-[SA: TODO: remove everything we do not use later, add strategy for making non-square Jacobian square]
-
-Here we introduce several strategies for computing Jacobian determinants.
-
-The vectorization has the useful property that for all matrices $A$, $X$, and $B$, $\vect(AYB) = (B^\top \otimes A) \vect(Y)$, where $\otimes$ is the Kronecker product.
-This allows us to convert expressions of matrices to expressions of vectors, from which the Jacobian can be read off.
-For example, let $X = AYB$. Then $\dv{X} = A \dv{Y} B$, and $\vect(\dv{X}) = (B^\top \otimes A) \vect(\dv{Y})$, and the Jacobian is $J = B^\top \otimes A$.
-
-When the previously discussed strategies are insufficient to compute the Jacobian, we bypass the Jacobian computation entirely and use the exterior product $\wedge$.
-For example, let $X \in \mathbb{R}^{N \times K}$.
-The exterior product
-\[
-  \dv{X_{11} }\wedge \dv{X_{21}} \wedge \ldots \wedge \dv{X_{NK}} = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X}_{ij}
-\]
-is another more explicit way of writing the integration measure $\dv{X_{11}} \dv{X_{21}} \dots \dv{X_{NK}}$ that has useful properties for computing Jacobian determinants.
-The basic properties we will use are
-
-\begin{itemize}
-  \item $a\dv{x} \wedge b\dv{y} = (ab)(\dv{x} \wedge \dv{y})$ for $a,b \in \mathbb{R}$
-  \item $(\dv{x} + \dv{y}) \wedge \dv{z} = \dv{x} \wedge \dv{z} + \dv{y} \wedge \dv{z}$
-  \item $\dv{x} \wedge \dv{y} = - \dv{y} \wedge \dv{x}$
-  \item $\dv{x} \wedge \dv{x} = 0$
-\end{itemize}
-
-The connection between the exterior product and the Jacobian determinant is given by the following expression.
-Given $f: y \mapsto x$ for $x,y \in \mathbb{R}^N$,
-\[\bigwedge_{i=1}^N \dv{x_i} = \det(J_f(y)) \bigwedge_{i=1}^N \dv{y_i}.\]
-
-We can then find the Jacobian determinant by first writing all differential forms $\dv{x_i}$ or $\dv{y_i}$ and then computing the exterior product until the expression looks like the above.
-For our purposes, we will ignore the sign of the Jacobian determinant.
-
-A related useful property is that for $X,Y,B \in \mathbb{R}^{N \times K}$, if $\dv{X} = B \dv{Y}$, then
-\[\bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_i} = \det(B)^K \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{Y_i}\]
-
-Following [REF, Edelman], we use the notation $(\cdot)^\wedge$ as a shorthand to represent the exterior product of the unique elements of $\cdot$.
-
-For example,
-\begin{itemize}
-  \item Rectangular matrix $\{X \in \mathbb{R}^{N \times K}\}$: $(\dv{X})^\wedge = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_{ij}}$
-  \item Lower-triangular rectangular matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i < j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{\min(i, K)} \dv{X_{ij}}$
-  \item Diagonal matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i \ne j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^{\min(N, K)} \dv{X_{ii}}$
-  \item Symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = X_{ji}\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^i \dv{X_{ij}}$
-  \item Skew-symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = -X_{ji}, X_{ii} = 0\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{i-1} \dv{X_{ij}}$
-\end{itemize}
-
-Using this notation, for example, $(B \dv{X})^\wedge = \det(B)^K (\dv{X})^\wedge$
+%notion of uniformity missing
+In this paper we consider transforms where $m = n$, for which the constraining transform is defined as $(\pi_Y, f)$: $\mathcal{Y} \rightarrow \mathcal{X}$, given that $\pi_{Y}(y)=\absdet{J_{f^{-1}}(y)}$.
 
@@
 A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
-values that sums to one.  As such, there are only $N$ degrees of
-freedom, because if $x$ is an $N$-simplex, then
+values that sums to one.  Simplexes are useful for representations of multinomial probabilities
+(e.g., probabilities of categories in a classification problem). The set of unit $N$-simplexes is conventionally denoted
 \[
-  x_N = -(x_1 + x_2 + \cdots + x_{N-1}).
-\]
-Simplexes are useful for representations of multinomial probabilities
-(e.g., probabilities of categories in a classification problem).
-
-The set of unit $N$-simplexes is conventionally denoted
-\[
-  \Delta^N = \setcomp{x \in \mathbb{R}^{N + 1}}{\textrm{sum}(x) = 1
-    \textrm{ and }
-    x_n \geq 0 \textrm{ for } n \in \rngto{N}}
+  \Delta^N = \setcomp{x \in \mathbb{R}_{\ge 0}^{N + 1}}{ \sum_{i=1}^{N} x_i = 1}
 \]
@@
 \begin{bmatrix} 0 & 1 & 0 \end{bmatrix}$,
-and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$.
+and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$. As such, there are only $N$ degrees of
+freedom, because if $x$ is an $N$-simplex, then
+\[
+  x_N = 1 - (x_1 + x_2 + \cdots + x_{N-1}).
+\]
+%n not n-1
+We will use $\Delta^N_-$ to denote $N$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
+\subsection{Stick-Breaking Transform}
 
-\subsection{StickBreaking Transform}
-
-The StickBreaking transform ican be be understood from the stick-breaking construction for Dirichlet \cite{sethurman}.
-Intuitively, this comprises of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
-Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  R^{N-1}$, for $1 \leq i \leq N$ as:
+The Stick-Breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
+It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
+Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \to  R^{N-1}$, for $1 \leq i \leq N$ as:
 \[
@@
 = \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
-   \right) \text{for break proportion} \, 
-   z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
+   \right) 
    \]
 
-The inverse transform $ f^{-1} \colon R^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  \Delta^{N-1}$ is defined as:
+for break proportion 
 \[ 
-x_i =
-\left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right) \text{for break proportion} \, z_i = \mathrm{logit}^{-1} \left( y_i
-                             + \log \left( \frac{1}{N - i}
-                                            \right)\right).
+ z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
+\]\\
+
+The inverse transform $ f^{-1} \colon R^{N-1} \to \Delta^{N-1}$ is defined as:
+
+\[
+x_i = \left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right)\\
 \]
 
-An $N$ dimensional unit simplex $\Delta^{N-1}$ has $N-1$ degrees of freedom (Notice that this inverse transform only maps the first $N-1$ elements, the last element of the simplex $x_{N} = 1 - \sum_1^{N-1}{x_i}$).
-The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix.
-Much like the alr transform, for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
-\begin{align*}
-\mathbf{J}_{i, i} &= \frac{\partial x_i}{\partial y_i}
-=
-\frac{\partial x_i}{\partial z_i} \,
-\frac{\partial z_i}{\partial y_i}\\
-\mathbf{J}_{i, i} &= \left(
-  1 - \sum_{k' = 1}^{k-1} x_{k'}
-   \right) z_k (1 - z_k),
-\end{align*}
+for break proportion \[z_i = \mathrm{logit}^{-1} \left( y_i  + \log \left( \frac{1}{N - i}
+                                            \right)\right)
+                                            \]\\
 
-Absolute determinant of the diagonal matrix $\mathbf{J}$ is the product of its diagonal entries:
-\begin{align*}
-	\abs{\, det \, \textbf{J} \,} = \prod_{i=1}^{N-1} \textbf{J}_{i,i}
-\end{align*}
-
-The correction term $p_Y(y) = p_X(f^{-1}(y))\,
-\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
+The determinant of the Jacobian
+\[
+   \abs{\mathbf{J}} = \prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right)
+\]
 \subsection{Additive log ratio transform}
-
 The unconstraining transform for the identified softmax is known as
@@
 where for $u \in \mathbb{R}^N$,
+
 \[
-  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
+  \textrm{softmax}(u) = \frac{exp(u)}{\sum exp(u)}
 \]
 
-To calculate the change of variables adjustment, we consider only the
-first $N-1$ coordinates of the result, because the last is defined in
-terms of the first $N-1$.  For convenience, we define a function 
-$s:\mathbb{R}^{N-1} \rightarrow \mathbb{R}^{N-1}$ that operates on
-$N-1$ unconstrained variables and returns the first $N-1$ components
-of the $\textrm{alr}^{-1}(y)$, which is
-defined for $y \in \mathbb{R}^{N-1}$ by
 \[
-  s(y) = \frac{\exp(y)}{\textrm{sum}(\exp(y)) + 1}
-  = \begin{bmatrix}
-    \textrm{alr}^{-1}(y)_1
-    & \cdots &
-    \textrm{alr}^{-1}(y)_{N-1}
-    \end{bmatrix}.
+\textrm{Here, } \abs{J} = \prod \exp(y)
+  \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
 \]
-Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
-we can transform to a density over unconstrained parameters $y \in
-\mathbb{R}^{N-1}$ by applying the inverse ALR transform and adjusting
-for the change of variables, which yields
-\[
-  p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{s}(y)},
-\]
-where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
-and $\absdet{J_s(y)}$ is the absolute value of its determinant.
-
-\subsubsection{Softmax Transform}
-
-To calculate the determinant of the Jacobian of the inverse transform,
-we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
-$\textrm{exp}$ is the elementwise exponential function and
-\textrm{norm} is defined by
-\[
-  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
-\]
-As such, the resulting Jacobian determinant is the product of the
-Jacobian determinants of the component functions,
-\[
-  \absdet{J_s(y)}
-  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
-\]
-where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
-function is diagonal, so the determinant is the product of the
-diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
-\[
-  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
-\]
-As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
-$\textrm{norm}$ to derive the Jacobian,
-\[
-  J_{\textrm{norm}}
-  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
-  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
-  \textrm{vector}_{N-1}(1)^{\top},
-\]
-where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
-$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
-the matrix determinant lemma,\footnote{The matrix determinant lemma
-  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
-    \textrm{det}(A).\]}
-we have
-\begin{eqnarray*}
-  \textrm{absdet}(J_{\textrm{norm}}(z))
-  & = &
-  \left(
-    1
-    + \textrm{vector}_{N-1}(1)^{\top}
-    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
-    \frac{-z}{(1 + \textrm{sum}(z))^2}
-    \right)
-    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
-        \right)
-  \\[6pt]
-  & = &
-  \left(
-    1 
-    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
-        \textrm{sum}(z))^2} \right)
-  \right)
-        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
-  \\[6pt]
-  & = &
-        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
-        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
-  \\[6pt]
-  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
-        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
-  \\[6pt]
-  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
-\end{eqnarray*}
-Thus the entire absolute determinant of the Jacobian is defined by the
-product, 
-\[
-  \absdet{J_s(y)}
-  \ = \
-  \textrm{prod}(\exp(y))
-  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
-\]
-and our final expression for densities for unconstrained $y \in
-\mathbb{R}^{N-1}$ is
-\[
-  p_Y(y)
-  = p_X(\textrm{alr}^{-1}(y))
-  \, \textrm{prod}(\exp(y))
-  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
+\subsection{Augmented-Softmax Transform}
+% Needs to be written like other transforms -> transform, inverse, |J|
+We define the transformation
+$\phi: \mathbb{R}^n \to \Delta^{N} \times \mathbb{R}_{>0}: y \mapsto
+(x_-, r)$, where  $\Delta_{-}^N$ and $x_{-}$  denote $N$ elements of the simplex and $x$, respectively. Here $r = \sum_{i=1}^n \textrm{exp}(y_i)$,
+$x_i = \frac{1}{r} \textrm{exp}(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function.
+If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $N$-vector of ones.\[
+  J = (I_{N-1} - x_- \boldsymbol{1}_{N-1}^\top) \operatorname{diag}(x_-),
 \]
 
-%\subsection{Softmax Transform}
-%
-%The softmax function can be understood from Multinomial Logistic
-%Regression employed for predicting probabilities of a Categorically
-%distributed variable. Geometrically it maps $R^K$ to the boundary of a
-%unit K-simplex(it is simply the convex hull of $k+1$ affinely
-%independent points in $R^K$. Essentially it transforms a vector of
-%size K to another vector of size K where the outputs sum to 1. It is
-%worth noting that the mapping is actually from $R^K$ to $R^{K-1}$, so
-%when a vector of size $K$ is transformed the $K_th$ vector is simply
-%$1$- sumof k-1 vectors
+Using Sylvester's determinant theorem,
+$|I_{N-1} - x_- \boldsymbol{1}_{N-1}^\top| = 1 -
+\boldsymbol{1}_{N-1}^\top x_- = 1 - \sum_{i=1}^{N-1} x_i = x_N$, so
+$$ |J| = x_n \prod_{i=1}^{N-1} x_i = \prod_{i=1}^{N} x_i = \exp\left(\sum_{i=1}^{N-1} y_i\right) \left(1 + \sum_{i=1}^{N-1} e^{y_i}\right)^{-N}$$
 
-\subsection{Simplex softmax parameterization}
+\section{Results}
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_density.png}
+    \caption{Probability Density plot of Effective Sample Size/Total LeapFrog Steps.
+Effective sample size is evaluated at 100 runs, each point is the value of ESS/Total Leapfrog steps taken by the sampler in one run.}
+    \label{fig:ess_density}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_cdf.png}
+    \caption{Cumulative Density plot of Effective Sample Size/Total LeapFrog Steps.
+Effective sample size is evaluated at 100 runs, each point is the value of ESS/Total Leapfrog steps taken by the sampler in one run.}
+    \label{fig:ess_cdf}
+\end{figure}
 
-Let $\Delta^n$ indicate the unit $n$-simplex with $n$ positive
-elements that sum to 1 and $\Delta^n_-$ indicate the first $n-1$
-elements, from which the final element can be uniquely determined.
-
-We define the transformation
-$\phi: \mathbb{R}^{n-1} \to \Delta^n_-: y \mapsto x_-$, where
-$x=\begin{pmatrix}x_- \\ \frac{1}{r}\end{pmatrix} \in \Delta^n$,
-$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$, and
-$r = 1 + \sum_{i=1}^{n-1} e^{y_i}$.
-
-First we compute the scalar derivatives:
-\[
-\begin{aligned}
-  \frac{\mathrm{d} r}{\mathrm{d} y_j}
-  &= e^{y_j} = r x_j\\
-  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
-  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i}
-  \frac{\mathrm{d} r}{\mathrm{d} y_j}
-  = \delta_{ij} x_i - x_i x_j, \quad 1 \le i \le n-1
-\end{aligned}
-\]
-where
-$\delta_{ij} = \begin{cases} 1 &\text{if } i = j \\ 0
-  &\text{otherwise}\end{cases}$ is the Kronecker delta.
-
-If $\mathrm{diag}(x)$ is the diagonal matrix whose diagonal are the
-elements of $x$, then the Jacobian is
-\[
-  J = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top) \operatorname{diag}(x_-),
-\]
-where $\boldsymbol{1}_n$ is the $n$-vector of ones.
-
-Using Sylvester's determinant theorem,
-$|I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = 1 -
-\boldsymbol{1}_{n-1}^\top x_- = 1 - \sum_{i=1}^{n-1} x_i = x_n$, so
-$$\mathrm{correction} = |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$
-
-\subsection{Simplex softmax parameterization through parameter expansion}
-
-We define the transformation
-$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
-(x_-, r)$, where $r = \sum_{i=1}^n e^{y_i}$ and
-$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$..
-
-First we compute the scalar derivatives:
-\[
-\begin{aligned}
-  \frac{\mathrm{d} r}{\mathrm{d} y_j}
-  &= e^{y_j} = r x_j
-  \\
-  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
-  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i} \frac{\mathrm{d} r}{\mathrm{d} y_j} = \delta_{ij} x_i - x_i x_j,
-\end{aligned}
-\]
-which corresponds to the Jacobian
-\[
-  J = \begin{pmatrix}I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top & -x_- \\
-    r \boldsymbol{1}_{n-1}^\top & r \end{pmatrix} \mathrm{diag}(x).
-\]
-
-For invertible $A$, the determinant of the block matrix
-$\begin{pmatrix}A & B \\ C & D\end{pmatrix}$ is $|A| |D-CA^{-1}B|$.  A
-square matrix is invertible iff its determinant is non-yero.  From the
-previous section,
-\[
-  |I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = x_n > 0,
-\]
-so the determinant of the Jacobian is
-\[
-  |J| = x_n \left|r + r \boldsymbol{1}_{n-1}^\top (I_{n-1} - x_-
-    \boldsymbol{1}_{n-1}^\top)^{-1} x_-\right|
-  \prod_{i=1}^n x_i.
-\]
-
-Let $w = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top)^{-1} x_-$. Then,
-\[
-\begin{aligned}
-    w - x_- \sum_{i=1}^{n-1} w_i &= x_-\\
-    w &= x_- \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
-    \sum_{i=1}^{n-1} w_i &= \sum_{i=1}^{n-1} \left( x_- (1 - \sum_{i=1}^{n-1} w_i) \right) = \left(\sum_{i=1}^{n-1} x_i \right) \left(1 - \sum_{i=1}^{n-1} w_i\right) = (1 - x_n)  \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
-    \sum_{i=1}^{n-1} w_i &= \frac{1 - x_n}{x_n} = \frac{1}{x_n} - 1\\
-    w &= x_- \left(1 - \left(\frac{1}{x_n} - 1\right)\right) = \frac{1}{x_n} x_-
-  \end{aligned}
-\]
-
-Then
-\[
-  |J| = x_n r \left|1 + \frac{1}{x_n}\sum_{i=1}^{n-1} x_i\right|
-  \prod_{i=1}^n x_i = r \prod_{i=1}^n x_i
-\]
-
-To keep the target distribution proper, we must select a prior
-distribution $\pi(r)$ for $r$.  If we choose $r \sim \chi_n$, then the
-product of the correction and the density of the prior for $r$ is
-proportional to
-
-\[
-  \mathrm{correction}
-  = \pi(r) |J| = r^n e^{-r^2/2} \prod_{i=1}^n x_i
-  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{2}\left(\sum_{i=1}^n
-      e^{y_i}\right)^2\right).
-\]
-
-Alternatively, if we choose $r \sim \mathrm{Gamma}(n, 1)$, then
-\[
-  \mathrm{correction} = \pi(r) |J| = r^n e^{-r} \prod_{i=1}^n x_i =
-  \exp\left(\sum_{i=1}^n y_i - \sum_{i=1}^n e^{y_i}\right).
-\]
-This latter correction is equivalent to the sampling procedure from
-the Dirichlet distribution with $\alpha_i=1$, where
-$z_i \sim \mathrm{Exponential}(1)$ and
-$y = \frac{z}{\sum_{i=1}^n z_i}$.
-
-Both of these corrections can be captured with the generalization
-\[
-  \mathrm{correction}
-  = \pi(r) |J|
-  = r^n e^{-r^p/p} \prod_{i=1}^n x_i
-  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{p} \left(\sum_{i=1}^n e^{y_i}\right)^p\right),
-\]
-for $p > 0$, which corresponds to $r \sim \text{Generalized-Gamma}(1, n, p)$.
-
-\section{Semi-orthogonal matrices (Stiefel manifold)}
-
-A matrix $X \in \mathbb{R}^{N \times K}$ for $K \le N$ is semi-orthogonal if and only if $X^\top X = I_K$.
-The semi-orthogonal matrices form a space called the Stiefel manifold.
-When $K=1$, $X$ is a point on the $(N-1)$-sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$.
-We will consider such unit vectors in the next section.
-When $K=N$, $X$ is an element of the orthogonal group $O(N)$ and additionally satisfies $\det(X) = \pm 1$.
-$O(N)$ is comprised of two disconnected submanifolds, one being the Special Orthogonal group $SO(N)$ with $\det(X) = +1$, and the other having $\det(X) = -1$.
-$SO(2)$ and $SO(3)$ are the groups of rotation matrices of 2D and 3D vectors, respectively.
-
-By choosing a set of rules such as Gram-Schmidt orthonormalization of the canonical basis in $\mathbb{R}^{N \times N}$, it is possible to analytically compute from $X$ an orthogonal complement $X^\perp \in \mathbb{R}^{N \times (N - K)}$, such that $\tilde{X}^\top \tilde{X} = I_N$, where $\tilde{X} = \begin{bmatrix}X & X^\perp \end{bmatrix} \in O(N)$.
-
-Note that if we differentiate both sides of the constraint $X^\top X = I_K$, we get $\dv{X}^\top X + X^\top \dv{X} = 0$, which leads to $(X^\top \dv{X})^\top = - X^\top \dv{X}$.
-Therefore, $X^\top \dv{X}$ is a $K \times K$ skew-symmetric matrix with $\frac{K(K-1)}{2}$ unique entries in the strict upper triangle.
-
-Densities of distributions on the Stiefel manifold are usually written with respect to the invariant metric on the Stiefel manifold, which we will define separately for the two cases $K=N$ and $K < N$.
-
-For $X \in O(N)$ with the $i$th column written $x_i$, we define the invariant measure as the exterior product of the unique elements of this skew-symmetric matrix:
-\[
-  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N (X^\top \dv{X})_{ij} = \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j.
-\]
-This measure is called invariant because if $f(X) = A X B$ for $A,B \in O(N)$ and $\mathcal{S} \subset O(N)$ then $\int_{\mathcal{S}} (X^\top \dv{X}) = \int_{f(\mathcal{S})} (X^\top \dv{X})$.
-It is also called the Haar measure on $O(N)$ \cite{muirhead2009aspects}..
-
-To choose an invariant measure for $K < N$, we first choose an $X^\perp$ analytically computed from $X$.
-Because $X^\perp$ is analytically computed, $\dv{X^\perp_{ij}} = 0$.
-We can then construct the invariant measure for $K<N$ from the invariant measure for $O(N)$:
-
-\[
-  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j = \bigwedge_{j=1}^K \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j,
-\]
-
-Though we needed to choose a $X^\perp$, the invariant measure does not depend on the choice of $X^\perp$.
-Moreover, this measure is invariant to left-actions by $O(N)$ and right-actions by $O(K)$.
-For more details, see \cite{muirhead2009aspects}.
-
-When $K=1$, the invariant measure is equivalent to the usual Hausdorff measure on the sphere.
-
-\subsection{Givens parameterization of Stiefel manifold}
-
-\subsection{Cayley parameterization of Stiefel manifold}
-
-\subsection{Polar parameterization of Stiefel manifold}
-
-\subsection{Householder parameterization of Stiefel manifold}
-
-\subsection{Matrix exponential parameterization of Special Orthogonal group}
-
-\section{Unit sphere}
-
-The unit sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$ consists of all unit vectors $x$ of length 1, i.e. $x^\top x = 1$.
-It is a special case of the Stiefel manifold.
-
-Densities on the sphere are generally defined with respect to the spherically uniform measure.
-This measure can be defined in many ways that are equivalent to within a constant factor that is not relevant for MCMC.
-We will take the same approach as with the Stiefel manifold and define the spherically uniform measure as the measure that is invariant to transformations by $N \times N$ orthogonal matrices (rotations and reflections).
-As discussed in the previous section, if $\tilde{X} \in O(N)$ is an orthogonal matrix whose first column is $x$, then $\tilde{X}^\top x = e_1$, where $e_1$ is the unit vector whose first element is 1, and $(\tilde{X}^\top \dv{x})_1 = 0$.
-For convenience, we will assume that the rules used to produce $\tilde{X}$ enforce that $\det(\tilde{X}) = +1$.
-The spherically uniform measure is then written
-\[
-    (\dv{x})^\wedge \equiv (\tilde{X}^\top \dv{x})^\wedge = \bigwedge_{i=2}^N (\tilde{X}^\top \dv{x})_i.
-\]
-
-\subsection{Embedding in real space using parameter expansion}
-
-Let $f\colon y \mapsto (r, x)$ for $y \in \mathbb{R}^N$, $r > 0$, and $x \in \mathbb{S}^{N-1}$.
-This map is bijective everywhere except $y=0$.
-
-To compute the Jacobian determinant of the map, we first compute the differential form of $y=rx$:
-
-\[\dv{y} = \dv{r} x + r \dv{x}.\]
-
-Multiplying both sides by $\tilde{X}^\top$, we find
-
-\[
-    \tilde{X}^\top \dv{y} = \dv{r} \tilde{X}^\top x + r \tilde{X}^\top \dv{x} = \dv{r} e_1 + r (\tilde{X}^\top \dv{x}).
-    \]
-
-Then
-
-\[
-    (\tilde{X}^\top \dv{y})_i = \begin{cases}
-        \dv{r} & \text{ if } i = 1 \\
-        r (\tilde{X}^\top \dv{x})_i & \text{ if } i \ne 1
-    \end{cases}.
-\]
-
-Note that $(\tilde{X}^\top \dv{y})^\wedge = \det({\tilde{X}}) (\dv{y})^\wedge = (\dv{y})^\wedge$.
-
-Taking the exterior product of both sides, we then find
-
-\[
-\begin{aligned}
-    (\dv{y})^\wedge &= \dv{r} \wedge \left(\bigwedge_{i=2}^N r (\tilde{X}^\top \dv{x})_i \right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge\right)\\
-    \dv{r} \wedge (\dv{x})^\wedge &= r^{1-N} (\dv{y})^\wedge = \lVert y \rVert^{1-N} (\dv{y})^\wedge
-\end{aligned}
-\]
-
-So the absolute Jacobian determinant is $\absdet{J_f(y)} = \lVert y \rVert^{1-N}$.
-
-In the constrained space there is a free radial parameter $r$ that is unconstrained on the positive reals.
-As a result, nothing prevents $r$ from growing arbitrarily large, which would cause the posterior to be improper, i.e. to have infinite mass.
-Thus, we must also specify a prior $\pi(r)$.
-While we may want to customize $\pi(r)$, a reasonable default choice is $r \sim \chi_N$, which has the density $\pi(r) \propto r^{N-1} \exp\left(-\frac{1}{2}r^2\right) $.
-
-The combined log density correction is then
-\[
-  \absdet{J_f(y)} \pi(r) = \absdet{J_f(y)} \pi(\lVert y \rVert) \propto \lVert y \rVert^{1-N} \lVert y \rVert^{N-1} \exp\left(-\frac{1}{2}\lVert y \rVert^2\right) = \exp\left(-\frac{1}{2}\lVert y \rVert^2\right).
-\]
-
-That is, if $x$ is uniformly distributed on the sphere and $r$ follows this prior, then $y$ has a zero-centered standard multivariate normal distribution.
-
-Since HMC is especially efficient at sampling normal distributions, this parameterization is very efficient when the marginal posterior distribution of $x$ is very diffuse on the sphere.
-However, as the target distribution becomes concentrated on $x$, the geometry near the singularity $y=0$ may become problematic.
-
-\subsection{Parameterization using hyperspherical coordinates}
-
-Given $z \in \mathbb{R}^N$, $r > 0$ and angles $\phi=\{\phi_1, \ldots, \phi_{N-1}\}$, where $\phi_{N-1} \in (0, 2\pi]$ and $\phi_i \in (0, \pi]$ for $i < N-1$, we define the relationship between $z$ and $(r, \phi)$ as
-
-\[
-  z_i = \left(r \prod_{j=1}^{i-1} \sin(\phi_j)\right) \begin{cases}
-    1 & \text{ if } i = N\\
-    \cos(\phi_i) & \text{otherwise}
-  \end{cases}.
-\]
-
-For the plane $N=2$ this works out the usual polar coordinate system:
-\[
-    \begin{aligned}
-        z_1 &= r \cos\phi_1\\
-        z_2 &= r \sin\phi_1
-    \end{aligned}
-\]
-
-For $N=3$ this is \footnote{
-    Note that this definition of hyperspherical coordinates differs for $N=3$ from the usual definition of spherical coordinates.
-    It can be made the same through the coordinate permutation $(z_1, z_2, z_3) \to (z_2, z_3, z_1)$.
-}
-\[
-    \begin{aligned}
-        z_1 &= r \cos\phi_1\\
-        z_2 &= r \sin\phi_1\cos\phi_2\\
-        z_3 &= r \sin\phi_1\sin\phi_2
-    \end{aligned}
-\]
-
-Our strategy is to compute the Jacobian determinant of the bijective map $g: (r, \phi) \mapsto z$ and relate it to the Jacobian determinant from the previous section to find the Jacobian determinant of the bijective map $f_2: \phi \mapsto x$ for $x \in \mathbb{S}^{N-1}$.
-
-To do so, note first that
-\[
-  z_i^2 = \left(r^2 \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \begin{cases}
-    1 & \text{ if } i = N\\
-    \cos^2(\phi_i) & \text{otherwise}
-  \end{cases}.
-\]
-
-Note also that
-\[
-    \begin{aligned}
-        z_N^2 &= r^2 \prod_{j=1}^{N-1} \sin^2(\phi_j)\\
-        z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \cos^2(\phi_{N-1}) + \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j) \right) \sin^2(\phi_{N-1}) = r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\\
-        z_{N-2}^2 + z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \cos^2(\phi_{N-2}) + \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \sin^2(\phi_{N-2}) = r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\\
-        &\vdots \\
-        \sum_{k=1}^i z_{N - i + k}^2 &= r^2 \prod_{j=1}^{N - i} \sin^2(\phi_j)\\
-        &\vdots \\
-        \sum_{k=1}^N z_{N - i + k}^2 &= r^2
-    \end{aligned}
-\]
-
-We then differentiate both sides of this expression to get the differential form:
-
-\[
-    \begin{aligned}
-        2 z_N \dv{z_N} &= 2r\dv{r} + 2 r^2 \left(\prod_{j=1}^{N-1} \sin^2(\phi_j)\right)\left( \sum_{j=1}^{N-1} \cot(\phi_j) \dv{\phi_j} \right)\\
-        2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-2} \cot(\phi_j)\right)\\
-        2z_{N-2}\dv{z_{N-2}} + 2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-3} \cot(\phi_j)\right)\\
-        &\vdots \\
-        2\sum_{k=1}^i z_{N - i + k} \dv{z_{N - i + k}} &= 2 r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N - i} \sin^2(\phi_j)\right) \left(\sum_{k=1}^{N-i} \cot(\phi_k) \dv{\phi_k}\right)\\
-        &\vdots \\
-        2\sum_{k=1}^N z_{N - i + k} \dv{z_{N - i + k}} &= 2r\dv{r}
-    \end{aligned}
-\]
-
-For each side separately, we now take the exterior product for all $i \in 1\ldots N$.
-On the left side, we start at the top with $(2z_N dz_n) \wedge (2 z_{N-1} \dv{z_{N-1}} + 2 z_N \dv{z_N}) = 2^2 z_{N} z_{N-1} (\dv{z_N} \wedge \dv{z_{N-1}})$.
-Note that the $\dv{z_N}$ from the second equation is eliminated since the exterior product already contains that term.
-This trend continues as we proceed to the bottom, and the result is
-\[
-    \bigwedge_{i=1}^N 2z_{N-i+1} \dv{z_{N-i+1}} = \bigwedge_{i=1}^N 2z_i \dv{z_i} = 2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge
-\]
-
-Note that
-
-\[
-  \prod_{i=1}^N z_i = r \prod_{i=1}^{N-1} r \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j) = r^N \prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j)
-\]
-
-On the right side, we start at the bottom instead with
-\[
-(2r\dv{r} + 2 r^2 \sin(\phi_1) \cos(\phi_1) \dv{\phi_1}) \wedge (2r\dv{r}) = 4 r^3 \sin(\phi_1) \cos(\phi_1) (\dv{\phi_1} \wedge \dv{r}).
-\]
-Again, the $\dv{r}$ term in the second-from-bottom equation cancels since $\dv{r}$ is already included in the bottom term.
-This trend also continues as we proceed to the top, and the result is then
-
-\[
-\begin{aligned}
-    \left(\bigwedge_{i=1}^{N - 1} 2 r^2 \left(\prod_{j=1}^{N - i - 1} \sin^2(\phi_j)\right) \sin(\phi_{N-i}) \cos(\phi_{N-i}) \dv{\phi_{N-i}} \right) \wedge (2r\dv{r}) \\
-    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_{N-i}) \cos(\phi_{N-i}) \prod_{j=1}^{N-i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
-    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
-    = 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
-\end{aligned}
-\]
-
-Equating the exterior products of each side, we have
-
-\[
-\begin{aligned}
-    2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge &= 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
-    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
-\end{aligned}
-\]
-
-As a result, the Jacobian determinant is
-
-\[
-    \absdet{J_g(r, \phi)} = r^{N-1} \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}
-\]
-
-As a quick check, for the circle ($N=1$), $\absdet{J_g(r, \phi)} = r$, and for the sphere ($N=2$), $\absdet{J_g(r, \phi)} = r^2 \sin(\phi_1)$, which are the expected results.
-A similar result was given in \cite{muirhead2009aspects}.
-
-Comparing with the previous section, we find
-
-\[
-\begin{aligned}
-    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge \right) \\
-    &= (-1)^{N-1} r^{N-1} \left((\dv{x})^\wedge \wedge \dv{r}\right)\\
-    (\dv{x})^\wedge &= (-1)^{N-1}\left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) (\dv{\phi})^\wedge
-\end{aligned},
-\]
-
-so the absolute Jacobian determinant for the bijective map from hyperspherical coordinates to the sphere is
-\[
-    \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}.
-\]
-
-However, we're not done yet.
-Recall that $\phi_j \in (0, \pi]$ for $i < N-1$ and $\phi_{N-1} \in (0, 2\pi]$.
-So far we've essentially changed the problem of sampling on the hypersphere in $N$ dimensions to the problem of sampling on $N-1$ intervals, or, alternatively, $N-2$ hemicircles and 1 circle.
-
-There are a few considerations when choosing the remaining parameterizations.
-First, and most importantly, we want a smooth geometry.
-Second, we want to avoid unnecessary work.
-For example, if we choose parameter expansion for these parameters, we will end up with $N-1$ extra degrees of freedom.
-Third, we want to avoid costly functions if possible.
-For example, computing $\sin$ and $\cos$ is about 3 times more costly than computing $\exp$ and about 20 times more costly than computing a square.
-
-\section{Cholesky factors of correlation matrices}
-
-A $K \times K$ correlation matrix $\Omega$ is positive definite, has a unit diagonal, and all off-diagonal entries are between -1 and 1 (exclusive).
-Because it is positive definite, it can be Cholesky factored (see \ref{cholesky}) to a $K \times K$ lower-triangular matrix $L$
-with positive diagonal elements such that $\Omega = L \cdot L^T$. Because the correlation matrix has a unit diagonal,
-
-\[
-  \Omega_{k,k} = L_k \cdot L_k^T = 1
-\]
-
-each row vector $L_k$ of the Cholesky factor is of unit length. The length and positivity
-constraint allow the diagonal elements of L to be calculated from the off-diagonal
-elements, so that a Cholesky factor for a $K \times K$ correlation matrix requires only $\binom{K}{2}$ unconstrained parameters.
-
-\subsection{Hyperbolic tangent transform}
-
-Suppose $y$ is a vector of $\binom{K}{2}$ unconstrained values. Let $z$ be a lower-triangular matrix with zero diagonal and below diagonal entries filled by row. For example, in the $3 \times 3$ case,
-\[
-z = \begin{bmatrix}
-0 & 0 & 0\\
-\tanh{y_1} & 0 & 0\\
-\tanh{y_2} & \tanh{y_3} & 0\\
-\end{bmatrix}
-\]
-
-The matrix $z$, with entries in the range $( -1, 1)$, is then transformed to the Cholesky factor $x$, by taking
-\[
-x_{i, j}=\left\{\begin{array}{lll}
-0 & \text { if } i<j & \text { [above diagonal] } \\
-\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i=j & \text { [on diagonal] } \\
-z_{i, j} \sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i>j & \text { [below diagonal] }
-\end{array}\right.
-\]
-
-In the $3 \times 3$ case, this yields
-
-\[
-x=\left[\begin{array}{ccc}
-1 & 0 & 0 \\
-z_{2,1} & \sqrt{1-x_{2,1}^{2}} & 0 \\
-z_{3,1} & z_{3,2} \sqrt{1-x_{3,1}^{2}} & \sqrt{1-\left(x_{3,1}^{2}+x_{3,2}^{2}\right)}
-\end{array}\right]
-\]
-
-where the $z_{i,j}\in (-1, 1)$ are the tanh-transformed $y$.
-
-\subsection{Absolute Jacobian determinant of inverse transform}
-
-The Jacobian of the full transform is the product of the Jacobians of its component
-transforms. First, for the inverse transform $z = \tanh{y}$, the derivative is
-\[
-\frac{d}{dy} \tanh y = \frac{1}{(\cosh y)^{2}}
-\]
-Second, for the inverse transform of $z$ to $x$, the resulting Jacobian matrix $J$ is of
-dimension $\binom{K}{2} \times \binom{K}{2}$, with indexes $(i, j)$ for $(i > j)$. The Jacobian matrix is lower
-triangular, so that its determinant is the product of its diagonal entries, of which
-there is one for each $(i, j)$ pair.
-\[
-|\operatorname{det} J|=\prod_{i>j}\left|\frac{d}{d z_{i, j}} x_{i, j}\right|
-\]
-
-where
-
-\[
-\frac{d}{d z_{i, j}} x_{i, j}=\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}}
-\]
-
-So the combined log Jacobian determinant
-of the complete inverse transform is given by
-
-\[
-\log |\operatorname{det} J|=-2 \sum_{n \leq(\substack{K \\ 2}} \log \cosh y+\frac{1}{2} \sum_{i>j} \log \left(1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}\right)
-\]
-
-\section{Positive definite matrices}
-
-[SA: all transformations below can be generalized to SPD matrices of fixed rank, but we'd need to know what the common reference measure is for densities of SPD matrices, and I can't recall seeing one.]
-
-An $N \times N$ positive definite (PD) matrix $X$ is a symmetric matrix that satisfies $v^\top X v > 0$ for all non-zero vectors $v$.
-Covariance matrices are examples of PD matrices.
-
-Density functions for distributions of PD matrices, such as the Wishart distribution, are generally defined with respect to the Lebesgue measure of the unique triangular elements $(\dv{X})^\wedge$.
-
-\subsection{Cholesky parameterization}\label{cholesky}
-
-Given a positive (semi-)definite matrix $X$ of rank $K$, we can uniquely decompose it as $X = L L^\top$ where $L$ is an $N \times N$ lower triangular matrix with $L_{jj} > 0$ for all $j \le K$ and $L_{ij} = 0$ for all $j > K$.
-$L$ has $\frac{K(K+1)}{2} + (N - K)K$ degrees of freedom.
-When $K=N$, $X$ is positive-definite.
-
-We can construct $X$ by composing 2 bijective maps.
-The first, $f_1$, enforces the constraint of the positive diagonal:
-
-\[L_{ij} = f_1(Y)_{ij} = \begin{cases} \exp(Y_{ij}) & \text{ if } i = j \le K \\ Y_{ij} & \text{otherwise} \end{cases}\]
-
-The Jacobian determinant for this map is $\absdet{J_{f_1}(Y)} = \exp(\sum_{i=1}^K Y_{ii})$.
-
-The second map is $f_2: L \mapsto L L^\top = X$.
-To compute the Jacobian of the map, we use the half-vectorization $\vectu(X)$.
-Note that $X_{ij} = \sum_{k=1}^{\min(i, j)} L_{ik} L_{jk}$.
-$X_{ij}$ is thus only dependent on elements of $L$ that are below and to the left of the $(i, j)$ position.
-As a result, the Jacobian of $\vectu \circ f_2 \circ \vectu^{-1}$ is triangular, and its determinant is only the product of its diagonal elements: $\absdet{J_{f_2}} = \prod_{i=1}^N \prod_{j=1}^i \frac{\partial X_{ij}}{\partial L_{ij}}$.
-
-In this index range, $\frac{\partial X_{ij}}{\partial L_{ij}} = (1 + \delta_{ij}) L_{ii}$.
-As a result
-
-\[
-  \absdet{J_{f_2}(Z)} = 2^N \prod_{i = 1}^N \prod_{j = 1}^i L_{ii} = 2^N \prod_{i=1}^N |L_{ii}|^{N - i + 1} \\
-\]
-
-Composing the two bijective maps as $f = f_2 \circ f_1$, the resulting Jacobian determinant is
-
-\[
-  \absdet{J_{f}(Y)} = 2^N \exp\left( \sum_{i=1}^N (N - i + 2) Y_{ii}\right).
-\]
-
-\subsection{Matrix exponential parameterization}
-
-For every PD matrix $X \in \mathbb{R}^{N \times N}$, there is a unique symmetric matrix $Y$ such that $X = \Exp(Y)$, where $\operatorname{Exp}(\cdot)$ is the matrix exponential, defined as
-\[\Exp(Y) = \sum_{k=0}^\infty \frac{1}{k!} Y^k.\]
-
-Let $Y = U \Lambda U^\top$ be the symmetric eigendecomposition of $Y$, where the matrix of eigenvectors $U$ is orthogonal, and the $\Lambda = \operatorname{diag}(\lambda)$ for $\lambda \in \mathbb{R}^N$.
-Then $Y^2 = U \Lambda U^\top U \Lambda U^\top = U \Lambda^2 U^\top$.
-In fact, $Y^c = U \Lambda^c U^\top$ for all real $c$.
-As a result,
-\[\Exp(Y) = U \Exp(\Lambda) U^\top,\]
-where $\Exp(\Lambda)_{ij} = \exp(\lambda_i) \delta_{ij}$.
-
-$\Exp(Y)$ is one example of a matrix function $f$, which we can roughly define for symmetric matrices as any function that obeys
-\[f(Y) = U f(\Lambda) U^\top,\]
-where $f(\Lambda)_{ij} = f(\lambda_i)\delta_{ij}$.
-\footnote{
-  In a similar way, we can generalize other scalar functions like the inverse, trigonometric, and hyperbolic functions to functions of matrices.
-  All results in this section apply to these functions as well.
-}
-
-For such functions $f$, we can compute a common form of the Jacobian determinant.
-Differentiating the expression, we find
-\[
-\begin{aligned}
-  \dv{(f(Y))} &= \dv{U} f(\Lambda) U^\top + U f'(\Lambda) \dv{\Lambda} U^\top + U f(\Lambda) \dv{U}^\top\\
-  U^\top \dv{(f(Y))} U &= U^\top \dv{U} f(\Lambda) + f(\Lambda) \dv{U}^\top U + f'(\Lambda)\dv{\Lambda},
-\end{aligned}
-\]
-where $f'(\Lambda)_{ij} = \frac{\dv{(f(\lambda_i))}}{\dv{\lambda_i}}$.
-
-Recall that $U^\top \dv{U}$ is skew-symmetric.
-Note also that for skew-symmetrix $Z = -Z^\top$ and diagonal matrix $\operatorname{diag}(v)$,
-\[(Z\operatorname{diag}(v) + \operatorname{diag}(v)Z^\top)_{ij} = Z_{ij} v_j + v_i Z_{ji} = Z_{ij} v_j - v_i Z_{ij} = Z_{ij} (v_j - v_i) = Z \circ V,\]
-where $V_{ij} = v_j - v_i$ and $\circ$ is the Hadamard (elementwise) product.
-
-From the right-hand side, note then that
-\[(U^\top \dv{(f(Y))} U)_{ij} = \begin{cases} (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) & \text{ if } i \ne j \\ f'(\lambda_i) \dv{f(\lambda_i)} &\text{ if } i = j \end{cases}\]
-
-The exterior product of the diagonal part is
-\[
-  \bigwedge_{i=1}^N f'(\lambda_i) \dv{\lambda_i} = \left(\prod_{i=1}^N f'(\lambda_i)\right) (\dv{\Lambda})^\wedge,
-\]
-while the exterior product of the sub-diagonal part is
-\[
-  \bigwedge_{i=1}^N \bigwedge_{j=i+1}^N (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) = \left( \prod_{i=1}^N \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i)) \right) \left(U^\top \dv{U} \right)^\wedge
-\]
-
-Lastly, note for any symmetric $Z$ and orthogonal matrix $U$,
-\[(U^\top \dv{Z} U)^\wedge = \det(U^\top)^N \det(U)^N (\dv{Z})^\wedge = (\dv{Z})^\wedge.\]
-
-As a result, we can compute $(\dv(f(Y)))^\wedge$ by taking the exterior product of the sub-diagonal and diagonal parts: 
-\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i))\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]
-
-If $f$ is the identity map $f(Y) = Y$, then
-\[ (\dv{Y})^\wedge = \left(\prod_{i=1}^N \prod_{j=1}^{i-1} (\lambda_j - \lambda_i)\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]
-
-Combining the latter two expressions:
-\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} \frac{f(\lambda_j) - f(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge\]
-
-[REF, Edelman's notes] previously found this result.
-
-For $X = f(Y) = \Exp(Y)$ then,
-\[ (\dv{X})^\wedge = \left(\prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge,\]
-
-so the Jacobian determinant of the exponential map is $\absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right|$.
-
-Note that this expression is undefined when any eigenvalues are repeated.
-This is not a problem for MCMC, since the eigenvalues are non-unique with probability 0, but non-unique eigenvalues could occur when maximizing a log-density.
-
-Because $\lim_{h \to 0} \frac{f(x) - f(x + h)}{h} = f'(x)$, we can then modify the expression to handle such cases:
-\[
-  \absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \prod_{j=1}^i \begin{cases}
-    \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right| &\text{ if } \lambda_i \ne \lambda_j \\ 
-    \exp(\lambda_i) &\text{ if } \lambda_i = \lambda_j
-  \end{cases}.
-\]
-
-If the eigenvalues are sorted in ascending order, then the log Jacobian determinant can be more stably computed as
-
-\[
-  \log\absdet{J_{\Exp}(Y)} = \sum_{j=1}^N j\lambda_j + \sum_{i=1}^j \begin{cases}
-    \log(1 - \exp(\lambda_i - \lambda_j)) - \log(\lambda_j - \lambda_i) &\text{ if } \lambda_i \ne \lambda_j \\ 
-    0 &\text{ if } \lambda_i = \lambda_j
-  \end{cases}.
-\]
-
-While the cost of computing the Jacobian determinant is $O(N^2)$, the cost of computing the eigendecomposition naively is $O(N^3)$, so this transformation is expensive.
-An attractive alternative would be to define the transform directly from the eigendecomposition of $X$.
-
-\subsection{Eigendecomposition parameterization}
-
-To define a transform directly from the symmetric eigendecomposition of $X$, we first need to note that the eigendecomposition is not unique; if we swap any two columns $i$ and $j$ of $U$ as well as swapping the eigenvalues $\lambda_i$ and $\lambda_j$, the result is the same matrix.
-Moreover, if we reverse the sign of any given eigenvector, it has no effect on the eigenvalues or $X$.
-
-When the eigenvalues are all unique, we can make the eigendecomposition also unique by enforcing an ascending order of eigenvalues $0 < \lambda_i < \lambda_j$ for $j > i$ and by forcing the first entry of each eigenvector to be positive.
-
-So we can construct a bijective map with the maps $f_1$ that produces an ordered vector of positive eigenvalues and $f_2$ that produces an orthogonal matrix whose first row contains only positive entries.
-Consider $\lambda = f_1(y_1)$ and $U = f_2(y_2)$.
-
-Let $f_1$ be the map that computes each $\lambda_i$ as
-\[
-  \lambda_i = \begin{cases}
-   \exp\bigl((y_1)_1\bigr) &\text{ if } i = 1\\
-    \lambda_{i-1} + \exp\bigl((y_1)_i\bigr) &\text{ if } 1 < i \le N
-  \end{cases}
-\]
-The Jacobian of this map is triangular, so its determinant is $\absdet{J_{f_1}(y_1)} = \exp\left(\sum_{i=1}^N (y_1)_i\right)$.
-
-We can then choose for $f_2$ any of the transforms to the orthogonal matrices previously discussed, though care must be taken to avoid multimodality.
-For example, if the transform covers the entire orthogonal group, then $2^N$ matrices map to the same orthogonal matrix with positive first row, which will manifest as $2^N$ modes in the unconstrained distribution.
-If, however, we can modify one of the transforms to natively produce only matrices with this constraint, then we can eliminate these undesirable modes.
-
-The combined Jacobian determinant of $f: (y_1, y_2) \mapsto f_2(y_2) \operatorname{diag}(f_1(y_1)) f_2(y_2)^\top = X$ is then 
-\[ \absdet{J_{f}(y_1, y_2)} = \absdet{J_{f_2}(y_2)} \exp\left(\sum_{i=1}^N (y_1)_i\right).\]
-
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1.2\textwidth]{figures/simplex/rmse.png}
+    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps.
+At every iteration in a run, cumulative mean of samples is used to calculate RMSE, which is plotted against the cumulative leapfrog steps till that iteration}
+    \label{fig:rmse}
+\end{figure}
+\clearpage
 \subsubsection*{Acknowledgements}
@@
 
+\include{appendix}
 \end{document}

### 2022-07-29 paper.tex
sync

@@
 \usepackage{graphicx}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{hyperref}
+    \hypersetup{
+        colorlinks=true,
+        linkcolor=MidnightBlue,
+        anchorcolor=MidnightBlue,
+        citecolor=MidnightBlue,
+        filecolor=MidnightBlue,
+        urlcolor=MidnightBlue,
+        pdftitle={Efficient Unconstraining Transforms for HMC},
+        pdfauthor={Meenal Jhajaria et al.},
+        pdfpagemode=FullScreen,
+    }
+\usepackage[title]{appendix}
 
@@
 Markov chain Monte Carlo (MCMC) methods.  The most effective MCMC
-method in high dimensions is Hamiltonian Monte Carlo (HMC).  HMC works
-by simulating the Hamiltonian dynamics of a fictitious particle
+method in high dimensions is Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc}.
+HMC works by simulating the Hamiltonian dynamics  a fictitious particle
 representing the value being sampled coupled with a momentum term.
@@
 the constrained values to unconstrained values before sampling
-\cite{JSSv076i01}.
-For example, the Probabilistic  Programming language Stan, defines distributions with an unconstrained support, in favour of making sampling an easier process.
-The variables with constraints (or with constrained support in this case) are transformed to an unconstrained space.
-Inverse transforms of these variables are used for log density adjustments.
-These are usually in the form of a Jacobian matrix comprised of the gradient $\frac{\partial x}{\partial y}$, where $x$ and $y$ are the constrained and unconstrained parameters respectively.
+\cite{JSSv076i01, radul2021base, fjelde2020bijectors}.
 
-These gradients are computationally expensive, certain transforms are entirely inefficient, and some work well on certain distributions or parametrizations and vice versa.
-This presents the issue of selecting which transform to use
-among an infinite set of options, which is the topic of this paper.
-We begin by defining the general theory of transforms, followed by transforms from a unit simplex $\Delta^N$ to the $R^M$.
-Finally, we discuss notions of "better" transforms defined on the simplex, using statistical measures like Effective Sample Size, Leapfrog steps taken by the sampler etc.
+Our goal in this paper is to evaluate a standard and less well known transforms for constrained data types in regular use in statistical modeling.
+We will evaluate transforms in the head, body, and tail of distributions in terms of computational efficiency of both the transform, its Jacobian determinant and associated gradients, the geometry induced in the unconstrained space (e.g., log convexity and conditioning), as well as Hamiltonian Monte Carlo sampling efficiency.
 
-\section{Unconstraining Transforms}
+We consider transforms for several constrained data types, including scalars with lower- and/or upper-bounds, vectors with simplex, sum-to-zero or unit length constraints, and matrices (or their Cholesky factors) constrained to be positive definite (e.g., covariance matrices), positive definite with unit diagonal (e.g., correlation matrices), or orthonormal.
 
-To draw samples from a parameter defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y} \in \mathbb{R}^m$ for $m \ge n$.
-Let $p_Y(y)$ be an improper density function defined over $\mathcal{Y} \rightarrow(0, \infty)$ with support over all elements of $\mathcal{Y}$ and a surjective, smooth and continuous map $g: \mathcal{Y} \rightarrow \mathcal{X}$.
-Then, for any proper density $\pi_{X}$ over $\mathcal{X}$, the constraining transform can be defined as $\left(p_{Y}, g\right)$: $\mathcal{Y} \rightarrow \mathcal{X}$.
-In this case, the density $p_{Y}(y)=\pi_{Y}(y) \cdot \pi_{X}(g(y))$ is proper, and $Y \sim \pi_Y \implies g(Y) \sim \pi_X$.
-The required change of variables for this unconstraining transform is:
+\section{Constraining transforms}
+
+In this paper, we restrict our attention to a general class of surjective constraining transforms $f : \mathcal{Y} \rightarrow \mathcal{X}$ from an unconstrained space $\mathcal{Y}$ onto a constrained space $\mathcal{X}$.
+For all of our transforms, we have $\mathcal{Y} = \mathbb{R}^M$ and $\mathcal{X} \subset \mathbb{R}^N$.
+Along with each transform, we have an additional (perhaps improper) density function $h(y)$ defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(x) \sim \textrm{uniform}(\mathcal{X}).$
+
+The above definition is informal in that it's not possible to sample over an improper density.
+Instead, we can characterize our end-to-end goal, which is that if we have a proper density $p_X(x)$ defined for $x \in \mathcal{X}$, then we want the induced distribution on $y$, namely
+$$
+p_Y(y) = h(y) \, p_X(f(y)),
+$$
+to be proper and be such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
+
+\subsection{Bijective constraining transforms}
+
+The traditional way to define constraining transforms is through the usual change-of-variables formula.
+Suppose we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$ and we have a smooth and bijective unconstraining transform $g : \mathcal{X} \rightarrow \mathcal{Y}$.
+We then define our constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$ as the inverse of $g$, $f = g^{-1}$.
+The appropriate density on $\mathcal{Y}$ is then defined as the absolute value of the determinant of the constraining transform
 \[
-  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
+h(y) = \absdet{J_f(y)}.
+\]
+That way, for any given $p_X(x)$, we have
+\[
+p_Y(y) = p_X(f(y)) \, \absdet{J_f(y)},
 \]
@@
 \[
-  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
+  J_{f^{-1}}(y) \ = \ \nabla_y \, f^{-1}(y) \ = \ \frac{\partial}{\partial y} \, f^{-1}(y).
 \]
+Because we have applied a standard change of variables adjustment, we are guaranteed that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
 
-%notion of uniformity missing
-In this paper we consider transforms where $m = n$, for which the constraining transform is defined as $(\pi_Y, f)$: $\mathcal{Y} \rightarrow \mathcal{X}$, given that $\pi_{Y}(y)=\absdet{J_{f^{-1}}(y)}$.
+\subsection{Surjective constraining transforms}
 
-\section{Unit simplex}
+Not all of our transforms are simple bijective changes of variables as described in the previous subsection.
+In some cases, we will take the dimensionality of the unconstrained space $\mathcal{Y}$ to be higher than that of the constrained space $\mathcal{X}$ and thus need additional constraints to ensure the correct distributions on $\mathcal{X}$.
+In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transform coupled with a standard normal distribution over $\mathcal{Y}$,
+\[
+  h(y) = \textrm{normal}(y \mid 0, \textrm{I}) \ \absdet{J_f(y)}.
+\]  
+In each case, we will have to ensure that $p_Y(y) = h(y) \, p_X(f(y))$ is such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
+
+\section{Vector transforms}
+
+In this section, we consider transforms from unconstrained space to constrained vectors, including unit-simplex, sum-to-zero, and unit-length constraints.
+
+\subsection{Unit simplex}
 
@@
 \]
-%n not n-1
 We will use $\Delta^N_-$ to denote $N$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
@@
 
-The Stick-Breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
+The stick-breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
 It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
@@
  z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
-\]\\
+\]
 
@@
                                             \right)\right)
-                                            \]\\
+\]
 
@@
 \]
+
 \subsection{Additive log ratio transform}
@@
 \subsection{Augmented-Softmax Transform}
-% Needs to be written like other transforms -> transform, inverse, |J|
+%TODO:  Needs to be written like other transforms -> transform, inverse, |J|
 We define the transformation
@@
 
+%TODO: better captions
 \section{Results}
-\begin{figure}[h!]
+
+\begin{table}[!ht]
+    \centering
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+        ~ & stan & softmax-augmented & softmax & stickbreaking \\ \hline
+        $\alpha = 0.1, \, N = 10$ & 6.908s & 11.712s & 10.548s & 13.386s \\ 
+        $\alpha = 0.1, \, N = 100$ & 74.274s & 33.757s & 61.826s & 159.016s \\ 
+        $\alpha = 0.1, \, N = 1000$ & 1108.48s & 307.463s & 657.791s & 11773.4s \\ 
+        $\alpha = 1, \, N = 10$ & 4.072s & 6.328s & 5.337s & 5.67s \\ 
+        $\alpha = 1, \, N = 100$ & 20.805s & 27.151s & 25.94s & 51.45s \\ 
+        $\alpha = 1, \, N = 1000$ & 306.942s & 171.824s & 184.126 & 3023.24 \\ 
+        $\alpha = 10, \, N = 10$ & 4.123s & 6.375s & 5.563s & 5.606s \\ 
+        $\alpha = 10, \, N = 100$ & 23.6s & 29.091s & 30.747s & 48.834s \\ 
+        $\alpha = 10, \, N = 1000$ & 205s.627 & 161.11s & 150.751s & 1866.83s \\ \hline
+    \end{tabular}
+\end{table}
+
+\begin{figure}[t!]
     \centering
@@
 \end{figure}
-\newpage
-\begin{figure}[h!]
+
+\begin{figure}[t!]
     \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[t!]
     \centering
@@
 \end{figure}
-\clearpage
+
 \subsubsection*{Acknowledgements}

### 2022-08-28 paper.tex
add appendix
sync

@@
 For all of our transforms, we have $\mathcal{Y} = \mathbb{R}^M$ and $\mathcal{X} \subset \mathbb{R}^N$.
-Along with each transform, we have an additional (perhaps improper) density function $h(y)$ defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(x) \sim \textrm{uniform}(\mathcal{X}).$
+Along with each transform, we have an additional (perhaps improper) density function $h(y)$ defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\mathcal{X}).$
 
@@
 \]
-
 \subsection{Additive log ratio transform}

### 2022-10-30 appendix.tex
Updates from Overleaf

@@
 \]
+
+\subsection{Hyperspherical simplex transforms}
+
+\subsubsection{Hyperspherical coordinate transform}
+
+We use the following definition of hyperspherical coordinates.
+For $\phi_i \in (0, \pi/2]$ and $u \in \mathbb{S}^N_{>0}$, the positive orthant of the $N$-sphere,
+\[
+  u_i = \left(\prod_{k=1}^{i-1} \sin(\phi_k)\right) \begin{cases}
+    1 & \text{ if } i = N + 1\\
+    \cos(\phi_i) & \text{otherwise}
+  \end{cases}.
+\]
+
+Let $z_i = \sin^2(\phi_i)$ and $x_i = u_i^2$, so that $x \in \Delta^N$.
+Then
+\[
+  x_i = \left(\prod_{k=1}^{i-1} z_k\right) \left(\begin{cases}
+    1 & \text{ if } i = N + 1\\
+    (1 - z_i) & \text{otherwise}
+  \end{cases}\right) = (1 - z_i (1 - \delta_{i,N+1})) \prod_{k=1}^{i-1} z_k.
+\]
+
+\subsubsection{Hyperspherical coordinates inverse transform}
+
+To derive the inverse transform, note that
+\[
+    \begin{aligned}
+        x_{N+1} &= \prod_{k=1}^{N} z_k\\
+        x_N + x_{N+1} &= (1 - z_N)\left(\prod_{k=1}^{N-1} z_k\right) + z_N \left( \prod_{k=1}^{N-1} z_k \right) = \prod_{k=1}^{N-1} z_k = \frac{x_N}{1 - z_N}\\
+        x_{N-1} + x_{N} + x_{N+1} &= (1-z_{N-1}) \left(\prod_{k=1}^{N-2} z_k\right) + z_{N-1} \left(\prod_{k=1}^{N-2} z_k\right) = \prod_{k=1}^{N-2} z_k = \frac{x_{N-1}}{1 - z_{N-1}}\\
+        &\vdots \\
+        \sum_{k=i}^{N+1} x_{k} &= \prod_{k=1}^{i - 1} z_k = \frac{x_i}{1 - z_i}.
+    \end{aligned}
+\]
+
+From the final expression, we can define the inverse transform:
+\[
+z_i = 1 - \frac{x_i}{\sum_{k=i}^{N+1} x_k}
+\]
+
+\subsubsection{Jacobian of hyperspherical coordinates transform}
+
+To derive the Jacobian, we note that $x_i$ depends only on $z_k$ for $k \le i$.
+As a result, the Jacobian of the map from $z$ to $x_-$ is lower triangular, and its determinant depends only on its diagonal, which consists of the terms
+$$J_{ii} = -\frac{\partial x_i}{\partial z_i} = \prod_{k=1}^{i-1} z_k,$$
+so
+$$
+|J| = \prod_{i=1}^N |J_{ii}|
+= \prod_{i=1}^N \prod_{k=1}^{i-1} z_k
+= \prod_{k=1}^{N-1} \prod_{i=k+1}^N z_k
+= \prod_{i=1}^{N-1} z_i^{N-i} = \prod_{i=1}^N z_i^{N-i}
+$$
+
+Due to the form of the Jacobian correction, if $x$ is uniformly distributed on the unit simplex, then $z_i$ and $z_j$ are uncorrelated for all $i \ne j$.
+
+\subsubsection{Hyperspherical angular transform}
+
+The map from $y$ to $z$ is elementwise, so its Jacobian is diagonal.
+Each element is
+$$\sin^{-1}(\sqrt{z_i}) = \phi_i = \frac{\pi}{2} \operatorname{logit}^{-1} (y_i).$$
+Differentiating all sides with respect to $y_i$, we find
+$$
+\begin{aligned}
+\frac{1}{\sqrt{1 - z_i^2}} \frac{1}{2 \sqrt{z_i}} \frac{\partial z_i}{\partial y_i} &= \frac{\pi}{2} \operatorname{logit}^{-1} (y_i) (1 - \operatorname{logit}^{-1} (y_i)) = \phi_i \left(1 - \frac{2}{\pi} \phi_i\right)\\
+\frac{\partial z_i}{\partial y_i} &= 2 \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin(\phi_i) \cos(\phi_i).
+\end{aligned}
+$$
+
+As a result, the combined transform $f: y \mapsto x$ is
+\[
+  x_i = \left(1 - \cos^2\left(\frac{\pi}{2} \operatorname{logit}^{-1} (y_i)\right) (1 - \delta_{i,N+1})\right)\left(\prod_{k=1}^{i-1} \sin^2\left(\frac{\pi}{2} \operatorname{logit}^{-1} (y_k)\right)\right),
+\]
+and its inverse is
+\[
+z_i = \operatorname{logit}\left(\frac{2}{\pi}\sin^{-1}\left(\sqrt{1 - \frac{x_i}{\sum_{k=i}^{N+1} x_k}}\right)\right).
+\]
+
+The Jacobian determinant is
+$$
+\begin{aligned}
+    |J_f| &= \prod_{i=1}^N z_i^{N-i} 2 \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin(\phi_i) \cos(\phi_i)\\
+          &= 2^N \prod_{i=1}^N \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin^{2(N-i)+1}(\phi_i) \cos(\phi_i).
+\end{aligned}
+$$
+
+\subsubsection{Hyperspherical logit transform}
+
+For $z_i \in (0, 1]$, $z_i^k \in (0, 1]$ is bijective for $k \ne 0$.
+Let $g(w)$ be the density function for a distribution on $\mathbb{R}$ and $G(w)$ be its cumulative distribution function.
+By the definition of the CDF, $\frac{\mathrm{d}}{\mathrm{d}w} G(w) = g(w)$.
+Let $z_i^{N-i+1} = G(y_i)$.
+
+The combined transform $f: y \mapsto x$ is then
+\[
+  x_i = \left(1-G(y_i)^{1/(N-i+1)}(1 - \delta_{i,N+1})\right)\left(\prod_{k=1}^{i-1} G(y_k)^{1/(N-k+1)}\right),
+\]
+and its inverse is 
+\[
+y_i = G^{-1}\left(\left(1 - \frac{x_i}{\sum_{k=i}^{N+1} x_k}\right)^{N-i+1}\right).
+\]
+
+Then
+$$
+\begin{aligned}
+(N-i+1) z_i^{N-i} \frac{\partial z_i}{\partial y_i} &= g(y_i)\\
+\frac{\partial z_i}{\partial y_i} &= \frac{g(y_i)}{(N-i+1) z_i}.
+\end{aligned}
+$$
+
+As a result,
+$$
+\begin{aligned}
+|J_f| &= \prod_{i=1}^{N} z_i^{N-i} \frac{g(y_i)}{(N-i+1) z_i^{N-i}}\\ 
+&= \left(\prod_{i=1}^N \frac{1}{N-i+1} \right) \prod_{i=1}^{N} g(y_i)\\
+&= \frac{1}{N!} \prod_{i=1}^{N} g(y_i).
+\end{aligned}
+$$
+
+This particular transform allows us to transform the uniform distribution on the simplex to any distribution on $\mathbb{R}^N$ whose coordinates are IID distributed according to $g$.
+
+If we select $G$ to be the logistic function, then $g(w) = \operatorname{logit}^{-1}(w) (1 - \operatorname{logit}^{-1}(w))$ is the density of the logistic distribution with mean $0$ and scale term $1$, and $y_i$ is logistically distributed.
+This choice of $G$ defines the hyperspherical logit transform.
+
+\subsubsection{Hyperspherical probit transform}
+
+Similarly, to define the hyperspherical probit transform, we select $G$ to be the inverse probit function.
+Then $g$ is the density function of the standard normal distribution.
+
+\subsubsection{Transforming the Dirichlet distribution}
+
+Recall that the Dirichlet distribution has density $p_X(x)$ equal to
+\[
+p_X(x) \propto \prod_{i=1}^{N+1} x_i^{\alpha_i - 1}.
+\]
+
+Note that
+\[
+\begin{aligned}
+\prod_{i=1}^{N+1} x_i^{\alpha_i - 1} &= \prod_{i=1}^{N+1} \left(\left(1-G(y_i)^{1/(N-i+1)}(1 - \delta_{i,N+1})\right)\left(\prod_{k=1}^{i-1} G(y_k)^{1/(N-k+1)}\right)\right)^{\alpha_i - 1}\\
+&= \left(\prod_{i=1}^{N} \left(1-G(y_i)^{1/(N-i+1)}\right)^{\alpha_i - 1}\right) \prod_{i=1}^{N+1}\prod_{k=1}^{i-1} G(y_k)^{(\alpha_i - 1)/(N-k+1)}\\
+&= \prod_{i=1}^{N} \left(1-G(y_i)^{1/(N-i+1)}\right)^{\alpha_i - 1} G(y_i)^{-1+\sum_{k=i+1}^{N+1}\alpha_k/(N-i+1)}\\
+\end{aligned}
+\]
+The transformed density on $y$ is then
+\[
+p_Y(y) \propto \prod_{i=1}^{N} \left(1-G(y_i)^{1/(N-i+1)}\right)^{\alpha_i - 1} G(y_i)^{-1+\sum_{k=i+1}^{N+1}\alpha_k/(N-i+1)} g(y_i) = \prod_{i=1}^{N} h(y_i, \alpha_{i:N+1}),
+\]
+where $h$ is a density function.
+Therefore, regardless of the choice of $g$ or $\alpha$, the individual coordinates of the transformed distribution are independently distributed.
 \end{appendices}

### 2022-10-30 notes.tex
Updates from Overleaf

@@
 \prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
+
 \subsection{Additive log ratio transform}
@@
   p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{s}(y)},
+\]
+where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
+and $\absdet{J_s(y)}$ is the absolute value of its determinant.
+
+\subsection{Isometric Log Ratio Transform}
+
+We may define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform.
+By noting that the ALR transformation may be written as
+
+\[
+  \begin{aligned}
+ \textrm{alr}(x) &= \begin{bmatrix}\displaystyle
+    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
+  \end{bmatrix} \\
+  &= \log(\mathbf{x}) \cdot \begin{bmatrix}\displaystyle
+                    1 & 0 & \cdots & 0 \\
+                    0 & 1 & \cdots & 0 \\
+                    \vdots & \vdots & \ddots & \vdots \\
+                    0 & 0 & \cdots & 1 \\
+                    -1 & -1 & \cdots & -1
+                    \end{bmatrix}.
+  \end{aligned}
+\]
+Take any unitary matrix $\mathbf{V} \in D \times D - 1$ then the ILR is defined for $x \in \Delta^{N-1}$ by
+\[
+  \textrm{ilr}(x)
+  = \log(\mathbf{x}) \cdot \mathbf{V}.
+\]
+
+Unitary $\mathbf{V}$ matrices may be constructed through the Gram-Schmidt process or by normalizing the rows of a Helmert contrast matrix using the L2 norm.
+
+The inverse isometric log ratio transform maps values in
+$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
+\mathbb{R}^{N-1}$ by
+\[
+  \textrm{ilr}^{-1}(y)
+  = \textrm{softmax}(\begin{bmatrix} \exp(\mathbf{V^{-1}} \mathbf{x} ) \end{bmatrix}),
+\]
+where for $u \in \mathbb{R}^N$,
+\[
+  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
+\]
+
+The affine transformation of the multiplication by the inverse unitary matrix does not change
+the calculation of the change of variables adjustment proceeds similarly to the ALR case.
+In fact, any affine transformation may be completed at this stage which generalizes the ILR to power transformations as in \cite{tsagris2011data}\footnote{
+   Let $\alpha$ be the scalar power transformation parameter in $\mathbb{R}$ then the inverse \textit{power} based transformation is given by $(V^{-1} \mathbf{x} \alpha + 1) / \alpha$.
+See \cite{tsagris2011data} for more information.
+}.
+
+Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
+we can transform to a density over unconstrained parameters $y \in
+\mathbb{R}^{N-1}$ by applying the inverse ILR transform and adjusting
+for the change of variables, which yields
+\[
+  p_Y(y) = p_X(\textrm{ilr}^{-1}(y)) \absdet{J_{s}(y)},
 \]
@@
 }
-
 For such functions $f$, we can compute a common form of the Jacobian determinant.

### 2022-10-30 paper.tex
Updates from Overleaf

@@
     colorlinks=true,
-        linkcolor=MidnightBlue,
-        anchorcolor=MidnightBlue,
-        citecolor=MidnightBlue,
-        filecolor=MidnightBlue,
-        urlcolor=MidnightBlue,
-        pdftitle={Efficient Unconstraining Transforms for HMC},
-        pdfauthor={Meenal Jhajaria et al.},
-        pdfpagemode=FullScreen,
+    allcolors=MidnightBlue,
+    pdftitle={Efficient unconstraining parameter transforms for Hamiltonian Monte Carlo},
+    pdfpagemode=FullScreen
   }
@@
 
-\title{Efficient Unconstraining
-  Parameter Transforms for Hamiltonian Monte Carlo}
-\author{Meenal Jhajharia \\ \small Flatiron Institute 
-\and Seth Axen \\ \small University of T\"ubingen 
-\and Adam Haber \\ \small Weizmann Institute 
-\and Sean Pinkney \\ \small Omnicom Media Group
-\and Bob Carpenter \\ \small Flatiron Institute}
+\title{Efficient Simplex Parameterizations for Hamiltonian Monte Carlo}
+\author{Meenal Jhajharia
+        \\ \small Center for Computational Mathematics
+        \\ \small Flatiron Institute
+\and \large Seth D. Axen
+     \\ \small Cluster of Excellence Machine Learning
+     \\ \small New Perspectives for Science
+     \\ \small University of T\"ubingen
+\and Adam Haber
+     \\ \small Weizmann Institute of Science
+\and Sean Pinkney
+     \\ \small Center of Excellence
+     \\ \small Omnicom Media Group
+\and Bob Carpenter
+     \\ \small Center for Computational Mathematics
+     \\ \small Flatiron Institute
+}
 \date{DRAFT: \today}
@@
 method in high dimensions is Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc}.
-HMC works by simulating the Hamiltonian dynamics  a fictitious particle
+HMC works by simulating the Hamiltonian dynamics of a fictitious particle
 representing the value being sampled coupled with a momentum term.
@@
 
-Our goal in this paper is to evaluate a standard and less well known transforms for constrained data types in regular use in statistical modeling.
-We will evaluate transforms in the head, body, and tail of distributions in terms of computational efficiency of both the transform, its Jacobian determinant and associated gradients, the geometry induced in the unconstrained space (e.g., log convexity and conditioning), as well as Hamiltonian Monte Carlo sampling efficiency.
+Our goal in this paper is to evaluate both standard and less well known transforms for constrained data types in regular use in statistical modeling.
+We will evaluate transforms in the head, body, and tail of distributions in terms of computational efficiency of the transform, its Jacobian determinant and associated gradients, the geometry induced in the unconstrained space (e.g., log convexity and conditioning), as well as HMC sampling efficiency.
 
-We consider transforms for several constrained data types, including scalars with lower- and/or upper-bounds, vectors with simplex, sum-to-zero or unit length constraints, and matrices (or their Cholesky factors) constrained to be positive definite (e.g., covariance matrices), positive definite with unit diagonal (e.g., correlation matrices), or orthonormal.
+We consider transforms for several constrained data types, including scalars with lower- and/or upper-bounds; vectors with simplex, sum-to-zero, or unit length constraints; and matrices (or their Cholesky factors) constrained to be positive definite (e.g., covariance matrices), positive definite with unit diagonal (e.g., correlation matrices), or orthonormal.
 
@@
 For all of our transforms, we have $\mathcal{Y} = \mathbb{R}^M$ and $\mathcal{X} \subset \mathbb{R}^N$.
-Along with each transform, we have an additional (perhaps improper) density function $h(y)$ defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\mathcal{X}).$
+Along with each transform, we have an additional (perhaps improper) density function $h(y)$ defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.
 
-The above definition is informal in that it's not possible to sample over an improper density.
-Instead, we can characterize our end-to-end goal, which is that if we have a proper density $p_X(x)$ defined for $x \in \mathcal{X}$, then we want the induced distribution on $y$, namely
+The above definition is informal in that it is not possible to sample over an improper density.
+Instead, we can characterize our end-to-end goal, which is that if we have a proper density $p_X(x)$ defined for $x \in \mathcal{X}$, then we want the induced distribution on $y$, namely, the one with the density function
 $$
@@
 
-The traditional way to define constraining transforms is through the usual change-of-variables formula.
+The conventional way to define constraining transforms is through the change-of-variables formula.
 Suppose we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$ and we have a smooth and bijective unconstraining transform $g : \mathcal{X} \rightarrow \mathcal{Y}$.
 We then define our constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$ as the inverse of $g$, $f = g^{-1}$.
-The appropriate density on $\mathcal{Y}$ is then defined as the absolute value of the determinant of the constraining transform
+The induced density on $\mathcal{Y}$ is adjusted by a factor of the absolute value of the determinant of the Jacobian of the constraining transform,
 \[
@@
 \]
-Because we have applied a standard change of variables adjustment, we are guaranteed that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
+With this standard change of variables adjustment, we are guaranteed that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
 
@@
 
-Not all of our transforms are simple bijective changes of variables as described in the previous subsection.
+Not all of our transforms are smooth, bijective changes of variables as described in the previous subsection.
 In some cases, we will take the dimensionality of the unconstrained space $\mathcal{Y}$ to be higher than that of the constrained space $\mathcal{X}$ and thus need additional constraints to ensure the correct distributions on $\mathcal{X}$.
-In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transform coupled with a standard normal distribution over $\mathcal{Y}$,
+In each of these cases, we seek to construct an expanded parameter space $\mathcal{Z} = \mathcal{X} \times \mathcal{R}$ with parameters $z = (x, r)$ for newly introduced constrained parameters $r \in \mathcal{R}$.
+$\mathcal{R}$ is chosen so that the new constrained space $\mathcal{Z}$ has the same dimensionality as $\mathcal{Y}$.
+Then, we choose a prior $p_R(r | x)$ so that
 \[
-  h(y) = \textrm{normal}(y \mid 0, \textrm{I}) \ \absdet{J_f(y)}.
+p_Y(y) = h(y) p_Z(z) = h(y) p_X(x) p_R(r | x).
 \]
-In each case, we will have to ensure that $p_Y(y) = h(y) \, p_X(f(y))$ is such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
+We then choose a bijective transform $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, that allows us to apply the usual bijective change-of-variables approach:
+\[
+h(y) = |J_{\tilde{f}}(y)|
+\]
+Discarding $r$ after sampling is equivalent to marginalizing it out, and in practice, $r$ may never be directly computed.
+It is convenient to choose $p_R(r | x) = p_R(r)$, and we do so in this paper.
+
+% In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transform coupled with a standard normal distribution over $\mathcal{Y}$,
+% \[
+%   h(y) = \textrm{normal}(y \mid 0, \textrm{I}) \, \absdet{J_f(y)},
+% \]  
+% where $\textrm{I}$ is the identity matrix.
+% For each case, we have to ensure that $p_Y(y) = h(y) \, p_X(f(y))$ is such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
 
@@
 
-A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
-values that sums to one.  Simplexes are useful for representations of multinomial probabilities
-(e.g., probabilities of categories in a classification problem). The set of unit $N$-simplexes is conventionally denoted
+Simplexes are useful for representations of multinomial probabilities - points on the standard $N$-simplex in $R^{N+1}$ are the space of possible parameters (probabilities) of the categorical distribution on $n+1$ possible outcomes.
+The set of unit simplexes of dimension $N$ is
 \[
-  \Delta^N = \setcomp{x \in \mathbb{R}_{\ge 0}^{N + 1}}{ \sum_{i=1}^{N} x_i = 1}
+  \Delta^N = \setcomp{x \in \mathbb{R}^{N + 1}}{x_n \geq 0 \textrm{ for } 1 \leq n \leq N+1 \textrm{ and } \sum_{n=1}^{N + 1} x_n = 1}
 \]
-Geometrically, an $N$-simplex is the convex closure of $N+1$ points
-that are 1 in one coordinate and 0 elsewhere.  For example, the
-3-simplex is the complex closure of
+Geometrically, $\Delta^N$ is the convex closure of $N+1$ ``one hot'' points that take value one in one coordinate and zero elsewhere.
+For example, $\Delta^2$ is the convex closure of the three vectors
 $\begin{bmatrix}1 & 0 & 0 \end{bmatrix},
@@
 \[
-  x_N = 1 - (x_1 + x_2 + \cdots + x_{N-1}).
+  x_{N+1} = 1 - (x_1 + x_2 + \cdots + x_N).
 \]
-We will use $\Delta^N_-$ to denote $N$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
-\subsection{Stick-Breaking Transform}
-
-The stick-breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}.
-It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$.
-Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \to  R^{N-1}$, for $1 \leq i \leq N$ as:
+We introduce the notation $\Delta^N_-$ to denote the first $N$ elements of a simplex, i.e., a sequence of non-negative integers that sum to less than or equal to one.
 \[
-y_i
-= \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
-   \right) 
+\Delta^N_-
+= 
+\setcomp{x \in \mathbb{R}^N}{x_n \geq 0 \textrm{ for } 1 \leq n \leq N \textrm{ and } \sum_{n=1}^N x_n \leq 1}.
+\]
+There is a bijective mapping between $\Delta^N_-$ and $\Delta^N$ defined for $x \in \Delta^N_-$ by 
+\[
+f(x) = \begin{bmatrix} x_1 & x_2 & \cdots x_N & 1 - \sum_{n=1}^N x_n \end{bmatrix}.
 \]
 
-for break proportion 
-\[ 
- z_i = \frac{x_i}{1 - \sum_{i' = 1}^{i-1} x_{i'}}.
-\]
+\subsection{Aitchison Geometry}
 
-The inverse transform $ f^{-1} \colon R^{N-1} \to \Delta^{N-1}$ is defined as:
+Compositional data analysis deals with vectors of strictly positive quantities that adhere to a constant sum constraint.
+\cite{aitchison1982statistical} defined compositional data on a simplical space:
+$$
+\Delta^N=\left\{\mathbf{x}=\left[x_1, x_2, \ldots, x_D\right] \in \mathbb{R}^N \mid x_i>0, i=1,2, \ldots, N ; \sum_{i=1}^N x_i=\kappa\right\}
+$$
 
-\[
-x_i = \left( 1 - \sum_{i'=1}^{i-1} x_{i'} \right)\\
-\]
+Relative information enclosed in compositional data is preserved under non-negative scalar multiplication, thus we often consider $\mathcal{S}^D$ to be a standard simplex, where the constant $\kappa$ is 1.
+This can be achieved by normalization, defined by a closure operator in Aitchison Geometry as follows:
+$$
+\mathcal{C}\left[x_1, x_2, \ldots, x_N\right]=\left[\frac{x_1}{\sum_{i=1}^N x_i}, \frac{x_2}{\sum_{i=1}^N x_i}, \ldots, \frac{x_N}{\sum_{i=1}^N x_i}\right].
+$$
 
-for break proportion \[z_i = \mathrm{logit}^{-1} \left( y_i  + \log \left( \frac{1}{N - i}
-                                            \right)\right)
-\]
+Aitchison defines three linear isomorphisms on this simplex - Additive Log Ratio, Centered Log Ratio and Isometric Log Ratio.
+In the rest of the paper, we discuss transforms on the simplex - including these three classical ones or augmented forms of them that are bijective.
 
-The determinant of the Jacobian
-\[
-   \abs{\mathbf{J}} = \prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right)
-\]
-\subsection{Additive log ratio transform}
+\subsubsection{Additive log ratio transform}
 The unconstraining transform for the identified softmax is known as
@@
   = \begin{bmatrix}\displaystyle
-    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
+    \log \frac{x_1}{x_N}, \cdots , \log \frac{x_{N-1}}{x_N}
   \end{bmatrix}
@@
 \[
-  \textrm{softmax}(u) = \frac{exp(u)}{\sum exp(u)}
+  \textrm{softmax}(u) = \frac{\exp(u)}{\sum \exp(u)}
 \]
@@
 \]
-\subsection{Augmented-Softmax Transform}
+
+Geometrically, coordinates obtained from an Additive Log Ratio transform do not correspond to an orthogonal basis, rather it is an oblique basis with respect to the Aitchison metric.
+Also, the transform is not symmetrical in all the N components, rather only the first N-1.
+This distorts euclidean distances measured from the ALR coordinates
+\subsubsection{Stick-Breaking Transform}
+
+The stick-breaking transform is derived from the standard construction for Dirichlet processes \cite{sethuraman1994constructive}.
+The inverse mapping from unconstrained to constrained parameters involves taking a unit-length stick and recursively breaking pieces off until it is divided into $N + 1$ pieces, the lengths of which provide the entries in the simplex.
+Following the transforms used by Stan \cite{stan2022ref}, we add an offset so that the unconstraining transform of the uniform simplex $x = \begin{bmatrix} 1/N & \cdots & 1/N \end{bmatrix}$ maps to $y = \begin{bmatrix} 0 & \cdots & 0 \end{bmatrix}$, a vector of zeros.
+
+\paragraph{Unconstraining transform}
+The unconstraining transform $f : \Delta^{N} \to  \mathbb{R}^N$ is defined so that if $f(x) = y$, then
+\[
+y_n
+= \mathrm{logit}(z_n)
+  - \mbox{log}\left(\frac{1}{N + 1 - n} \right),
+\]
+where 
+\[ 
+ z_n = \frac{x_i}{1 - \sum_{i = 1}^{n-1} x_{i}}.
+\]
+
+\paragraph{Constraining inverse transform}
+The inverse transform follows the stick breaking construction, with $f^{-1} \colon \mathbb{R}^N \to \Delta^N$ defined by $f^{-1}(y) = x$, where
+\[
+x_n = z_n \left( 1 - \sum_{i=1}^{n-1} x_i \right), \\
+\]
+with
+\[
+z_n = \mathrm{logit}^{-1}\!\left(
+y_n  + \log \left( \frac{1}{N - n} \right)
+\right).
+\]
+
+\paragraph{Change-of-variables adjustment} 
+The absolute determinant of the Jacobian of the inverse transform is
+\[
+\abs{J_{f^{-1}}} 
+= \prod_{n=1}^{N-1}
+   z_n \, (1 - z_n)
+   \left( 1 - \sum_{i=1}^{n-1} x_{i} \right).
+\]
+
+% \subsubsection{Additive log ratio transform}
+% The unconstraining transform for the identified softmax is known as
+% the additive log ratio (ALR) transform
+% \cite{aitchison1982statistical}, which is a bijection
+% $\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
+% $x \in \Delta^{N-1}$ by
+% \[
+%   \textrm{alr}(x)
+%   = \begin{bmatrix}\displaystyle
+%     \log \frac{x_1}{x_N}, \cdots , \log \frac{x_{N-1}}{x_N}
+%   \end{bmatrix}
+% \]
+% The inverse additive log ratio transform maps values in
+% $\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
+% \mathbb{R}^{N-1}$ by
+% \[
+%   \textrm{alr}^{-1}(y)
+%   = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
+% \]
+% where for $u \in \mathbb{R}^N$,
+
+% \[
+%   \textrm{softmax}(u) = \frac{\exp(u)}{\sum \exp(u)}
+% \]
+
+% \[
+% \textrm{Here, } \abs{J} = \prod \exp(y)
+%   \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
+% \]
+
+\subsubsection{Isometric Log Ratio Transform}
+
+We may define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform.
+By noting that the ALR transformation may be written as
+
+\[
+  \begin{aligned}
+ \textrm{alr}(x) &= \begin{bmatrix}\displaystyle
+    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
+  \end{bmatrix} \\
+  &= \log(\mathbf{x}) \cdot \begin{bmatrix}\displaystyle
+                    1 & 0 & \cdots & 0 \\
+                    0 & 1 & \cdots & 0 \\
+                    \vdots & \vdots & \ddots & \vdots \\
+                    0 & 0 & \cdots & 1 \\
+                    -1 & -1 & \cdots & -1
+                    \end{bmatrix}.
+  \end{aligned}
+\]
+Take any unitary matrix $\mathbf{V} \in D \times D - 1$ then the ILR is defined for $x \in \Delta^{N-1}$ by
+\[
+  \textrm{ilr}(x)
+  = \log(\mathbf{x}) \cdot \mathbf{V}.
+\]
+
+Unitary $\mathbf{V}$ matrices may be constructed through the Gram-Schmidt process or by normalizing the rows of a Helmert contrast matrix using the L2 norm.
+
+The inverse isometric log ratio transform maps values in
+$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
+\mathbb{R}^{N-1}$ by
+\[
+  \textrm{ilr}^{-1}(y)
+  = \textrm{softmax}(\begin{bmatrix} \exp(\mathbf{V^{-1}} \mathbf{x} ) \end{bmatrix}),
+\]
+where for $u \in \mathbb{R}^N$,
+\[
+  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
+\]
+
+The affine transformation of the multiplication by the inverse unitary matrix does not change
+the calculation of the change of variables adjustment proceeds similarly to the ALR case.
+In fact, any affine transformation may be completed at this stage which generalizes the ILR to power transformations as in \cite{tsagris2011data}\footnote{
+   Let $\alpha$ be the scalar power transformation parameter in $\mathbb{R}$ then the inverse \textit{power} based transformation is given by $(V^{-1} \mathbf{x} \alpha + 1) / \alpha$.
+See \cite{tsagris2011data} for more information.
+}.
+
+Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
+we can transform to a density over unconstrained parameters $y \in
+\mathbb{R}^{N-1}$ by applying the inverse ILR transform and adjusting
+for the change of variables, which yields
+\[
+  p_Y(y) = p_X(\textrm{ilr}^{-1}(y)) \absdet{J_{s}(y)},
+\]
+where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
+and $\absdet{J_s(y)}$ is the absolute value of its determinant.
+
+\subsubsection{Augmented-Softmax Transform}
 %TODO:  Needs to be written like other transforms -> transform, inverse, |J|
-We define the transformation
-$\phi: \mathbb{R}^n \to \Delta^{N} \times \mathbb{R}_{>0}: y \mapsto
-(x_-, r)$, where  $\Delta_{-}^N$ and $x_{-}$  denote $N$ elements of the simplex and $x$, respectively. Here $r = \sum_{i=1}^n \textrm{exp}(y_i)$,
+We define the constraining transformation
+$f: \mathbb{R}^N \to \Delta^{N} \times \mathbb{R}_{>0}: y \mapsto
+(x_-, r)$, where and $x_{-} \in \Delta_{-}^N$. Here $r = \sum_{i=1}^N \textrm{exp}(y_i) > 0$,
 $x_i = \frac{1}{r} \textrm{exp}(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function.
-If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $N$-vector of ones.\[
-  J = (I_{N-1} - x_- \boldsymbol{1}_{N-1}^\top) \operatorname{diag}(x_-),
+If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $N$-vector of ones, then
+\[
+  J_f = (I_{N-1} - x_- \boldsymbol{1}_{N-1}^\top) \operatorname{diag}(x_-),
 \]
@@
 \boldsymbol{1}_{N-1}^\top x_- = 1 - \sum_{i=1}^{N-1} x_i = x_N$, so
-$$ |J| = x_n \prod_{i=1}^{N-1} x_i = \prod_{i=1}^{N} x_i = \exp\left(\sum_{i=1}^{N-1} y_i\right) \left(1 + \sum_{i=1}^{N-1} e^{y_i}\right)^{-N}$$
+$$ |J_f| = x_n \prod_{i=1}^{N-1} x_i = \prod_{i=1}^{N} x_i = \exp\left(\sum_{i=1}^{N-1} y_i\right) \left(1 + \sum_{i=1}^{N-1} \exp(y_i)\right)^{-N}$$
+
+\subsubsection{Hyperspherical coordinate transforms}
+
+Let $u$ be a unit vector with $N+1$ elements in the positive orthant of the unit sphere $\mathbb{S}_{>0}^N$, which has the constraints $\sum_{i=1}^{N+1} u_i^2 = 1$ and $u_i > 0$.
+There is a unique bijective map $f_1\colon \mathbb{S}_{>0}^N \to \Delta^N \colon u \mapsto x$ given by $x_i = u_i^2$.
+We then define $f_2$ as the map from hyperspherical coordinates to Cartesian coordinates of the hypersphere.
+Taking $z_i = \sin^2(\phi_i) \in (0, 1]$ for $\phi_i \in (0, \frac{\pi}{2}]$, the composed map $f_1 \circ f_2$ is written
+
+\[
+  x_i = (1 - z_i (1 - \delta_{i,N+1})) \prod_{k=1}^{i-1} z_k.
+\]
+
+The inverse of this transform is
+$$z_i = 1 - \frac{x_i}{1 - \sum_{k=1}^{i - 1} x_k}.$$
+
+To complete the transform $f = f_1 \circ f_2 \circ f_3$, we must select a constraining transform $f_3: y \mapsto \phi$.
+We consider 3 different approaches.
+
+\subsubsection{Hyperspherical angular transform}
+
+Here we use a (scaled) logistic function to map directly from the unconstrained space to the constrained angle:
+$$\sin^{-1}(\sqrt{z_i}) = \phi_i = \frac{\pi}{2} \operatorname{logit}^{-1} (y_i).$$
+
+In terms of $\phi$,
+
+$$
+|J_f| = 2^N \prod_{i=1}^N \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin^{2(N-i)+1}(\phi_i) \cos(\phi_i).
+$$
+This transform was discussed in \cite{betancourt2012cruising}.
+
+\subsubsection{Hyperspherical logit transform}
+
+If we define $f_3$ to map from the unconstrained space directly to a conveniently-chosen power of $z_i$ function using the logistic function 
+$$z_i^{N-i+1} = \operatorname{logit}^{-1} (y_i),$$
+then the Jacobian determinant is
+$$
+|J_f| = \frac{1}{N!} \prod_{i=1}^N \operatorname{logit}^{-1}(y_i) (1 - \operatorname{logit}^{-1}(y_i)).
+$$
+If $x$ is uniformly distributed on the unit simplex, then $y_i$ is logistic-distributed with mean 0 and variance $\frac{\pi^2}{3}$.
+
+\subsubsection{Hyperspherical probit transform}
+
+Similarly, if we use the inverse probit function instead of the inverse logit, then $f_3$ is
+$$z_i^{N-i+1} = \operatorname{probit}^{-1} (y_i),$$
+where $\operatorname{probit}^{-1}(\cdot)$ is the cumulative distribution function of the standard normal distribution.
+The resulting Jacobian determinant is
+\[
+|J_f| = \frac{1}{N!} \prod_{i=1}^N \mathcal{N}(y_i),
+\]
+where $\mathcal{N}(\cdot)$ is the density function of the standard normal distribution.
+As a result, if $x$ is uniformly distributed on the unit simplex, then $y_i$ is standard normally distributed.
 
@@
     \centering
-    \begin{tabular}{|l|l|l|l|l|}
+    \begin{tabular}{r|r|r|r|r|r|}
     \hline
-        ~ & stan & softmax-augmented & softmax & stickbreaking \\ \hline
-        $\alpha = 0.1, \, N = 10$ & 6.908s & 11.712s & 10.548s & 13.386s \\ 
-        $\alpha = 0.1, \, N = 100$ & 74.274s & 33.757s & 61.826s & 159.016s \\ 
-        $\alpha = 0.1, \, N = 1000$ & 1108.48s & 307.463s & 657.791s & 11773.4s \\ 
-        $\alpha = 1, \, N = 10$ & 4.072s & 6.328s & 5.337s & 5.67s \\ 
-        $\alpha = 1, \, N = 100$ & 20.805s & 27.151s & 25.94s & 51.45s \\ 
-        $\alpha = 1, \, N = 1000$ & 306.942s & 171.824s & 184.126 & 3023.24 \\ 
-        $\alpha = 10, \, N = 10$ & 4.123s & 6.375s & 5.563s & 5.606s \\ 
-        $\alpha = 10, \, N = 100$ & 23.6s & 29.091s & 30.747s & 48.834s \\ 
-        $\alpha = 10, \, N = 1000$ & 205s.627 & 161.11s & 150.751s & 1866.83s \\ \hline
+        $\alpha$ & $N$ & Stan & Augmented-Softmax & softmax & stickbreaking \\ \hline
+        0.1 & 10  & 6.9s & 11.712s & 10.548s & 13.386s \\ 
+        0.1 & 100 & 74.274s & 33.757s & 61.826s & 159.016s \\ 
+        0.1 & 1000 & 1108.48s & 307.463s & 657.791s & 11773.4s \\ 
+        1 & 10 & 4.072s & 6.328s & 5.337s & 5.67s \\ 
+        1 & 100 & 20.805s & 27.151s & 25.94s & 51.45s \\ 
+        1, & 1000 & 306.942s & 171.824s & 184.126 & 3023.24 \\ 
+        10 & 10 & 4.123s & 6.375s & 5.563s & 5.606s \\ 
+        10 & 100 & 23.6s & 29.091s & 30.747s & 48.834s \\ 
+        10 & 1000 & 205s.627 & 161.11s & 150.751s & 1866.83s \\ \hline
+    \end{tabular}
+\end{table}
+
+\begin{table}[!ht]
+    \centering
+    \begin{tabular}{c|r||r|r|r|r|}
+        $\alpha$ & $N$ & stan & aug-softmax & softmax & stick-break \\ \hline 
+        0.1 & 10  & 7s & 11s & 11s & 13s \\ 
+        0.1 & 100 & 74s & 34s & 62s & 160s \\ 
+        0.1 & 1000 & 1100s & 300s & 660s & 12000s \\ 
+        1 & 10 & 4s & 6s & 5s & 6s \\ 
+        1 & 100 & 21s & 27s & 26s & 51s \\ 
+        1 & 1000 & 300s & 170s & 180s & 3000s \\ 
+        10 & 10 & 4s & 6s & 6s & 5s \\ 
+        10 & 100 & 24s & 29s & 31s & 49s \\ 
+        10 & 1000 & 200s & 160s & 150s & 1900s \\
     \end{tabular}
@@
     \includegraphics[width=1.2\textwidth]{figures/simplex/ess_density.png}
-    \caption{Probability Density plot of Effective Sample Size/Total LeapFrog Steps.
-Effective sample size is evaluated at 100 runs, each point is the value of ESS/Total Leapfrog steps taken by the sampler in one run.}
+    \caption{Probability Density Plot of Effective Sample Size/Total Leapfrog Steps.
+Effective sample size is evaluated at 100 runs, and each point is the value of ESS/Total Leapfrog steps taken by the sampler in one run.}
     \label{fig:ess_density}
@@
     \includegraphics[width=1.2\textwidth]{figures/simplex/ess_cdf.png}
-    \caption{Cumulative Density plot of Effective Sample Size/Total LeapFrog Steps.
+    \caption{Cumulative Density Plot of Effective Sample Size/Total Leapfrog Steps.
 Effective sample size is evaluated at 100 runs, each point is the value of ESS/Total Leapfrog steps taken by the sampler in one run.}
@@
 easy-to-use symbolic matrix derivative calculator.
+This work was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) under Germany’s Excellence Strategy – EXC number 2064/1 – Project number 390727645.
 

### 2022-11-04 paper.tex
some rewriting

@@
   \noindent
-  This paper evaluates the statistical and computational
-  efficiency of unconstraining parameter transforms for Hamiltonian
-  Monte Carlo sampling.
+  We consider the task of sampling on a unit $N$ - Simplex using Hamiltonian Monte Carlo.
+In bayesian computation, parameters are often constrained on a simplex, e.g.
+for modelling categorical data.
+Unconstraining transforms mapping a unit $N$ - Simplex to $R^N$ are used to avoid the challenges of constructing HMC samplers for a constrained space.
+Arbitrary transformations can serve as this mapping, not all of them work well when sampled from - owing to weak convexity, incompatible posterior geometry, numerical instability among other reasons.
+In this paper we evaluate the statistical properties,  computational efficiency and geometry for unconstraining transforms between a unit $N$ - simplex and $R^N$.
+
   \end{abstract}
@@
 
-In statistical computing, we often need to compute high-dimensional
-integrals over densities $\pi(x)$ (e.g., Bayesian estimation or
-prediction, $p$-value calculations, etc.).  The only black-box
-techniques that work for general high-dimensional integrals are
-Markov chain Monte Carlo (MCMC) methods.  The most effective MCMC
-method in high dimensions is Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc}.
-HMC works by simulating the Hamiltonian dynamics of a fictitious particle
-representing the value being sampled coupled with a momentum term.
+In statistical computing, we often need to compute high-dimensional integrals over densities $\pi(x)$ (e.g., Bayesian estimation or prediction, $p$-value calculations, etc.).
+The only black-box techniques that work for general high-dimensional integrals are Markov chain Monte Carlo (MCMC) methods.
+The most effective MCMC method in high dimensions is Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc}.
+HMC works by simulating the Hamiltonian dynamics of a fictitious particle representing the value being sampled coupled with a momentum term.
 
-Although it is possible to write HMC samplers that work for simply
-constrained values such as upper- and/or lower-bounds
-\cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is
-much more challenging to do the same for complex constraints such as
-simplexes or positive definite matrices or for densities involving
-multiple constrained values.  Instead, it is far more common to map
-the constrained values to unconstrained values before sampling
-\cite{JSSv076i01, radul2021base, fjelde2020bijectors}.
+Although it is possible to construct HMC samplers that work for simple constraints such as upper- and/or lower-bounds \cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is much more challenging to do the same for complex constraints such as simplexes, positive definite matrices or for densities involving multiple constrained values.
+Instead, it is far more common to map the constrained values to unconstrained values before sampling \cite{JSSv076i01, radul2021base, fjelde2020bijectors}.
 
-Our goal in this paper is to evaluate both standard and less well known transforms for constrained data types in regular use in statistical modeling.
-We will evaluate transforms in the head, body, and tail of distributions in terms of computational efficiency of the transform, its Jacobian determinant and associated gradients, the geometry induced in the unconstrained space (e.g., log convexity and conditioning), as well as HMC sampling efficiency.
+This approach is not without ramifications, posterior geometry obtained from a transformed variable can be very different from the original.
+In context of Hamiltonian Monte Carlo, there is a lack of formal analysis or comparison for these transformations.
+We evaluate both standard and relatively unvisited transforms for constrained data types that find regular use in statistical modeling.
+Evaluation takes into account specificities such as head, body, and tail of distributions in terms of computational efficiency of the transform, its Jacobian determinant and associated gradients, the geometry induced in the unconstrained space (e.g., log convexity and conditioning), as well as HMC sampling efficiency.
 
-We consider transforms for several constrained data types, including scalars with lower- and/or upper-bounds; vectors with simplex, sum-to-zero, or unit length constraints; and matrices (or their Cholesky factors) constrained to be positive definite (e.g., covariance matrices), positive definite with unit diagonal (e.g., correlation matrices), or orthonormal.
-
+In this paper, we focus on transforms on a unit $N$-simplex.
+Simplexes are useful for representations of multinomial probabilities - points on the standard $N$-simplex in $R^{N+1}$ are the space of possible parameters (probabilities) of the categorical distribution on $n+1$ possible outcomes.
+Lastly, we define Aitchison Geometry and compare metric distortions of canonical transforms (or their augmented forms) in Compositional Data Analysis.
 \section{Constraining transforms}
 
-In this paper, we restrict our attention to a general class of surjective constraining transforms $f : \mathcal{Y} \rightarrow \mathcal{X}$ from an unconstrained space $\mathcal{Y}$ onto a constrained space $\mathcal{X}$.
-For all of our transforms, we have $\mathcal{Y} = \mathbb{R}^M$ and $\mathcal{X} \subset \mathbb{R}^N$.
-Along with each transform, we have an additional (perhaps improper) density function $h(y)$ defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.
-
-The above definition is informal in that it is not possible to sample over an improper density.
-Instead, we can characterize our end-to-end goal, which is that if we have a proper density $p_X(x)$ defined for $x \in \mathcal{X}$, then we want the induced distribution on $y$, namely, the one with the density function
-$$
-p_Y(y) = h(y) \, p_X(f(y)),
-$$
-to be proper and be such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
+A surjective constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$ is mapped from an unconstrained space $\mathcal{Y} = \mathbb{R}^M$ onto a constrained space $\mathcal{X} \subset \mathbb{R}^N$, and $h(y)$ is a proper density defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.
+It is not possible to sample over an improper density.
+To this end, we stipulate the induced distribution on $y$ to be proper, given a proper density $p_X(x)$ defined for $x \in \mathcal{X}$ such that if $y \sim p_Y(\cdot)$ then $f(y) \sim p_X(\cdot)$ .
+In this case, density of the distribution induced on $y$ is
+\[
+p_Y(y) = h(y) \, p_X(f(y))
+\]
 
@@
 
-The conventional way to define constraining transforms is through the change-of-variables formula.
-Suppose we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$ and we have a smooth and bijective unconstraining transform $g : \mathcal{X} \rightarrow \mathcal{Y}$.
-We then define our constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$ as the inverse of $g$, $f = g^{-1}$.
-The induced density on $\mathcal{Y}$ is adjusted by a factor of the absolute value of the determinant of the Jacobian of the constraining transform,
+ Suppose, we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$.
+A smooth, bijective unconstraining transform is defined as $g : \mathcal{X} \rightarrow \mathcal{Y}$, and $g^{-1}$ is the constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$.
+Using the standard change of variables, the induced density on $\mathcal{Y}$ is adjusted by a constant factor
+
+ \[\absdet{J_f(y)} = \absdet{\nabla_y \, f(y)}.
+ \]
+Thereby, for any given $p_X(x)$,
+
  \[
-h(y) = \absdet{J_f(y)}.
+ y \sim p_Y(\cdot) \implies f(y) \sim p_X(\cdot) \, \, \text{where} \, \, p_Y(y) = p_X(f(y)) \, \absdet{J_f(y)}
 \]
-That way, for any given $p_X(x)$, we have
-\[
-p_Y(y) = p_X(f(y)) \, \absdet{J_f(y)},
-\]
-where the Jacobian of the inverse transform is defined by
-\[
-  J_{f^{-1}}(y) \ = \ \nabla_y \, f^{-1}(y) \ = \ \frac{\partial}{\partial y} \, f^{-1}(y).
-\]
-With this standard change of variables adjustment, we are guaranteed that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.
 
@@
 
-Not all of our transforms are smooth, bijective changes of variables as described in the previous subsection.
-In some cases, we will take the dimensionality of the unconstrained space $\mathcal{Y}$ to be higher than that of the constrained space $\mathcal{X}$ and thus need additional constraints to ensure the correct distributions on $\mathcal{X}$.
-In each of these cases, we seek to construct an expanded parameter space $\mathcal{Z} = \mathcal{X} \times \mathcal{R}$ with parameters $z = (x, r)$ for newly introduced constrained parameters $r \in \mathcal{R}$.
-$\mathcal{R}$ is chosen so that the new constrained space $\mathcal{Z}$ has the same dimensionality as $\mathcal{Y}$.
-Then, we choose a prior $p_R(r | x)$ so that
+For a constraining transform $f: R^\mathcal{M} \rightarrow R^\mathcal{N}$, if $\mathcal{M} >  \mathcal{N}$, additional constraints are needed to ensure the correct distributions on $\mathcal{X}$.
+Bijective change-of-variables method cannot be applied here, instead an expanded parameter space $\mathcal{Z} = \mathcal{X} \times \mathcal{R}$ is constructed with parameters $z = (x, r)$.
+Here, $r \in \mathcal{R}$ is introduced such that $\mathcal{Y}$ and $\mathcal{Z}$ have the same dimensionality.
+Taking prior $p_R(r | x)$,
 \[
@@
 \]
-We then choose a bijective transform $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, that allows us to apply the usual bijective change-of-variables approach:
+We define a bijective transform $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, this permits obtaining change-of-variables,
 \[
-h(y) = |J_{\tilde{f}}(y)|
+h(y) = |J_{\tilde{f}}(y)|.
 \]
 Discarding $r$ after sampling is equivalent to marginalizing it out, and in practice, $r$ may never be directly computed.
-It is convenient to choose $p_R(r | x) = p_R(r)$, and we do so in this paper.
+On acount of convenience,  we choose $p_R(r | x) = p_R(r)$ in the rest of the paper.
 
@@
 
-\section{Vector transforms}
+\section{Unit simplex}
 
-In this section, we consider transforms from unconstrained space to constrained vectors, including unit-simplex, sum-to-zero, and unit-length constraints.
-
-\subsection{Unit simplex}
-
-Simplexes are useful for representations of multinomial probabilities - points on the standard $N$-simplex in $R^{N+1}$ are the space of possible parameters (probabilities) of the categorical distribution on $n+1$ possible outcomes.
 The set of unit simplexes of dimension $N$ is
 \[
-  \Delta^N = \setcomp{x \in \mathbb{R}^{N + 1}}{x_n \geq 0 \textrm{ for } 1 \leq n \leq N+1 \textrm{ and } \sum_{n=1}^{N + 1} x_n = 1}
+  \Delta^N = \setcomp{x \in \mathbb{R}^{N + 1}}{\sum_{i=1}^{N + 1} x_i = 1,\,x_i \ge 0 \, \, \forall i }
 \]
-Geometrically, $\Delta^N$ is the convex closure of $N+1$ ``one hot'' points that take value one in one coordinate and zero elsewhere.
+An affine transformation $(x_0, \ldots, x_n) \mapsto \sum_{i=0}^n x_i v_i \,$ maps a unit $N$-simplex to an arbitrary $N$-simplex.
+Geometrically, $\Delta^N$ is the convex hull of $N+1$ standard unit vectors comprising the vertices.
 For example, $\Delta^2$ is the convex closure of the three vectors
@@
 \]
-We introduce the notation $\Delta^N_-$ to denote the first $N$ elements of a simplex, i.e., a sequence of non-negative integers that sum to less than or equal to one.
+We introduce the notation $\Delta^N_-$ to denote the probability simplex, which is a face of the unit simplex (each face of an $N$ - simplex is an affine $N-1$ - simplex),
+
 \[

### 2022-11-11 paper.tex
more updates

@@
 Lastly, we define Aitchison Geometry and compare metric distortions of canonical transforms (or their augmented forms) in Compositional Data Analysis.
-\section{Constraining transforms}
+\section{Theory of constrained parametrizations}
 
-A surjective constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$ is mapped from an unconstrained space $\mathcal{Y} = \mathbb{R}^M$ onto a constrained space $\mathcal{X} \subset \mathbb{R}^N$, and $h(y)$ is a proper density defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.
+A surjective constraining transformation $f : \mathcal{Y} \rightarrow \mathcal{X}$ is mapped from an unconstrained space $\mathcal{Y} = \mathbb{R}^M$ onto a constrained space $\mathcal{X} \subset \mathbb{R}^N$, and $h(y)$ is a proper density defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.
 It is not possible to sample over an improper density.
@@
  Suppose, we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$.
-A smooth, bijective unconstraining transform is defined as $g : \mathcal{X} \rightarrow \mathcal{Y}$, and $g^{-1}$ is the constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$.
+A smooth, bijective unconstraining transformationis defined as $g : \mathcal{X} \rightarrow \mathcal{Y}$, and $g^{-1}$ is the constraining transformation $f : \mathcal{Y} \rightarrow \mathcal{X}$.
 Using the standard change of variables, the induced density on $\mathcal{Y}$ is adjusted by a constant factor
@@
 
-For a constraining transform $f: R^\mathcal{M} \rightarrow R^\mathcal{N}$, if $\mathcal{M} >  \mathcal{N}$, additional constraints are needed to ensure the correct distributions on $\mathcal{X}$.
+For a constraining transformation $f: R^\mathcal{M} \rightarrow R^\mathcal{N}$, if $\mathcal{M} >  \mathcal{N}$, additional constraints are needed to ensure the correct distributions on $\mathcal{X}$.
 Bijective change-of-variables method cannot be applied here, instead an expanded parameter space $\mathcal{Z} = \mathcal{X} \times \mathcal{R}$ is constructed with parameters $z = (x, r)$.
@@
 \]
-We define a bijective transform $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, this permits obtaining change-of-variables,
+We define a bijective transformation $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, this permits obtaining change-of-variables,
 \[
@@
 Discarding $r$ after sampling is equivalent to marginalizing it out, and in practice, $r$ may never be directly computed.
-On acount of convenience,  we choose $p_R(r | x) = p_R(r)$ in the rest of the paper.
+On account of convenience,  we choose $p_R(r | x) = p_R(r)$ in the rest of the paper.
 
-% In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transform coupled with a standard normal distribution over $\mathcal{Y}$,
+% In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transformationcoupled with a standard normal distribution over $\mathcal{Y}$,
 % \[
@@
 
-\section{Unit simplex}
+\section{Simplex}
 
@@
 Compositional data analysis deals with vectors of strictly positive quantities that adhere to a constant sum constraint.
-\cite{aitchison1982statistical} defined compositional data on a simplical space:
+Aitchison\cite{aitchison1982statistical} defined compositional data on a simplical space:
 $$
@@
 
-\subsubsection{Additive log ratio transform}
-The unconstraining transform for the identified softmax is known as
-the additive log ratio (ALR) transform
+\section{Reparametrizations for sampling from a simplex}
+
+\subsection{Additive Log Ratio}
+\paragraph{Unconstraining transformation}
+The unconstraining transformation for the identified softmax is known as
+the additive log ratio (ALR) parametrization
 \cite{aitchison1982statistical}, which is a bijection
@@
 \]
-The inverse additive log ratio transform maps values in
+\paragraph{Constraining transformation}
+The inverse additive log ratio transformation maps values in
 $\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
@@
 
+\paragraph{Change-of-variables adjustment}
 \[
@@
 
-Geometrically, coordinates obtained from an Additive Log Ratio transform do not correspond to an orthogonal basis, rather it is an oblique basis with respect to the Aitchison metric.
-Also, the transform is not symmetrical in all the N components, rather only the first N-1.
-This distorts euclidean distances measured from the ALR coordinates
-\subsubsection{Stick-Breaking Transform}
+\subsection{Stick-Breaking}
 
-The stick-breaking transform is derived from the standard construction for Dirichlet processes \cite{sethuraman1994constructive}.
+The stick-breaking transformation is derived from the standard construction for Dirichlet processes \cite{sethuraman1994constructive}.
 The inverse mapping from unconstrained to constrained parameters involves taking a unit-length stick and recursively breaking pieces off until it is divided into $N + 1$ pieces, the lengths of which provide the entries in the simplex.
-Following the transforms used by Stan \cite{stan2022ref}, we add an offset so that the unconstraining transform of the uniform simplex $x = \begin{bmatrix} 1/N & \cdots & 1/N \end{bmatrix}$ maps to $y = \begin{bmatrix} 0 & \cdots & 0 \end{bmatrix}$, a vector of zeros.
+Following the transforms used by Stan \cite{stan2022ref}, we add an offset so that the uniform simplex $x = \begin{bmatrix} 1/N & \cdots & 1/N \end{bmatrix}$ maps to $y = \begin{bmatrix} 0 & \cdots & 0 \end{bmatrix}$, a vector of zeros.
 
 \paragraph{Unconstraining transform}
-The unconstraining transform $f : \Delta^{N} \to  \mathbb{R}^N$ is defined so that if $f(x) = y$, then
+The unconstraining transformation $f : \Delta^{N} \to  \mathbb{R}^N$ is defined so that if $f(x) = y$, then
 \[
@@
 \paragraph{Constraining inverse transform}
-The inverse transform follows the stick breaking construction, with $f^{-1} \colon \mathbb{R}^N \to \Delta^N$ defined by $f^{-1}(y) = x$, where
+The inverse transformation follows the stick breaking construction, with $f^{-1} \colon \mathbb{R}^N \to \Delta^N$ defined by $f^{-1}(y) = x$, where
 \[
@@
 \paragraph{Change-of-variables adjustment} 
-The absolute determinant of the Jacobian of the inverse transform is
+The absolute determinant of the Jacobian of the inverse transformationis
 \[
@@
 
-% \subsubsection{Additive log ratio transform}
-% The unconstraining transform for the identified softmax is known as
+% \subsection{Additive log ratio transform}
+% The unconstraining transformationfor the identified softmax is known as
 % the additive log ratio (ALR) transform
@@
 % \]
-% The inverse additive log ratio transform maps values in
+% The inverse additive log ratio transformationmaps values in
 % $\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
@@
 
-\subsubsection{Isometric Log Ratio Transform}
+\subsection{Isometric Log Ratio Transform}
 
-We may define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform.
+\paragraph{Unconstraining transformation}
+We define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform.
 By noting that the ALR transformation may be written as
@@
 
-The inverse isometric log ratio transform maps values in
+\paragraph{Constraining transformation}
+The inverse isometric log ratio transformationmaps values in
 $\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
@@
 }. 
-
+\paragraph{Change-of-variables adjustment}
 Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
-we can transform to a density over unconstrained parameters $y \in
-\mathbb{R}^{N-1}$ by applying the inverse ILR transform and adjusting
+we can transform it to a density over unconstrained parameters $y \in
+\mathbb{R}^{N-1}$ by applying the inverse ILR transformation and adjusting
 for the change of variables, which yields
@@
 \]
-where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
-and $\absdet{J_s(y)}$ is the absolute value of its determinant.
 
-\subsubsection{Augmented-Softmax Transform}
+\subsection{Augmented-Softmax Transform}
 %TODO:  Needs to be written like other transforms -> transform, inverse, |J|
@@
 
-\subsubsection{Hyperspherical coordinate transforms}
-
+\subsection{Hyperspherical coordinate transforms}
+\paragraph{Unconstraining transformation}
 Let $u$ be a unit vector with $N+1$ elements in the positive orthant of the unit sphere $\mathbb{S}_{>0}^N$, which has the constraints $\sum_{i=1}^{N+1} u_i^2 = 1$ and $u_i > 0$.
@@
 \]
+\paragraph{Constraining transformation}
 
-The inverse of this transform is
+The inverse of this transformation is
 $$z_i = 1 - \frac{x_i}{1 - \sum_{k=1}^{i - 1} x_k}.$$
 
-To complete the transform $f = f_1 \circ f_2 \circ f_3$, we must select a constraining transform $f_3: y \mapsto \phi$.
+To complete the transformation $f = f_1 \circ f_2 \circ f_3$, we must select a constraining transformation $f_3: y \mapsto \phi$.
 We consider 3 different approaches.
 
-\subsubsection{Hyperspherical angular transform}
+\subsection{Hyperspherical angular transform}
 
@@
 $$\sin^{-1}(\sqrt{z_i}) = \phi_i = \frac{\pi}{2} \operatorname{logit}^{-1} (y_i).$$
+
+\paragraph{Change-of-variables adjustment}
 
@@
 $$
-This transform was discussed in \cite{betancourt2012cruising}.
+This transformation was discussed in \cite{betancourt2012cruising}.
 
-\subsubsection{Hyperspherical logit transform}
+\subsection{Hyperspherical logit transform}
 
-If we define $f_3$ to map from the unconstrained space directly to a conveniently-chosen power of $z_i$ function using the logistic function 
+We define $f_3$ to map from the unconstrained space directly to a conveniently-chosen power of $z_i$ function using the logistic function 
 $$z_i^{N-i+1} = \operatorname{logit}^{-1} (y_i),$$
-then the Jacobian determinant is
+\paragraph{Change-of-variables adjustment}
 $$
@@
 
-\subsubsection{Hyperspherical probit transform}
+\subsection{Hyperspherical probit transform}
 
@@
 where $\operatorname{probit}^{-1}(\cdot)$ is the cumulative distribution function of the standard normal distribution.
-The resulting Jacobian determinant is
+\paragraph{Change-of-variables adjustment}
+
 \[
@@
 
+\section{Metric Geometry of reparametrizations}
+
+Geometrically, coordinates obtained from an Additive Log Ratio transformationdo not correspond to an orthogonal basis, rather it is an oblique basis with respect to the Aitchison metric.
+Also, the transformationis not symmetrical in all the N components, rather only the first N-1.
+This distorts euclidean distances measured from the ALR coordinates.
 %TODO: better captions
-\section{Results}
+\section{Evaluations}
 
@@
 \end{figure}
+\subsection{Squared erorr}
+\subsection{Effective Sample Size}
+\subsection{Transient bias}
+\subsection{Autocorrelation time at stationarity}
+\subsection{Approximation quality of Laplace at mode in unconstrained  
+space}
 
-\subsubsection*{Acknowledgements}
+\section{Conclusion}
+\subsection*{Acknowledgements}
 

### 2022-11-11 appendix.tex

@@
 \begin{appendices}
+\section{Aitchison Geometry}
+\textbf{Note for coauthors: These are taken from notes on Aitchison Geometry, more trimming and rewriting needs to happen to make them relevant + citing.
+Also, here $\mathcal{S}^N$ is the probability simplex $\Delta^{N}$ which is equivalent to the face of a unit simplex $\Delta^{N}$.\\}
+
+A vector $\mathbf{x}=\left[x_1, x_2, \ldots, x_N\right]$ is a composition if $x_i \in \mathbb{R}_{+}^N$ and each $x_i$ represents relative information here.
+For any vector of $N$ real positive components, the closure of $\mathbf{x}$ is defined as
+$$
+\mathcal{C}(\mathbf{x})=\left[\frac{\kappa \cdot z_1}{\sum_{i=1}^N z_i}, \frac{\kappa \cdot z_2}{\sum_{i=1}^N z_i}, \cdots, \frac{\kappa \cdot z_N}{\sum_{i=1}^N z_i}\right] .
+$$
+The result is the same vector re-scaled so that the sum of its components is $\kappa$.\\
+
+\textbf{General principles}
+
+\begin{itemize}
+	\item Scale Invariance: Two compositions $\mathbf{x}, \mathbf{y}$ are equivalent if there exists a positive scalar $\lambda \in \mathbb{R}^{+}$such that $\mathbf{x}=\lambda \cdot \mathbf{y}$ and, equivalently, $\mathcal{C}(\mathbf{x})=\mathcal{C}(\mathbf{y})$.
+For a scale invariant function $f(.)$ and a positive scalar $\lambda$, we notice $f(\lambda \mathbf{x})=$ $f(\mathbf{x})$.
+Common choices of such functions are log-ratios of the parts in $\mathbf{x}$.
+The ratio $f(\mathbf{x})=x_1 / x_2=\left(\lambda \cdot x_1\right) /\left(\lambda \cdot x_2\right)$ is scale invariant.
+	\item A function is permutation-invariant if it yields equivalent results when the ordering of the parts in the composition is changed. 
+	\item Subcompositional coherence: Composition results obtained for a subset of parts of a composition (a subcomposition) should remain the same as in a composition
+
+\end{itemize}
+Perturbation of a composition $\mathbf{x} \in \mathcal{S}^N$ by a composition $\mathbf{y} \in$ $\mathcal{S}^N$
+$$
+\mathbf{x} \oplus \mathbf{y}=\mathcal{C}\left[x_1 y_1, x_2 y_2, \ldots, x_N y_N\right]
+$$
+Power transformation or powering of a composition $\mathbf{x} \in \mathcal{S}^N$ by a constant $\alpha \in \mathbb{R}$,
+$$
+\alpha \odot \mathbf{x}=\mathcal{C}\left[x_1^\alpha, x_2^\alpha, \ldots, x_N^\alpha\right] .
+$$
+The simplex,$\left(\mathcal{S}^N, \oplus, \odot\right)$, with perturbation and powering, is a vector space.
+This means the following properties hold, making them analogous to translation and scalar multiplication:
+$\left(\mathcal{S}^N, \oplus\right)$ has a commutative group structure
+Powering satisfies the properties of an external product.
+
+Inner product of $\mathbf{x}, \mathbf{y} \in \mathcal{S}^N$,
+$$
+\langle\mathbf{x}, \mathbf{y}\rangle_a=\frac{1}{2 D} \sum_{i=1}^N \sum_{j=1}^N \ln \frac{x_i}{x_j} \ln \frac{y_i}{y_j}
+$$
+
+Norm of $\mathbf{x} \in \mathcal{S}^N$
+$$
+\|\mathbf{x}\|_a=\sqrt{\frac{1}{2 D} \sum_{i=1}^N \sum_{j=1}^N\left(\ln \frac{x_i}{x_j}\right)^2}
+$$
+
+Distance between $\mathbf{x}$ and $\mathbf{y} \in \mathcal{S}^N$
+$$
+d_a(\mathbf{x}, \mathbf{y})=\|\mathbf{x} \ominus \mathbf{x}\|_a=\sqrt{\frac{1}{2 D} \sum_{i=1}^N \sum_{j=1}^N\left(\ln \frac{x_i}{x_j}-\ln \frac{y_i}{y_j}\right)^2}
+$$
+The algebraic-geometric structure of $\mathcal{S}^N$ satisfies standard properties, like compatibility of the distance with perturbation and powering, i.e.
+$$
+\mathrm{N}_a(\mathbf{p} \oplus \mathbf{x}, \mathbf{p} \oplus \mathbf{y})=\mathrm{N}_a(\mathbf{x}, \mathbf{y}), \quad \mathrm{N}_a(\alpha \odot \mathbf{x}, \alpha \odot \mathbf{y})=|\alpha| \mathrm{N}_a(\mathbf{x}, \mathbf{y})
+$$
+for any $\mathbf{x}, \mathbf{y}, \mathbf{p} \in \mathcal{S}^N$ and $\alpha \in \mathbb{R}$. Other typical properties of metric spaces are valid for $\mathcal{S}^N$.
+
+\subsection{Log ratios}
+Given a composition $\mathbf{x}=\left[\begin{array}{llll}x_1 & x_2 & \cdots & x_N\end{array}\right]$, a general LR is defined as
+$$
+\begin{aligned}
+\operatorname{LR}(j, k)= & \log \left(x_j / x_k\right), \text{where } j \neq k \text{and } j, k \in\{1, \ldots, N\}\\
+=&  \log \left(x_j\right)-\log \left(x_k\right)
+\end{aligned}
+$$
+\subsubsection{ALR}
+$\begin{aligned} \operatorname{ALR}(\mathbf{x}) &=\left[\ln \left(\frac{x_1}{x_N}\right), \ldots, \ln \left(\frac{x_{N-1}}{x_N}\right)\right]^T \\ \operatorname{ALR}^{-1}(\mathbf{y}) &=\mathcal{C}\left(\left[\exp \left(y_1\right), \ldots, \exp \left(y_{N-1}\right), 1\right]^T\right) \end{aligned}$
+
+$y_i$ varies monotonically with $x_i$, ALR is symmetric in the first $N$-1 dimensions, but not in the $N_{th}$.
+The simplex obtained from ALR is a linear combination of vectors in $\mathbb{R}^{N-1}$ with weights $\ln \left(x_i\right)$.
+The first $(N-1)$ vectors are the standard basis vectors of $\mathbb{R}^{N-1},\left\{\mathbf{e}_1, \cdots, \mathbf{e}_{N-1}\right\}$, whereas the $N^{t h}$ vector is $-1$.
+Since $1=\left\|\mathbf{e}_i\right\|<\|-\mathbf{1}\|=\sqrt{N-1}$, for the binary case $N=2$ the ALR transform is symmetric between the two labels, but for $N>2$, changes in the last label's probability are magnified by $\sqrt{N-1}$.
+
+\subsubsection{ILR}
+$$
+\operatorname{ILR}(\mathbf{p})=U^T \ln (\mathbf{p})
+$$
+$$
+\mathrm{ILR}^{-1}(\mathbf{v})=\mathcal{C}(\exp (U \mathbf{v}))
+$$
+where $\mathcal{C}(\cdot)$ renormalizes and $U=\left[\mathbf{u}_1, \ldots, \mathbf{u}_{L-1}\right]$ is an orthonormal basis for the hyperplane of $\mathbb{R}^L$ orthogonal to 1 .
+In ILR, the vectors equidistant from each other and from the origin (forming the vertices of a generalized tetrahedron), making ILR symmetric in all N dimensions.
+ILR maps the Aitchison inner product to the Euclidean inner product (unlike ALR which maps to a skewed euclidean inner product)
+\\
+
+{{insert figure here}}
 
@@
 Therefore, regardless of the choice of $g$ or $\alpha$, the individual coordinates of the transformed distribution are independently distributed.
+
 \end{appendices}
